## Load library

```{r}
library(readr)    
library(dplyr)      
library(lme4)       
library(lmerTest)   
library(DHARMa) 
library(emmeans)
library(ggplot2)  
library(ggbeeswarm)
library(car)
library(glmmTMB)
```

## Set working directory

```{r}
setwd("~/Desktop/simon_lasio/data/behavior and activity/tracking/move_analysis")
```

## Load and prepare data

```{r}
# Load the data 
data_raw <- read_delim(file = "animal_ta_move_processed_with_arena.csv",
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

# pour que Treatment antibiotic, reinoculation et Year sont des facteurs
data <- data_raw %>%
  mutate(
    Treatment = factor(Treatment),                   # d’abord en facteur
    Treatment = relevel(Treatment, ref = "zero_no"),
    Antibiotic = factor(Antibiotic),
    Antibiotic = relevel(Antibiotic, ref = "zero"),
    Reinoculation = factor(Reinoculation),
    Reinoculation = relevel(Reinoculation, ref = "no"),
    Year = factor(Year), 
    Arena = factor(Arena)
  )

data$Antibiotic <- factor(data$Antibiotic, levels = c("zero", "low", "high"))

```

## Descriptive statistics

```{r}
#DESCRIPTIVES STATISTICS
moyenne_speed <- tapply(data$Average_Speed, data$Treatment, mean, na.rm = TRUE)
ecarts_speed <- tapply(data$Average_Speed, data$Treatment, sd, na.rm = TRUE)

resultats_speed <- data.frame(
  Moyenne = moyenne_speed,
  Ecart_Type = ecarts_speed
)

moyenne_prop <- tapply(data$Prop_time_moving, data$Treatment, mean, na.rm = TRUE)
ecarts_prop <- tapply(data$Prop_time_moving, data$Treatment, sd, na.rm = TRUE)

resultats_prop <- data.frame(
  Moyenne = moyenne_prop,
  Ecart_Type = ecarts_prop
)

```

### Export descriptive stats

```{r}
write.csv(resultats_speed, "avg_speed_table.csv", row.names = TRUE)
write.csv(resultats_prop, "prop_of_time_moving_table.csv", row.names = TRUE)
```

# Average speed analysis

## LLM full model

```{r}
hist(data$Average_Speed, breaks = 50)
qqnorm(data$Average_Speed)

model_lmm <- lmer(Average_Speed ~ Antibiotic * Reinoculation + (1 | Year:Arena), data = data)
simulateResiduals(model_lmm) |> plot()
testDispersion(model_lmm)
testResiduals(model_lmm)

s1 <- summary(model_lmm)
s1
a1 <- Anova(model_lmm, type=3)
a1 

s1


p_adjusted_BH <- p.adjust(p_interaction_activity, method = "BH")

bh_results <- data.frame(
  Outcome = labels_activity,
  Raw_p = p_interaction_activity,
  BH_adjusted_p = round(p_adjusted_BH, 4),
  Significant = p_adjusted_BH < 0.05
)

print(bh_results)
```

### Visualisation

```{r}
emm <- emmeans(model_lmm, ~ Antibiotic * Reinoculation)
emm_df <- as.data.frame(emm)


p1 <- ggplot() +
  # A. IC
  geom_rect(data = emm_df,
            aes(xmin = as.numeric(Antibiotic) - 0.3,
                xmax = as.numeric(Antibiotic) + 0.3,
                ymin = lower.CL,  
                ymax = upper.CL,  
                fill = Reinoculation),
            alpha = 0.3, 
            position = position_dodge(width = 0.8),
            show.legend = TRUE) + 

  # B. Predicted mean
  geom_errorbar(data = emm_df,
                aes(x = Antibiotic, 
                    ymin = emmean, 
                    ymax = emmean,
                    color = Reinoculation),
                width = 0.5, 
                position = position_dodge(width = 0.8), 
                size = 1.5) +
  
   # C. Raw data
geom_beeswarm(data = data,
                aes(x = Antibiotic, y = Average_Speed, color = Reinoculation),
                dodge.width = 0.8, cex = 1.5, alpha = 0.5) +
  theme_bw()+ 

  labs(y = "Average speed [cm/sec]", 
       x = "Antibiotic",
       color = "Re-inoculation",
       fill = "Re-inoculation") 
  #ggtitle("Predicted Average Speed with 95% CI")

print(p1)

```

### Save the plot

```{r}
ggsave("avg_speed_plot.pdf", p1, width = 9, height = 5)
```

### Additive model

```{r}
model_lmm_add <- lmer(Average_Speed ~ Antibiotic + Reinoculation + (1 | Year:Arena), data = data)
simulateResiduals(model_lmm_add) |> plot()
testDispersion(model_lmm_add)
testResiduals(model_lmm_add)

s2 <- summary(model_lmm_add)
s2
```

## Reduced model

```{r}
#Same stats but joining LOW and ZERO
data_raw <- read_delim(file = "animal_ta_move_processed_with_arena.csv",
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

data_raw$Antibiotic <- as.character(data_raw$Antibiotic)
data_raw$Antibiotic[data_raw$Antibiotic %in% c("low", "zero")] <- "zero"
data_raw<- data_raw %>%
  mutate(Treatment = paste(Antibiotic, Reinoculation, sep = "_"))

data <- data_raw %>%
  mutate(
    Treatment = factor(Treatment),                   # d’abord en facteur
    Treatment = relevel(Treatment, ref = "zero_no"),
    Antibiotic = factor(Antibiotic),
    Antibiotic = relevel(Antibiotic, ref = "zero"),
    Reinoculation = factor(Reinoculation),
    Reinoculation = relevel(Reinoculation, ref = "no"),
    Year = factor(Year),
    Arena = factor(Arena)
  )
data$Antibiotic <- factor(data$Antibiotic, levels = c("zero", "high"))
```

## LLM reduced

```{r}
hist(data$Average_Speed, breaks = 50)
qqnorm(data$Average_Speed)

model_lmm <- lmer(Average_Speed ~ Antibiotic * Reinoculation + (1 | Year:Arena), data = data)
simulateResiduals(model_lmm) |> plot()
testDispersion(model_lmm)
testResiduals(model_lmm)

s3 <- summary(model_lmm)
s3
a3 <- Anova(model_lmm, type=3)
a3 
```

### additive model

```{r}

#ADDITIVE model 
model_lmm_add <- lmer(Average_Speed ~ Antibiotic + Reinoculation + (1 | Year:Arena), data = data)
simulateResiduals(model_lmm_add) |> plot()
testDispersion(model_lmm_add)
testResiduals(model_lmm_add)

s4 <- summary(model_lmm_add)
s4
```

## Save the result in a .txt file

```{r}
sink("avg_speed_result.txt", split = TRUE)

cat("\n\n  Full model 3x2 for avg speed \n\n")
 
s1
a1
cat("\n\n additive \n\n")
s2


cat("\n\n Reduced model 2x2 for avg speed \n\n")
cat("\n")
s3
a3
cat("\n\n additive \n\n")
s4

sink()
```

# -----------------

# PROPORTION OF TIME MOVING

#-----------------

## Load and prepare data

```{r}
# Load the data 
data_raw <- read_delim(file = "animal_ta_move_processed_with_arena.csv",
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

data <- data_raw %>%
  mutate(
    Treatment = factor(Treatment),                   # d’abord en facteur
    Treatment = relevel(Treatment, ref = "zero_no"),
    Antibiotic = factor(Antibiotic),
    Antibiotic = relevel(Antibiotic, ref = "zero"),
    Reinoculation = factor(Reinoculation),
    Reinoculation = relevel(Reinoculation, ref = "no"),
    Year = factor(Year), 
    Arena = factor(Arena) 
  )
data$Antibiotic <- factor(data$Antibiotic, levels = c("zero", "low", "high"))
levels(data$Treatment)
```

# LLM Full model

```{r}
hist(data$Prop_time_moving, breaks = 35)
qqnorm(data$Prop_time_moving)
range(data$Prop_time_moving)

#Beta-regression should be the best

# transform y to (0,1) range
n <- length(data$Prop_time_moving)
y_adj <- (data$Prop_time_moving * (n - 1) + 0.5) / n
data$prop_time_adj <- y_adj
range(data$prop_time_adj)

model_beta <- glmmTMB(prop_time_adj ~ Antibiotic * Reinoculation + (1|Year/Arena),
                      family = beta_family(link="logit"),
                      data=data) 

#Verifications check
res <- simulateResiduals(model_beta)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta) 

# Models summary
s1 <- summary(model_beta)
s1
a1 <- Anova(model_beta, type=3)
a1 
```

### Visualisation

```{r}
emm <- emmeans(model_beta, ~ Antibiotic * Reinoculation, type = "response")  # back-transform to 0–1
emm_df <- as.data.frame(emm)

p2 <- ggplot() +
  # CI as rectangles
  geom_rect(data = emm_df,
            aes(xmin = as.numeric(Antibiotic) - 0.3,
                xmax = as.numeric(Antibiotic) + 0.3,
                ymin = asymp.LCL,
                ymax = asymp.UCL,
                fill = Reinoculation),
            alpha = 0.3, 
            position = position_dodge(width = 0.8)) +
  # Predicted mean as point
  geom_errorbar(data = emm_df,
                aes(x = Antibiotic, ymin = response, ymax = response,
                    color = Reinoculation),
                width = 0.5, position = position_dodge(width = 0.8), size = 1.5) +
  # Raw data points for context
  geom_beeswarm(data = data,
                aes(x = Antibiotic, y = prop_time_adj, color = Reinoculation),
                dodge.width = 0.8, cex = 1.5, alpha = 0.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_bw() +
  labs(y = "Proportion of time moving", x = "Antibiotic", color="Re-inoculation", fill="Re-inoculation")
  # + ggtitle("Predicted proportion of time moving with 95% CI")
p2


  





```

### Save the plot

```{r}
ggsave("Prop_time_moving_full_plot.pdf", p2,   width = 9, height = 5)
```

### Additive model (just to have it)

```{r}
model_beta_add <- glmmTMB(prop_time_adj ~ Antibiotic + Reinoculation + (1|Year/Arena),
                          family = beta_family(link="logit"),
                          data=data) 

#Verifications check
res <- simulateResiduals(model_beta_add)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta_add) 

#Résumé modele
s2 <- summary(model_beta_add)
s2
a2 <- Anova(model_beta_add, type=3)
a2
```

# Reduced model

```{r}
data_raw <- read_delim(file = "animal_ta_move_processed_with_arena.csv",
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

data_raw$Antibiotic <- as.character(data_raw$Antibiotic)
data_raw$Antibiotic[data_raw$Antibiotic %in% c("low", "zero")] <- "zero"
data_raw<- data_raw %>%
  mutate(Treatment = paste(Antibiotic, Reinoculation, sep = "_"))

data <- data_raw %>%
  mutate(
    Treatment = factor(Treatment),                   # d’abord en facteur
    Treatment = relevel(Treatment, ref = "zero_no"),
    Antibiotic = factor(Antibiotic),
    Antibiotic = relevel(Antibiotic, ref = "zero"),
    Reinoculation = factor(Reinoculation),
    Reinoculation = relevel(Reinoculation, ref = "no"),
    Year = factor(Year)                            
  )
data$Antibiotic <- factor(data$Antibiotic, levels = c("zero", "high"))

# Vérification
levels(data$Treatment)
```

# LLM reduced

```{r}
hist(data$Prop_time_moving, breaks = 35)
qqnorm(data$Prop_time_moving)
range(data$Prop_time_moving)

#Beta-regression should be the best
# transform y to (0,1) range
n <- length(data$Prop_time_moving)
y_adj <- (data$Prop_time_moving * (n - 1) + 0.5) / n
data$prop_time_adj <- y_adj
range(data$prop_time_adj)

model_beta <- glmmTMB(prop_time_adj ~ Antibiotic * Reinoculation + (1|Year/Arena),
                      family = beta_family(link="logit"),
                      data=data) 

#Verifications check
res <- simulateResiduals(model_beta)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta) 


#Résumé modele
s3 <- summary(model_beta)
s3
a3 <- Anova(model_beta, type=3)
a3
```

### Visualisation

```{r}
emm <- emmeans(model_beta, ~ Antibiotic * Reinoculation, type = "response")  # back-transform to 0–1
emm_df <- as.data.frame(emm)

p2 <- ggplot() +
  # CI as rectangles
  geom_rect(data = emm_df,
            aes(xmin = as.numeric(Antibiotic) - 0.3,
                xmax = as.numeric(Antibiotic) + 0.3,
                ymin = asymp.LCL,
                ymax = asymp.UCL,
                fill = Reinoculation),
            alpha = 0.3, 
            position = position_dodge(width = 0.8)) +
  # Predicted mean as point
  geom_errorbar(data = emm_df,
                aes(x = Antibiotic, ymin = response, ymax = response,
                    color = Reinoculation),
                width = 0.5, position = position_dodge(width = 0.8), size = 1.5) +
  # Raw data points for context
  geom_beeswarm(data = data,
                aes(x = Antibiotic, y = prop_time_adj, color = Reinoculation),
                dodge.width = 0.8, cex = 1.5, alpha = 0.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_bw() +
  labs(y = "Proportion of time moving", x = "Antibiotic", color="Reinoculation", fill="Reinoculation")
  # + ggtitle("Predicted proportion of time moving with 95% CI")

p2
```

### Save the plot

```{r}
ggsave("Prop_time_moving_reduced_plot.pdf", p2,   width = 9, height = 5)
```

### additive model

```{r}
model_beta_add <- glmmTMB(prop_time_adj ~ Antibiotic + Reinoculation + (1|Year/Arena),
                          family = beta_family(link="logit"),
                          data=data)

#Verifications check
res <- simulateResiduals(model_beta_add)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta) 

#Résumé modele
s4 <- summary(model_beta_add)
s4
a4 <- Anova(model_beta_add, type=3)
a4
```

## Save the result in a .txt file

```{r}
sink("prop_time_moving_result.txt", split = TRUE)

cat("\n\n  Full model 3x2 for prop of time moving \n\n")
cat("\n")

s1
a1

cat("\n\n additive \n\n")
cat("\n")

s2
a2


cat("\n\n Reduced model 2x2 for pro of time moving \n\n")
cat("\n")

s3
a3
cat("\n\n additive \n\n")
cat("\n")

s4
a4

sink()
```


### Save the combined plot

```{r}
ggsave("combined_activity_metrics.pdf", p_combined, width = 14, height = 6)
ggsave("combined_activity_metrics.png", p_combined, width = 14, height = 6, dpi = 300)
```


