##### *Lucie Kesnerova + Berra Erkosar Combe, 11/2016, UNIL*

##### *Script usage: To concatenate Ct and STDEV values from excel sheet exported from QuantStudio5 SW*

##### *qPCR data (.xlsx or .xls) will be exported in wd (working directory)*

##### *January 2018: changed libraries used: xlsx -\> readxl (Lucie)*

##### *March 2018: code adjusted to data exported by QuantStudio5*

##### *September 2018, UNIL, project Brains Lucie Kesnerova and Joanito Liberti code for first processing of data*

##### *January 2025: script adjusted by Greg to data exported from BioRad 96 qPCR machine*

##### *Simon, December 2025, UNIGE, add basic LMM model on qPCR number*

# Load packages

```{r}
library(readxl)
```

# Setting working directory

```{r}
setwd("~/Desktop/simon_lasio/data/qPCR")
```

# Read result and load a dataframe

```{r}
myFiles <- list.files(path= "~/Desktop/simon_lasio/data/qPCR/qPCR_data", 
                      recursive = F, pattern = (".(xls|xlsx)"), full.names = T)


empty = data.frame() # make empty variable which will be filled with data from the loop

for(x in myFiles){
  #x <- myFiles[2]
  #t3<- read_excel(myFiles[3], sheet = 1, skip =43, col_names=T) 
  t <- read_excel(x, sheet = 1, col_names=T) 
  # if problems occur, try to reinstall the readxl library. IMPORTANT: Excel files need to be resaved as xlxs (Excel version on qPCR computer is old)
  t <- t[c(2:97),c("Sample","Target","Cq")]
  names(t)[1:3] <- c("Sample.Name","Target.Name","Ct")
  t$Ct <- as.numeric(t$Ct) # NAs will be introduced at the place of 'undetermined' values
  t$Target.Name <- factor(t$Target.Name)
  t$Sample.Name <- factor(t$Sample.Name)
  
   
  for (k in seq(1, nrow(t), by=3)){
    if (is.na(t[k,3])==T &is.na(t[(k+1),3])==T & is.na(t[(k+2),3])==T) {
      t[k,3]=200
      t[k+1,3]=200
      t[k+2,3]=200} 
    # replace all NA values by "200" ONLY in case there is NA in all triplicates - temporary solution to treat negative samples and H2O negative control
  }
  ## another option which didint work is ddply()
  #t <- ddply(tt, c("Sample.Name", "Target.Name"), summarise,
  #m = mean(C??, na.rm=T),
  #sd   = sd(C??))
  t.m <- aggregate(Ct~Sample.Name+Target.Name, t, mean, na.action = na.omit) # calculate means, if NA present in triplicate it still calculate mean from other values in triplicate 
  t.sd <- aggregate(Ct~Sample.Name+Target.Name, t, sd, na.action = na.omit)
  t <- merge(t.m, t.sd, by=c("Sample.Name","Target.Name")) # merge Ct and SD values...
  
  t[t=="200"]<-NA # values '100' transfered back to NA after aggregate()
   
  # Remove in plate false positives - values higher than NTC
  water <- t[t$Sample.Name == "H2O",] 
  targets = levels(t$Target.Name)
  newt = data.frame()
  for (i in targets){
    TempWater=water[water$Target.Name ==i,] # to keep specificity for the target
    if (is.na(TempWater[1,3]) == T) {TempWater[1,3]=41} # if water is NA, change to '41' and then comparte to values
    TempData=t[t$Target.Name==i,]
    for (ii in 1:nrow(TempData)){ # loop in loop to be applied on subset of data based on Target.Name
      if (is.na(TempData[ii,3])==F) {
        if (TempData[ii,3]>=TempWater[1,3]) {TempData[ii,3]=NA} } # loop in loop to compare to water only if not NA
      
      #go to Tempdata (subset with the same target, all rows, 3rd column ~ Ct), compare to the TempWater (1st row, 3rd column) and if its >=, replace with NA
    }
    newt = rbind(newt, TempData)
  }
  # remove negative and positive samples
  newt <- newt[newt$Sample.Name!= "positive control",]
  newt <- newt[newt$Sample.Name!= "H2O",]
  
  empty <- rbind(empty, newt) # rbind() is adding data to 'empty' after each big loop
}

empty <- as.data.frame(empty) # change 'empty' to data frame
names(empty)[1:4] <- c("Sample_ID","Target","Ct","SD")
```

# Save the file

```{r}
write.csv(empty, file="~/Desktop/simon_lasio/data/qPCR/rawdata.csv") # export data in .csv
```

# You can start from here if you have the data

## Load packages

```{r}
library(readxl)
library(lme4)
library(dplyr)
library(ggplot2)
library(car)
```

# Set working directory

```{r}
setwd("~/Desktop/simon_lasio/data/qPCR")
```

# Load data

```{r}
cts = read.csv("rawdata.csv")
```

# Preparation of the data

```{r}
cts$X <- NULL
colnames(cts) <- c("Bee_ID", "Target", "Ct", "SD") # rename columns for better work

# rename levels of targets
cts$Target <- as.factor(cts$Target)
#levels(cts$Target)

# merge together with info about treatment
treatments <- read_excel("LasioExtractions.xlsx", sheet=1) # loads the sheet with info
dt <- merge(cts,treatments, by.x="Bee_ID", by.y="SampleID") # adds Treatment and exp#
#View(dt)

# check data distribution
#ggplot(dt[dt$Target=="UV",], aes(x = Bee_ID, y = Ct)) + geom_bar(stat = "identity") + ggtitle("UV")
#hist(dt$Ct[dt$Target=="UV"], breaks = 10)

# turn remained SD values to NA if Ct = NA
dt$SD[is.na(dt$Ct)] <- NA

# check proportion of not quality data
poor <- dt[dt$SD > 0.7 & dt$Ct < 30,]
poor <- poor[!is.na(poor$Ct),]
a = nrow(poor)
b = nrow(dt)
(a/b)*100   #1 % of data can be doubtful
```

## Calculate Copy number for each target

$$
\text{Copy number} = (\text{Efficiency}^\text{Intercept - C^t} ) \times \text{dilution factor}
$$

```{r}
StdCurves <- read_excel("StdCurves_20250312.xlsx", sheet=1) # export intercept, slope and threshold of detection values for each target
StdCurves <- StdCurves[c(1:2),c(2:7)] # select only data needed
StdCurves <- as.data.frame(StdCurves) # table from excel is loaded as 'tibble' -> needs to be fixed

# calculate CopyNums of 16S gene
dt$copynum <- ((StdCurves$Efficiency[2] ^ (StdCurves$Intercept[2] - dt$Ct))*200)  #number in millions
dt$copynum <- round(dt$copynum)
```

# Export the data

```{r}
write.csv(dt, file="~/Desktop/simon_lasio/data/qPCR/lasio_qpcr.csv") # export data in .csv
```

## Exploratory plot to discover the data

```{r}
dt$Treatment <- factor(dt$Treatment, levels= c( "BeforeExperiment", "Ctr","AntLow","AntLow+hom","AntHigh","AntHigh+hom", "GutHomFeedingSolution(SecondStep)", "GutHomogenate", "GlycerolPBScontrol(FirstStep)", "ControlFeedingSolution(SecondStep)"))
dt$Treatment <- relevel(dt$Treatment, ref = "Ctr")
dt$Replicate <- factor(dt$Replicate)


p1 <- ggplot(dt, aes(x = Treatment, y = copynum))+ 
  geom_boxplot(outlier.shape = NA) +
  #geom_beeswarm(aes(color(Treatment)),cex=1) +
  geom_point(aes(colour=Treatment), shape = 16, size = 3, show.legend = FALSE, alpha = 0.7) +
  scale_x_discrete()+ 
  scale_y_log10()+
  annotation_logticks(base = 10, sides = "l", scaled = TRUE,
                      short = unit(0.05, "cm"), mid = unit(0.1, "cm"), long = unit(0.2, "cm"),
                      colour = "black", size = 0.5, linetype = 1, alpha = 1, color = NULL)+

  theme_bw()
print(p1)
```
##Descriptive stats 

```{r}
library(dplyr)

# 1. Filtrer pour exclure les contrôles (si ce n'est pas déjà fait)
dt_clean <- dt %>% 
  filter(!Treatment %in% c("BeforeExperiment", 
                           "GutHomFeedingSolution(SecondStep)", 
                           "GutHomogenate", 
                           "GlycerolPBScontrol(FirstStep)", 
                           "ControlFeedingSolution(SecondStep)"))

# 2. Calculer les statistiques globales
overall_stats <- dt_clean %>%
  summarise(
    # Stats sur les valeurs brutes
    Median_Raw = median(copynum, na.rm = TRUE),
    Mean_Raw   = mean(copynum, na.rm = TRUE),
    SD_Raw     = sd(copynum, na.rm = TRUE),
    Min_Raw    = min(copynum, na.rm = TRUE),
    Max_Raw    = max(copynum, na.rm = TRUE),
    
    # Stats sur les valeurs Log10 (plus robustes)
    Mean_Log10 = mean(log10(copynum), na.rm = TRUE),
    SD_Log10   = sd(log10(copynum), na.rm = TRUE)
  )

cat("--- STATISTIQUES GLOBALES (EXPÉRIMENTALES) ---\n")
print(overall_stats)

# 3. Calculer les statistiques par traitement (Médiane et Log10)
stats_by_group <- dt_clean %>%
  group_by(Treatment) %>%
  summarise(
    N = n(),
    Median_Copynum = median(copynum, na.rm = TRUE),
    Mean_Log10 = mean(log10(copynum), na.rm = TRUE),
    SD_Log10   = sd(log10(copynum), na.rm = TRUE)
  ) %>%
  arrange(desc(Median_Copynum))

cat("\n--- STATISTIQUES PAR GROUPE ---\n")
print(stats_by_group)
```

## Statistics (linear mixed model)

### Using Unique Replicate as random effect

```{r}
dt <- dt %>% filter(!is.na(Antibiotic))
dt$UniqueReplicate <- paste(dt$Antibiotic, dt$Reinoculation, dt$Replicate, sep="_")
length(unique(dt$UniqueReplicate)) #verification
#View(dt)

dt$Antibiotic <- factor(dt$Antibiotic, levels = c("Zero", "Low", "High"))
dt$Reinoculation <- factor(dt$Reinoculation)
dt$Reinoculation <- relevel(dt$Reinoculation, ref = "No")

dt$Treatment <- interaction(dt$Antibiotic, dt$Reinoculation, drop = TRUE)

#Testing treatment (3X2 levels)
m2 <- lmer(log10(copynum) ~ Treatment + (1|UniqueReplicate), data = dt)

#Antibiotic treatment (3 levels)
m3 <- lmer(log10(copynum) ~ Antibiotic + (1|UniqueReplicate), data = dt)

#Antibiotic treatment (2 levels)
m4 <- lmer(log10(copynum) ~ Reinoculation + (1|UniqueReplicate), data = dt)

```

# Visualisation of qPCR

```{r}
p2 <- ggplot(dt, aes(x = Antibiotic, y = log10(copynum), fill = Reinoculation)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
       x = "Antibiotic dose",
       y = "Log10(Copies qPCR)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p2)
```

## Save the plots

```{r}
ggsave("qPCR_exploratory_plot.pdf",   p1,   width = 16, height = 4)
ggsave("qPCR_by_treatment.pdf", p2, width = 8, height = 4)
```

## Save the result in a .txt file

```{r}
output_file <- "qpcr_result.txt"
sink(output_file)

summary(m2)
anova(m2)
summary(m3)
anova(m3)
summary(m4)
anova(m4)

sink()
```



