# Behavior statistics

## Load library

```{r}
# Essential Statistics & Modeling
library(glmmTMB)   # Main GLMM engine
library(car)       # Significance testing (Anova)
library(DHARMa)    # Residual diagnostics
library(emmeans)   # Post-hoc means and predictions

# Multivariate Analysis (PCA & PERMANOVA)
library(dimensio)  # Modern PCA visualization
library(vegan)     # Adonis/Betadisper 

# Data Manipulation & Input
library(readr)     # Fast data loading
library(tidyr)     # Data reshaping
library(dplyr)     # Data wrangling

# Visualization
library(ggplot2)   # Main plotting engine
library(ggbeeswarm) # For spread-out points in plots

```

## Set working directory

```{r}
setwd("~/Desktop/simon_lasio/data/behavior and activity/behavior")
```

# 3 x 2 model

## Load and prepare data

```{r}
data <- read_delim(file = "detailed_petri_counts.csv", delim =  ";",
                   locale = locale(decimal_mark = ","),
                   trim_ws = TRUE,
                   col_names = TRUE,
                   progress = FALSE,
                   guess_max = 2000,
                   show_col_types = FALSE)

colnames(data)<- c("petridish.number", "category", "antibiotic", "reinoculate",
                   "Year", "nb.of.bump", "nb.of.avoidance", "nb.of.head.to.head",
                   "nb.of.touch", "nb.of.antennal.interaction")

data$antibiotic <- factor(data$antibiotic)
data$reinoculate <- factor(data$reinoculate)

#Make sure the reference is zero antibiotic and no recolonisation, year as a factor
data$antibiotic <- factor(data$antibiotic, levels = c("zero", "low", "high"))
data$reinoculate <- relevel(data$reinoculate, ref = "No")
data$Year <- factor(data$Year)
```

## DESCRIPTIVES STATISTICS

```{r}
moyenne_bump <- tapply(data$nb.of.bump, data$category, mean, na.rm = TRUE)
ecarts_bump <- tapply(data$nb.of.bump, data$category, sd, na.rm = TRUE)

resultats_bump <- data.frame(
  Moyenne = moyenne_bump,
  Ecart_Type = ecarts_bump
)

moyenne_avoidance <- tapply(data$nb.of.avoidance, data$category, mean, na.rm = TRUE)
ecarts_avoidance <- tapply(data$nb.of.avoidance, data$category, sd, na.rm = TRUE)

resultats_avoidance <- data.frame(
  Moyenne = moyenne_avoidance,
  Ecart_Type = ecarts_avoidance
)

moyenne_head_to_head <- tapply(data$nb.of.head.to.head, data$category, mean, na.rm = TRUE)
ecarts_head_to_head <- tapply(data$nb.of.head.to.head, data$category, sd, na.rm = TRUE)

resultats_head_to_head <- data.frame(
  Moyenne = moyenne_head_to_head,
  Ecart_Type = ecarts_head_to_head
)

moyenne_touch <- tapply(data$nb.of.touch, data$category, mean, na.rm = TRUE)
ecarts_touch <- tapply(data$nb.of.touch, data$category, sd, na.rm = TRUE)

resultats_touch <- data.frame(
  Moyenne = moyenne_touch,
  Ecart_Type = ecarts_touch
)

formater_stats <- function(moyenne, et) {
  paste0(round(moyenne, 2), " \u00B1 ", round(et, 2)) # \u00B1 is the symbol +/-
}

tableau_publication <- data.frame(
  Category = rownames(resultats_bump), 
  stringsAsFactors = FALSE
)

tableau_publication$`Random bump` <- formater_stats(
  moyenne_bump, 
  ecarts_bump
)

tableau_publication$`Short antennal touch` <- formater_stats(
  moyenne_touch, 
  ecarts_touch
)

tableau_publication$`Head-to-head interaction` <- formater_stats(
  moyenne_head_to_head, 
  ecarts_head_to_head
)

tableau_publication$Avoidance <- formater_stats(
  moyenne_avoidance, 
  ecarts_avoidance
)

colnames(tableau_publication)[1] <- "Traitement"

print(tableau_publication)

```

## Export descriptive stats

```{r}
write.csv(tableau_publication, "tableau_resultats_comportements.csv", row.names = FALSE)
```

## Linear mixed models

### Function to run DHARMa diagnostics
```{r}
check_my_model <- function(model, title = "Model Diagnostics") {
  cat("\n--- Diagnostics for:", title, "---\n")
  
  # 1. Simulate residuals
  res <- simulateResiduals(fittedModel = model, plot = FALSE)
  
  # 2. Plot QQ-plot and Residuals vs. Predicted
  plot(res)
  
  # 3. Specific tests
  print(testDispersion(res))      # Over/Under-dispersion
  print(testZeroInflation(res))   # Zero-inflation
  print(testOutliers(res))        # Outlier detection
}
```

### TEST the nb of single antennal touch

```{r}

aggregate(nb.of.touch ~ antibiotic, data, sum) 
var(data$nb.of.touch)
mean(data$nb.of.touch)

# Negative Binomial GLMM 
model_nb2_single_touch <- glmmTMB(
  nb.of.touch ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_single_touch)
Anova(model_nb2_single_touch, type=3)


#Additive model
model_nb2_add_single_touch <- glmmTMB(
  nb.of.touch ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_add_single_touch)
Anova(model_nb2_add_single_touch, type = 2)

#to test if the interaction if significant
anova(model_nb2_single_touch,model_nb2_add_single_touch)

```

### TEST the nb of head to head interaction

```{r}
aggregate(nb.of.head.to.head ~ antibiotic, data, sum)
var(data$nb.of.head.to.head) 
mean(data$nb.of.head.to.head) 


#Negative Binomial GLMM (NB2)
model_nb2_head <- glmmTMB(
  nb.of.head.to.head ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_head)
Anova(model_nb2_head, type=3)


#Additive model
model_nb2_head_add <- glmmTMB(
  nb.of.head.to.head ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_head_add)
Anova(model_nb2_head_add, type=2)

#to test if the interaction if significant
anova(model_nb2_head,model_nb2_head_add)
```

### TEST the nb of Avoidance

```{r}
aggregate(nb.of.avoidance ~ antibiotic, data, sum) 
var(data$nb.of.avoidance) 
mean(data$nb.of.avoidance) 

# Negative Binomial GLMM (NB2)
model_nb2_avoid <- glmmTMB(
  nb.of.avoidance ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_avoid)
Anova(model_nb2_avoid, type=3)


#Additive model
model_nb2_add_avoid <- glmmTMB(
  nb.of.avoidance ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_add_avoid)
Anova(model_nb2_add_avoid, type=2)

#to test if the interaction if significant
anova(model_nb2_avoid, model_nb2_add_avoid)
```

### TEST the nb of BUMP

```{r}
aggregate(nb.of.bump ~ antibiotic, data, sum) 
var(data$nb.of.bump) 
mean(data$nb.of.bump)

# Negative Binomial GLMM
model_nb2_bump <- glmmTMB(
  nb.of.bump ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_bump)
Anova(model_nb2_bump, type=3)


#Additive model
model_nb2_add_bump <- glmmTMB(
  nb.of.bump ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_add_bump)
Anova(model_nb2_add_bump, type = 2)

#to test if the interaction if significant
anova(model_nb2_bump, model_nb2_add_bump)
```

### Test residuals 

```{r}
# Diagnostics for Single Antennal Touch
check_my_model(model_nb2_single_touch, "3x2 Single Antennal Touch")

# Diagnostics for Head to Head
check_my_model(model_nb2_head, "3x2 Head to Head")

# Diagnostics for Avoidance
check_my_model(model_nb2_avoid, "3x2 Avoidance")

# Diagnostics for Bump
check_my_model(model_nb2_bump, "3x2 Bump")
```


## Visualisation all behavior categories

### Extraction of predictions (EMMEANS) for the 4 behavioral metrics

```{r}
get_emm_data <- function(model, label_name) {
  emm <- emmeans(model, ~ antibiotic * reinoculate, type = "response")
  df <- as.data.frame(emm)
  df$Behavior <- label_name
  return(df)
}

df_touch <- get_emm_data(model_nb2_single_touch, "Single Antennal Touch")
df_head  <- get_emm_data(model_nb2_head, "Head to Head")
df_avoid <- get_emm_data(model_nb2_avoid, "Avoidance")
df_bump  <- get_emm_data(model_nb2_bump, "Bump")

emm_all <- rbind(df_touch, df_head, df_avoid, df_bump)
```

### Preparation of raw data (long format)

```{r}
data_long <- data %>%
  pivot_longer(
    cols = c("nb.of.touch", "nb.of.head.to.head", "nb.of.avoidance", "nb.of.bump"),
    names_to = "variable",
    values_to = "Count"
  ) %>%
  mutate(Behavior = case_when(
    variable == "nb.of.touch" ~ "Single Antennal Touch",
    variable == "nb.of.head.to.head" ~ "Head to Head",
    variable == "nb.of.avoidance" ~ "Avoidance",
    variable == "nb.of.bump" ~ "Bump"
  ))

# Have the right order for the plot
emm_all$Behavior <- factor(emm_all$Behavior, 
                           levels = c("Single Antennal Touch", "Bump", "Head to Head", "Avoidance"))
data_long$Behavior <- factor(data_long$Behavior, 
                             levels = c("Single Antennal Touch", "Bump","Head to Head", "Avoidance"))
```

### Creation of combined plot

```{r}
p_combined <- ggplot() +
  # A. IC
  geom_rect(data = emm_all,
            aes(xmin = as.numeric(antibiotic) - 0.3,
                xmax = as.numeric(antibiotic) + 0.3,
                ymin = asymp.LCL,  # Limite basse (Lower Confidence Limit)
                ymax = asymp.UCL,  # Limite haute (Upper Confidence Limit)
                fill = reinoculate),
            alpha = 0.3, 
            position = position_dodge(width = 0.8),
            show.legend = FALSE) + # On cache la légende ici pour éviter les doublons
  
  # B. Predicted mean
  geom_errorbar(data = emm_all,
                aes(x = antibiotic, 
                    ymin = response, 
                    ymax = response,
                    color = reinoculate),
                width = 0.5, 
                position = position_dodge(width = 0.8), 
                size = 1.2) +
  
  # C. Raw data
  geom_beeswarm(data = data_long,
                aes(x = antibiotic, 
                    y = Count, 
                    color = reinoculate),
                dodge.width = 0.8, 
                cex = 1.2,        # Taille de dispersion
                size = 1.5,       # Taille des points
                alpha = 0.4) +    # Transparence
  
  # D. Facetting
  facet_wrap(~ Behavior, scales = "free_y") + # Echelles Y libres car les comptes varient
  theme_bw() +
  scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
  labs(y = "Number of interactions", 
       x = "Antibiotic dose",
       color = "Reinoculation",
       fill = "Reinoculation") +
  theme(strip.text = element_text(size = 12, face = "bold"),
        legend.position = "bottom")

print(p_combined)
```

## Save the plot

```{r}
ggsave("All_Behaviors_Plot.pdf", p_combined, width = 12, height = 8)
```

## PCA analysis

### Prepare the data

```{r}
data$antibiotic <- factor(data$antibiotic)
data$reinoculate <- factor(data$reinoculate)
data$Year <- factor(data$Year)
data$antibiotic <- factor(data$antibiotic, levels = c("zero", "low", "high"))
data$reinoculate <- relevel(data$reinoculate, ref = "No")
```

## PCA pour Treatment (3X2)

```{r}
data2 <- data[, !(names(data) %in% c("antibiotic", "reinoculate", "Year", "petridish.number", "nb.of.antennal.interaction", "total.interactions"))]
head(data2)
X <- dimensio::pca(data2, center = TRUE, scale = FALSE, sup_quali = "category")
get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

p1 <- viz_individuals(
  x = X,
  extra_quali = data2$category,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)

set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data2[, sapply(data2, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data$category)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat <- permutest(dispersion_mod, permutations = 999)
permu_treat
# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_treat <- adonis2(dist_matrix ~ category, strata=data$Year, data = data, method = "euclidean")
ado_treat

```

## PCA pour Antibiotic

```{r}
data3 <- data[, !(names(data) %in% c("category", "reinoculate", "Year", "petridish.number", "nb.of.antennal.interaction", "total.interactions"))]
head(data3)

X <- dimensio::pca(data3, center = TRUE, scale = FALSE, sup_quali = "antibiotic")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data3$antibiotic,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)


set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data3[, sapply(data3, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data$antibiotic)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_ant <- permutest(dispersion_mod, permutations = 999)
permu_ant
# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_ant <- adonis2(dist_matrix ~ antibiotic, strata=data$Year, data = data, method = "euclidean")
ado_ant
```

## PCA pour réinoculation

```{r}
data4 <- data[, !(names(data) %in% c("category", "antibiotic", "Year", "petridish.number", "nb.of.antennal.interaction", "total.interactions"))]
head(data4)

X <- dimensio::pca(data4, center = TRUE, scale = FALSE, sup_quali = "reinoculate")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data4$reinoculate,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)


set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data4[, sapply(data4, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data$reinoculate)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_reino <- permutest(dispersion_mod, permutations = 999)
permu_reino

# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_reino <- adonis2(dist_matrix ~ reinoculate, strata=data$Year, data = data, method = "euclidean")
ado_reino

```

# 2 x 2 model

```{r}
# We join togheter low and zero into the "zero category"
data$antibiotic[data$antibiotic %in% c("low", "zero")] <- "zero"
data$antibiotic <- factor(data$antibiotic)
levels(data$antibiotic)

#Make sure the reference is zero antibiotic and no recolonisation, year as a factor
data$antibiotic <- relevel(data$antibiotic, ref = "zero")
data$reinoculate <- relevel(data$reinoculate, ref = "No")
data$Year <- factor(data$Year)
```

## Linear mixed models

### TEST the nb of single antennal touch

```{r}
aggregate(nb.of.touch ~ antibiotic, data, sum) 

var(data$nb.of.touch)
mean(data$nb.of.touch)

# Negative Binomial GLMM 
model_nb2_single_touch_red <- glmmTMB(
  nb.of.touch ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_single_touch_red)


#Additive model
model_nb2_add_single_touch_red <- glmmTMB(
  nb.of.touch ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_add_single_touch_red)

anova(model_nb2_add_single_touch_red, model_nb2_single_touch_red)
```

### TEST the nb of head to head interaction

```{r}
aggregate(nb.of.head.to.head ~ antibiotic, data, sum)
var(data$nb.of.head.to.head) 
mean(data$nb.of.head.to.head) 


#Negative Binomial GLMM (NB2)
model_nb2_head_red <- glmmTMB(
  nb.of.head.to.head ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_head_red)


model_nb2_add_head_red <- glmmTMB(
  nb.of.head.to.head ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)
summary(model_nb2_add_head_red)


anova(model_nb2_head_red, model_nb2_add_head_red)
```

### TEST the nb of Avoidance

```{r}
aggregate(nb.of.avoidance ~ antibiotic, data, sum) 

var(data$nb.of.avoidance) 
mean(data$nb.of.avoidance) 

# Negative Binomial GLMM (NB2)
model_nb2_avoid_red <- glmmTMB(
  nb.of.avoidance ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_avoid_red)


#Additive model 
model_nb2_add_avoid_red <- glmmTMB(
  nb.of.avoidance ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_add_avoid_red)

anova(model_nb2_avoid_red, model_nb2_add_avoid_red)
```

### TEST the nb of BUMP

```{r}
aggregate(nb.of.bump ~ antibiotic, data, sum) 

var(data$nb.of.bump) 
mean(data$nb.of.bump)

# Negative Binomial GLMM
model_nb2_bump_red <- glmmTMB(
  nb.of.bump ~ antibiotic * reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_bump_red)

#Additive model
model_nb2_add_bump_red <- glmmTMB(
  nb.of.bump ~ antibiotic + reinoculate + (1 | Year),
  family = nbinom2,
  data = data
)

summary(model_nb2_add_bump_red)

anova(model_nb2_bump_red, model_nb2_add_bump_red)
```


# Models comparison between 3x2 and 2x2 (if both are significant)

```{r}
anova(model_nb2_add_single_touch, model_nb2_add_single_touch_red)

anova(model_nb2_add_head_red, model_nb2_head_add)

anova(model_nb2_add_avoid, model_nb2_add_avoid_red)

anova(model_nb2_add_bump, model_nb2_add_bump_red)
```

## PCA analysis

### Data preparation

```{r}
data$antibiotic <- factor(data$antibiotic)
data$reinoculate <- factor(data$reinoculate)
data$Year <- factor(data$Year)

data$antibiotic <- factor(data$antibiotic, levels = c("zero", "high"))
data$reinoculate <- relevel(data$reinoculate, ref = "No")

data$category[data$category %in% c("low_ant", "control")] <- "control"
data$category[data$category %in% c("low_gut", "gut_ctrl")] <- "gut_ctrl"
```

## PCA pour Treatment (2X2)

```{r}
data2 <- data[, !(names(data) %in% c("antibiotic", "reinoculate", "Year", "petridish.number", "nb.of.antennal.interaction", "total.interactions"))]
head(data2)

X <- dimensio::pca(data2, center = TRUE, scale = FALSE, sup_quali = "category")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data2$category,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)



set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data2[, sapply(data2, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data$category)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat2x2 <- permutest(dispersion_mod, permutations = 999)
permu_treat2x2
# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_treat2x2 <- adonis2(dist_matrix ~ category, strata=data$Year, data = data, method = "euclidean")
ado_treat2x2

```

## PCA pour antibiotic

```{r}
data3 <- data[, !(names(data) %in% c("category", "reinoculate", "Year", "petridish.number", "nb.of.antennal.interaction", "total.interactions"))]
head(data3)

X <- dimensio::pca(data3, center = TRUE, scale = TRUE, sup_quali = "antibiotic")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data3$antibiotic,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)



set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data3[, sapply(data3, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data$category)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_ant2x2 <- permutest(dispersion_mod, permutations = 999)
permu_ant2x2
# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_ant2x2 <- adonis2(dist_matrix ~ category, strata=data$Year, data = data, method = "euclidean")
ado_ant2x2


```

## PCA pour réinoculation

```{r}
data4 <- data[, !(names(data) %in% c("category", "antibiotic", "Year", "petridish.number", "nb.of.antennal.interaction", "total.interactions"))]
head(data4)

X <- dimensio::pca(data4, center = TRUE, scale = TRUE, sup_quali = "reinoculate")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data4$reinoculate,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)


set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data4[, sapply(data4, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data$category)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_reino2x2 <- permutest(dispersion_mod, permutations = 999)
permu_reino2x2

# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_reino2x2 <- adonis2(dist_matrix ~ category, strata=data$Year, data = data, method = "euclidean")
ado_reino2x2

```

# Save the result in a .txt file

```{r}

# ===============================
# Save all statistical results to a .txt file
# ===============================

sink("behavior_statistics_results.txt", split = TRUE)

cat("=====================================\n")
cat("DESCRIPTIVE STATISTICS\n")
cat("=====================================\n\n")

print(tableau_publication)

cat("\n\n=====================================\n")
cat("3 x 2 MODEL — GLMM RESULTS\n")
cat("=====================================\n\n")

cat("\n--- Single Antennal Touch ---\n")
summary(model_nb2_single_touch)
summary(model_nb2_add_single_touch)

anova(model_nb2_single_touch, model_nb2_add_single_touch)

cat("\n--- Head-to-Head Interaction ---\n")
summary(model_nb2_head)
Anova(model_nb2_head, type = 3)
summary(model_nb2_head_add)
Anova(model_nb2_head_add, type = 2)

anova(model_nb2_head, model_nb2_head_add)

cat("\n--- Avoidance ---\n")
summary(model_nb2_avoid)
Anova(model_nb2_avoid, type = 3)
summary(model_nb2_add_avoid)
Anova(model_nb2_add_avoid, type = 2)

anova(model_nb2_avoid, model_nb2_add_avoid)

cat("\n--- Bump ---\n")
summary(model_nb2_bump)
Anova(model_nb2_bump, type = 3)
summary(model_nb2_add_bump)
Anova(model_nb2_add_bump, type = 2)

anova(model_nb2_bump, model_nb2_add_bump)

cat("\n--- PCA 3x2 ---\n")
permu_treat 
ado_treat
permu_ant 
ado_ant
permu_reino  
ado_reino

cat("\n\n=====================================\n")
cat("2 x 2 MODEL — GLMM RESULTS\n")
cat("=====================================\n\n")

cat("\n--- Single Antennal Touch ---\n")
summary(model_nb2_single_touch_red)
summary(model_nb2_add_single_touch_red)

cat("\n--- Head-to-Head Interaction ---\n")
summary(model_nb2_head_red)
summary(model_nb2_add_head_red)

cat("\n--- Avoidance ---\n")
summary(model_nb2_avoid_red)
summary(model_nb2_add_avoid_red)

cat("\n--- Bump ---\n")
summary(model_nb2_bump_red)
summary(model_nb2_add_bump_red)


cat("\n--- PCA 2x2 ---\n")

permu_treat2x2
ado_treat2x2
permu_ant2x2
ado_ant2x2
permu_reino2x2
ado_reino2x2

cat("\n=====================================\n")
cat("END OF ANALYSIS\n")
cat("=====================================\n")

sink()
```
