---
output: html_document
editor_options: 
  chunk_output_type: inline
---

# Lasioglossum malachurum antiobiotic exposure microbiome experiment - Joanito Liberti and Simon Levy

Liberti Lab - Analyses of 16S rRNA gene amplicon-sequencing

## If you have the ps object ("ps_dada2taxa_Lasio.rds"), you can start later with it directly

## Number of reads per sample

```{bash}
cd /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles
for f in *fastq.gz; do
echo $f
echo $(zless $f | wc -l)/4|bc
done
```

## total number of reads

```{bash}
cd /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles

# sum reads for all fastq.gz files
total_reads=0
for f in *.fastq.gz; do
    # count lines
    lines=$(zless "$f" | wc -l)
    # divide by 4 to get reads
    reads=$((lines / 4))
    # add to total
    total_reads=$((total_reads + reads))
done

echo "Total reads across all files: $total_reads"
```

## Check data quality with FastQC

```{bash}
mkdir /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2
mkdir /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis
mkdir /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/00_FASTQC/
  cd /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles

for f in *.fastq.gz; do 
fastqc $f -o /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/00_FASTQC/; 
done
```

## Clean up sample names

```{bash}
cd /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles
for f in *fastq.gz; do mv $f ${f/_001_*/.fastq.gz}; done
for f in *fastq.gz; do mv $f ${f/_L1/}; done
```

## Remove primers and anything before and after with Cutadapt

```{bash}

mkdir /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/02_dada2
cd /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles
  for f in *R1.fastq.gz; do /Users/SEL/Library/Python/3.12/bin/cutadapt -G GGACTACHVGGGTWTCTAAT -g GTGCCAGCMGCCGCGGTAA -o ${f/.fastq/.cut.fastq} -p  ${f/R1.fastq/R2.cut.fastq} $f ${f/R1.fastq/R2.fastq} --pair-filter=any; done
```

## Move trimmed reads into a separate folder

```{bash}
mkdir /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/01_trimmed
cd  /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles
for f in *cut.fastq.gz; do mv $f /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/01_trimmed/$f; done
```

## Number of reads per sample after cutadapt

```{bash}
cd /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles/


echo -e "Sample""\t""Raw""\t""Cutadapt"

for f in $(ls *R1.fastq.gz | cut -d '_' -f 1|sort|uniq); do
raw=`echo $(zless /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/RawFiles/$f\\_R1.fastq.gz | wc -l)/4|bc`
cut=`echo $(zless /Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/01_trimmed/$f\\_R1.cut.fastq.gz | wc -l)/4|bc`

echo -e $f"\t"$raw"\t"$cut
done
```

## Now we can feed the quality-filtered reads to the DADA2 pipeline

## First, install required packages if needed (uncomment all installation lines)

```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(version = "3.10")
# BiocManager::install("dada2", version = "3.10")
# BiocManager::install("DECIPHER")
# BiocManager::install("phyloseq")
# BiocManager::install("biomformat")
# BiocManager::install("ShortRead")
# BiocManager::install("Biostrings")
# BiocManager::install("genefilter")
# BiocManager::install("decontam")
# BiocManager::install("DESeq2")

# install.packages("ggplot2")
# install.packages("phangorn")
# install.packages("vegan")
# install.packages("dplyr")
# install.packages("scales")
# install.packages("RColorBrewer")
# install.packages(“reshape2”)
# install.packages("cowplot")
# install.packages("tidyverse")
# install.packages("readxl")
# install.packages("ggbeeswarm")
# install.packages("magrittr")
# install.packages("ggpubr")  
# install.packages('dendextend')
# install.packages("multcomp")
# install.packages("GUniFrac")

# Load libraries

library(dada2)
library(ShortRead)
library(Biostrings)
library(ggplot2)
library(biomformat)
```

## Set the working directory

```{r}
path <- "/Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/02_dada2"
knitr::opts_knit$set(root.dir = normalizePath(path)) 
```

# Sample names

```{r}
library(dada2); packageVersion("dada2")
pathraw <-"/Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/01_trimmed"

list.files(pathraw)
# Forward and reverse fastq filenames have format: SAMPLENAME_R1_001.trim.fastq and SAMPLENAME_R2_001.fastq
fnFs <- sort(list.files(pathraw, pattern="R1.cut.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(pathraw, pattern="R2.cut.fastq.gz", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "\\_"), `[`, 1)
sample.names
```

## Quality scores

```{r}
path<-"/Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/01_trimmed"
dev.new()
# Quality scores of R1 reads
plotQualityProfile(fnFs[1:20]) 
plotQualityProfile(fnFs[21:40])
plotQualityProfile(fnFs[41:60])

# Quality scores of R2 reads
plotQualityProfile(fnRs[1:20]) 
plotQualityProfile(fnRs[21:40])
plotQualityProfile(fnRs[41:60])

```

## Trim the data

```{r}
# Place filtered files in filtered/ subdirectory
path<-"/Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/analysis/02_dada2"
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

sys_str <- Sys.time()
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(232,231), # truncLen[[1]] + truncLen[[2]] > amplicon_length+25
                     maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,      
                     compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE
sys_str[2] <- Sys.time()
sys_str
rm(sys_str)

derepFs <- derepFastq(filtFs)
derepRs <- derepFastq(filtRs)
names(derepFs) <- sample.names
names(derepRs) <- sample.names

head(out)
out
```

## Learn the error rates

```{r}
sys_str <- Sys.time()
set.seed(1) # Since we randomize what samples we use for learning the error rates, we use set.seed() to make the analysis reproducible
errF <- learnErrors(filtFs, randomize=TRUE, nbases=3e8, multithread=TRUE) # here we need to increase nbases to sample more than only a few samples, default is 1e8
set.seed(1)
errR <- learnErrors(filtRs, randomize=TRUE, nbases=3e8, multithread=TRUE) # here we need to increase nbases to sample more than only a few samples, default is 1e8
plotErrors(errF, nominalQ=TRUE)

# In the plots, the black line is the error model, the dots are the actual errors
sys_str[2] <- Sys.time()
sys_str
rm(sys_str)
```

## Sample inference

```{r}
sys_str <- Sys.time()
dadaFs <- dada(derepFs, err=errF, multithread=TRUE)
dadaRs <- dada(derepRs, err=errR, multithread=TRUE)
sys_str[2] <- Sys.time()
sys_str
rm(sys_str)

dadaFs[[1]]
dadaRs[[1]]
```

## Merge paired reads

```{r}
mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, verbose=TRUE, trimOverhang=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```

## Construct sequence table

```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```

## Remove sequences that appear too short or long to be real

```{r}
seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 250:274]
```

## Remove chimeras

```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab2, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim)/sum(seqtab2)
```

## Track reads through the pipeline

```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
tail(track)
```

## Assign taxonomy

```{r}
#assignTaxonomy using Silva 
taxa <- assignTaxonomy(seqtab.nochim, "/Volumes/s-genev-liberti/users/Liberti/ErikFrank_Collaborations/MelinaKienitz/AmpliconSequencing_Melina/DADA2/silva_nr99_v138.2_toSpecies_trainset.fa", multithread=TRUE)
taxa <- addSpecies(taxa, "/Volumes/s-genev-liberti/users/Liberti/ErikFrank_Collaborations/MelinaKienitz/AmpliconSequencing_Melina/DADA2/silva_v138.2_assignSpecies.fa")

taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

```{bash}
mkdir /Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/DADA2/analysis/03_Taxonomy
```

```{r}
pathtax <- "/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/DADA2/analysis/03_Taxonomy/"

# Save taxonomy assignments from dada2 pipeline
write.csv2(file=paste(pathtax,"Taxtable_dada2",sep=""),taxa)
```

# Convert to fasta

```{r}
uniquesToFasta(seqtab.nochim, fout='/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/DADA2/analysis/03_Taxonomy/seqtab.nochim.fasta', ids=colnames(seqtab.nochim))
```

# Prepare a phylogenetic tree with phangorn

```{r}
library(DECIPHER) # Make alignment with DECIPHER...
library(phangorn) # ...then build tree with phangorn
asv.seqs <- getSequences(seqtab.nochim )
names(asv.seqs) <- asv.seqs
asv.alignment <- AlignSeqs(DNAStringSet(asv.seqs), anchor=NA)

# convert to phangorn and run phylogenetic tree reconstruction
phang.align <- phyDat(as(asv.alignment, "matrix"), type="DNA")
dm <- dist.ml(phang.align)
treeNJ <- NJ(dm) # Note, tip order != sequence order
fit = pml(treeNJ, data=phang.align)

fitGTR <- update(fit, k=4, inv=0.2)
fitGTR <- optim.pml(fitGTR, model="GTR", optInv=TRUE, optGamma=TRUE,
                      rearrangement = "stochastic", control = pml.control(trace = 0))
detach("package:phangorn", unload=TRUE)
```

## Evaluate accuracy 

```{r}
unqs.mock <- seqtab.nochim["Mock",] 
unqs.mock <- sort(unqs.mock[unqs.mock>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.mock), "sample sequences present in the Mock community.\n")

mock.ref <- getSequences(file.path(path, "/Volumes/s-genev-liberti/users/MSc_students/tajada_paula/Project_DrosophilaTracking/MolecularData/AmpliconSequencing/DADA2/mock_seq_plasmids.fasta"))
match.ref <- sum(sapply(names(unqs.mock), function(x) any(grepl(x, mock.ref))))
cat("Of those,", sum(match.ref), "were exact matches to the expected reference sequences.\n")
```

## Load and filter the data in phyloseq

```{r}
library(ggplot2)
library(vegan) # ecological diversity analysis
library(dplyr)
library(scales) # scale functions for vizualizations
library(grid)
library(RColorBrewer)
library(reshape2) # data manipulation package
library(cowplot)
library(phyloseq)
library(tidyverse)
library(readxl)

# setting the working directory
setwd("/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/DADA2/")

# Set plotting theme
theme_set(theme_bw())

#Data frame containing sample information
samdf = read.table("/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/Metadata_Lasio.txt", header = T, fill=TRUE, sep="\t") # fill=TRUE allows to read a table with missing entries
rownames(samdf) = samdf$SampleID

#Create a phyloseq object
ps.raw <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf),
               phy_tree(fitGTR$tree),
               tax_table(taxa))
sample_data(ps.raw) 

# Make phyloseq objects containing Mitochondria and Chloroplast ASVs so we can see how many there were
ps.mit <- subset_taxa(ps.raw, Family=="Mitochondria", prunesamples=TRUE) # Depending on the classifier used, double check if your Chloroplasts or Mitochondria are in the Order and Family levels...
tax_table(ps.mit)
ps.chl <- subset_taxa(ps.raw, Order=="Chloroplast", prunesamples=TRUE)
tail(tax_table(ps.chl))
```

## Export raw ASV table

```{r}
table = merge( tax_table(ps.raw),t(otu_table(ps.raw)), by="row.names")

write.table(table, "/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/DADA2/RawASVtable_dada2taxa.txt", sep="\t")
```

## Filter the table and clean the classification

```{r}
#Filter out Eukaryota, mitochondria and chloroplast reads
ps <- subset_taxa(
  ps.raw,
  (Kingdom != "Eukaryota" | is.na(Kingdom)) &
  (Kingdom != "Unclassified" | is.na(Kingdom)),
  # Family != "Mitochondria" &
  # Order  != "Chloroplast",
  prunesamples = TRUE
)

# give new names to ASV's
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps

# Instead of having NAs in the taxa names, at all levels of taxonomical assignment we replace them with the lowest level that the classifier could achieve
tax <- data.frame(tax_table(ps))

tax.clean <- data.frame(row.names = row.names(tax),
Kingdom = str_replace(tax[,1], "D_0__",""),
Phylum = str_replace(tax[,2], "D_1__",""),
Class = str_replace(tax[,3], "D_2__",""),
Order = str_replace(tax[,4], "D_3__",""),
Family = str_replace(tax[,5], "D_4__",""),
Genus = str_replace(tax[,6], "D_5__",""),
Species = str_replace(tax[,7], "D_6__",""), # uncheck this if using the Silva classification from DADA2, as that one also has the species column
stringsAsFactors = FALSE)
tax.clean[is.na(tax.clean)] <- ""

for (i in 1:7){ tax.clean[,i] <- as.character(tax.clean[,i])} 
####### Fill holes in the tax table
tax.clean[is.na(tax.clean)] <- ""
for (i in 1:nrow(tax.clean)){

#Fill in missing taxonomy
if (tax.clean[i,2] == ""){
kingdom <- paste("Kingdom_", tax.clean[i,1], sep = "")
tax.clean[i, 2:7] <- kingdom  
} else if (tax.clean[i,3] == ""){
phylum <- paste("Phylum_", tax.clean[i,2], sep = "")
tax.clean[i, 3:7] <- phylum
} else if (tax.clean[i,4] == ""){
class <- paste("Class_", tax.clean[i,3], sep = "")
tax.clean[i, 4:7] <- class
} else if (tax.clean[i,5] == ""){
order <- paste("Order_", tax.clean[i,4], sep = "")
tax.clean[i, 5:7] <- order
} else if (tax.clean[i,6] == ""){
family <- paste("Family_", tax.clean[i,5], sep = "")
tax.clean[i, 6:7] <- family
} else if (tax.clean[i,6] == ""){
tax.clean$Species[i] <- paste("Genus",tax.clean$Genus[i], sep = "_")
}
}

tax_table(ps) <- as.matrix(tax.clean)
```

## Save the phyloseq object so next time you do not need to rerun the analysis from scratch

```{r}
saveRDS(ps, "/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/DADA2/ps_dada2taxa_Lasio.rds")

# Grab the XStringSet object with the OTU/ASV sequences
rs <- refseq(ps)
rs # inspect
# Get strings with the full taxonomy for each OTU/ASV
tax <- tax_table(ps)
head(tax) # inspect
tax_strings <- apply(tax, 1, paste, collapse=";")
head(tax_strings) # inspect
# Create new names for the sequences in the refseq object; these will become
# the sequence headers in the fasta file. Here, I set these to be the OTU/ASV
# name and the tax string separated by one space (determined by the sep
# parameter)
new_names <- paste(taxa_names(ps), tax_strings, sep = " ")
head(new_names) # inspect
# Update the refeq names and save as a fasta file
names(rs) <- new_names
rs # inspect
Biostrings::writeXStringSet(rs, "/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/DADA2/analysis/03_Taxonomy/AllRawASVs.fasta")
```

## Blast all ASVs to complement the taxonomy

```{bash}
cat AllRawASVs.fasta | parallel --block 2000k --recstart '>' --pipe /work/Joanito_new/Blast/ncbi-blast-2.11.0+/bin/blastn -query AllRawASVs.fasta -db /work/nucleotide/nt -outfmt \"6 std stitle\" -evalue 1e-3 -max_target_seqs 5 -num_threads 35 > AllRawASVs_Blast.txt
```

## Filter Blast results to retain only first hit of each ASV

```{r}
bl <- read.table("/Volumes/My\ Passport\ for\ Mac/MiSeqRun_Analyses/Bee_tracking/DADA2/analysis/03_Taxonomy/AllRawASVs_Blast.txt", header=F, sep = "\t")
bl <- bl[!duplicated(bl$V1),]
write.table(bl, "/Volumes/My\ Passport\ for\ Mac/MiSeqRun_Analyses/Bee_tracking/DADA2/analysis/03_Taxonomy/AllRawASVs_Blast_OnlyFirstHit.txt", sep = "\t")
write.table(tax_table(ps), "/Volumes/My\ Passport\ for\ Mac/MiSeqRun_Analyses/Bee_tracking/DADA2/analysis/03_Taxonomy/TaxTable_ps.txt", sep = "\t")
```

# The analysis can be started from here if the ps object was saved

```{r}
library(ggplot2)
library(vegan) # ecological diversity analysis
library(dplyr)
library(scales) # scale functions for vizualizations
library(grid)
library(RColorBrewer)
library(reshape2) # data manipulation package
library(cowplot)
library(phyloseq)
library(tidyverse)
library(readxl)
library(dada2)

ps <- readRDS("/Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/DADA2/ps_dada2taxa_Lasio.rds")

#Data frame containing sample information
samdf = read.table("/Users/SEL/Desktop/simon_lasio/data/Amplicon-sequencing_new/Metadata_Lasio.txt", header = T, fill=TRUE,  sep="\t") # fill=TRUE allows to read a table with missing entries

```

## Or if you saved the whole session, you can load it like this

```{r}
#load("/Volumes/My\ Passport\ for\ Mac/MiSeqRun_Analyses/Bee_tracking/DADA2/BeeTracking_Ampliseq.rdata")
```

## Plot distribution of reads per sample

```{r}
# Make a data frame with a column for the read counts of each sample
sample_sum_df <- data.frame(sum = sample_sums(ps))

# Histogram of sample read counts
dev.new()
ggplot(sample_sum_df, aes(x = sum)) + 
  geom_histogram(color = "black", fill = "indianred", binwidth = 1000) +
  ggtitle("Distribution of sample sequencing depth") + 
  xlab("Read counts") +
  theme(axis.title.y = element_blank())

# mean, max and min of sample read counts
smin <- min(sample_sums(ps))
smean <- mean(sample_sums(ps))
smax <- max(sample_sums(ps))

# printing the results
cat("The minimum sample read count is:",smin)
cat("The average sample read count is:",smean)
cat("The maximum sample read count is:",smax)

# setting the seed to one value in order to create reproducible results
set.seed(1)  
```

## Rarefaction curves

```{r}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

n = 4
cols = gg_color_hue(n)

ps@sam_data$Sample_Cols<-cols[as.factor(ps@sam_data$Sample_Type)]


dev.new()
pdf(file="Rarefactions.pdf",width=5,height=4, pointsize=4)
rarecurve(as.matrix(data.frame(otu_table(ps))), ylab = "Number of ASVs", xlab = "Number of sequences", col= sample_data(ps)$Sample_Cols, step= 50, cex=0.5)
dev.off()

rarecurve(as.matrix(data.frame(otu_table(ps))), ylab = "Number of ASVs", xlab = "Number of sequences", col= ps@sam_data$Sample_Cols, step= 50, cex=0.5)
```

## Retain only ASVs that have at least 1% relative abundance in minimum 1 sample

```{r}
library(genefilter)
mothur_proportions = transform_sample_counts(ps, function(x) {x/sum(x)})
Filtered = filter_taxa(mothur_proportions, filterfun(kOverA(1, 0.01)), TRUE)
```

## Plot bacterial stacked bars

```{r}
mdf = psmelt(Filtered)
# mdf <- mdf[order(factor(mdf$Treatment, levels = c("Axenic", "Native", "Core", "Negative", "Mock"), ordered=TRUE), mdf$Sample), ]  # Let's first sort by Treatment, then hive, then sample ID
mdf$Sample <- factor(mdf$Sample, levels = unique(mdf$Sample), ordered=TRUE)  # we need to do this to retain the order in plot.
tail(mdf, 4)

library(RColorBrewer)
cbPalette<-colorRampPalette(brewer.pal(12, "Paired"))

set.seed(7) # Here we use set.seed to make the colors reproducible, because in the scale_fill_manual we use "sample" to shuffle the order of colors in the palette
p1 = ggplot(mdf, aes_string(x = "Sample", y = "Abundance", fill = "Genus")) +
        geom_bar(stat = "identity", position = "stack", color = "black") +
        ylab("Relative abundance") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position="bottom") +
        scale_fill_manual(values=sample(cbPalette(48))) 
  
print(p1, width = 1000, height = 200)
```

## See the composition of the gut homogenate 

```{r}
library(dplyr)
library(ggplot2)

# --- 1. Subset the correct object ---
# Use 'ps' (or whichever object gave you the list of 11 treatments above)
ps_homogenate <- subset_samples(ps, Treatment == "GutHomogenate")

# Remove ASVs that aren't present in the homogenate to keep it clean
ps_homogenate <- prune_taxa(taxa_sums(ps_homogenate) > 0, ps_homogenate)

# --- 2. Calculate Relative Abundance (%) ---
ps_hom_rel <- transform_sample_counts(ps_homogenate, function(x) 100 * x / sum(x))

# --- 3. Create a Summary Table (Top 10 Genera) ---
hom_summary <- psmelt(ps_hom_rel) %>%
  group_by(Genus) %>%
  summarise(Mean_Abundance = mean(Abundance)) %>%
  arrange(desc(Mean_Abundance))

print("Top 10 Genera in the Gut Homogenate:")
print(head(hom_summary, 10))

# Plot the top 15 Genera
ggplot(hom_summary[1:15, ], aes(x = reorder(Genus, Mean_Abundance), y = Mean_Abundance, fill = Genus)) +
  geom_bar(stat = "identity", color = "black") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Composition of Gut Homogenate Inoculum",
    subtitle = "Top 15 Genera by Relative Abundance",
    x = "Genus",
    y = "Average Relative Abundance (%)"
  )
```

## Plot stacked bars of the various control samples (blanks, mock, negative PCR controls)

## Plotting only ASVs that have at least 1% relative abundance in 1 sample

```{r}
ps.blanks <- subset_samples(ps, Sample_Type == "Control") 
ps.mock <- subset_samples(ps, Treatment == "Mock")
# ps.H2O <- subset_samples(ps, Treatment == "H2O")
# ps.inoculum <- subset_samples(ps, Treatment == "inoculum")

library(genefilter)
proportions.blanks = transform_sample_counts(ps.blanks, function(x) {x/sum(x)})
Filtered.blanks = filter_taxa(proportions.blanks, filterfun(kOverA(1, 0.01)), TRUE)
proportions.mock = transform_sample_counts(ps.mock, function(x) {x/sum(x)})
Filtered.mock = filter_taxa(proportions.mock, filterfun(kOverA(1, 0.01)), TRUE)
# proportions.H2O = transform_sample_counts(ps.H2O, function(x) {x/sum(x)})
# Filtered.H2O = filter_taxa(proportions.H2O, filterfun(kOverA(1, 0.001)), TRUE)
# proportions.inoculum = transform_sample_counts(ps.inoculum, function(x) {x/sum(x)})
# Filtered.inoculum = filter_taxa(proportions.inoculum, filterfun(kOverA(1, 0.001)), TRUE)
```

```{r}
mdf.blanks = psmelt(Filtered.blanks)
mdf.mock = psmelt(Filtered.mock)

# Make a color palette
library(RColorBrewer)
cbPalette<-colorRampPalette(brewer.pal(15, "Paired"))

p2 = ggplot(mdf.blanks, aes_string(x = "Sample", y = "Abundance", fill = "Genus")) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  ylab("Relative abundance in blank DNA") +
  xlab(element_blank()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position="bottom") +
  scale_fill_manual(values=cbPalette(35)) 

p3 = ggplot(mdf.mock, aes_string(x = "Sample", y = "Abundance", fill = "Genus")) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.2) +
  ylab("Relative abundance in mock DNA") +
  xlab(element_blank()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position="bottom") +
  scale_fill_manual(values=cbPalette(9)) 



print(p2, width = 1000, height = 200)
print(p3, width = 1000, height = 200)

```

## Save plots to pdf

```{r}
dev.new()
pdf(file="Barplots_allsample-LasioProject.pdf",width=12,height=6, pointsize=4)
p1
dev.off()

dev.new()
pdf(file="Barplots_ControlSamples-LasioProject.pdf", width=14,height=20, pointsize=4)
plot_grid(p2, p3, labels = c("A", "B"), ncol = 1, nrow = 4, rel_heights = c(2,1))
dev.off()
```

## Distribution of library sizes

```{r}
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Negative)) + geom_point() 
ggsave(file="DistributionLibrarySizes.pdf", width=8, height=6, useDingbats=FALSE)
```

## Check which sequences are likely to be contaminants (frequency method)

```{r}
library(decontam); packageVersion("decontam")
contamdf.freq <- isContaminant(ps, method="frequency", conc="Picogreen")
head(contamdf.freq)
table(contamdf.freq$contaminant)

which(contamdf.freq$contaminant)
```

## Plot frequencies of identified contaminants vs. their concentrations after PCR

```{r}
plot_frequency(ps, taxa_names(ps)[sample(which(contamdf.freq$contaminant), )], conc="Picogreen") + 
  xlab("DNA Concentration (PicoGreen ng/μl)")
ggsave(file="ContaminantFreqVsPicogreenConc.pdf", width=12, height=10, useDingbats=FALSE)
```

## Check which sequences are likely to be contaminants (prevalence method)

```{r}
contamdf.prev <- isContaminant(ps, neg = "Negative", normalize = TRUE, threshold = 0.1, detailed = T)
table(contamdf.prev$contaminant)

head(which(contamdf.prev$contaminant))
```

## Plot prevalences of identified contaminants

```{r}
# Make phyloseq object of presence-absence in negative controls and true samples
ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Negative == "TRUE", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Negative == "FALSE", ps.pa)
# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                      contaminant=contamdf.prev$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point(size=3, position=position_jitter(h=0.05,w=0.05)) +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
ggsave(file="ContaminantPrevalences.pdf", width=8, height=6, useDingbats=FALSE)
```

## Both methods combined

```{r}
contamdf.both <- isContaminant(ps, conc = "Picogreen", neg = "Negative", method = "either", normalize = TRUE, detailed = T)
table(contamdf.both$contaminant)

which(contamdf.both$contaminant)
```

## New phyloseq object after filtering out contaminants

```{r}
ps.noncontam <- prune_taxa(!contamdf.both$contaminant, ps)
ps.noncontam
```

## Remove non experimental samples before replotting MiSeq results

```{r}

ps.noncontam.filt <- prune_samples(ps.noncontam@sam_data$Sample_Type == "Experimental", ps.noncontam)
ps.noncontam.filt = filter_taxa(ps.noncontam.filt, function(x) sum(x) > 0, TRUE)
ps.noncontam.filt

```

## Save filtered ASV table (not normalised by qPCR yet)

```{r}
write.table(otu_table(ps.noncontam.filt), "LasioProject_ASVtable_OnlyExpSamples.txt", sep="\t")
```

## Histogram of qPCR 16S rRNA copy numbers

```{r}


library(ggh4x)
library(ggplot2)

mdf = sample_data(ps.noncontam.filt)

mdf$Treatment <- factor(mdf$Treatment, levels = c("Ctr", "AntLow", "AntHigh", "AntLow+hom", "AntHigh+hom"))
mdf$Replicate_num <- gsub("[^0-9]", "", mdf$Replicate)

p1 <- ggplot(mdf, aes(x = SampleID, y = CopyNum)) +
    theme_minimal() +
    geom_bar(stat="identity", color = "Black", fill = "indianred") +
    ylab("16S rRNA gene copies") +
    scale_y_log10(limits=c(1,1e9), breaks = c(1e1, 1e2, 1e3, 1e4, 1e5, 1e6, 1e7, 1e8)) +
    coord_cartesian(ylim = c(1e3,1e8)) +
    theme(
      axis.title.y = element_text(size = 10.5, face = "bold"),
      axis.text.y = element_text(size = 9, color = "black"),
      axis.title.x = element_blank(),
      axis.text.x = element_blank(), 
      axis.ticks.x = element_blank(),
      legend.position = "bottom",
      strip.text = element_text(size = 10, face = "bold") # Optionnel : harmonise les titres de facettes
    ) +
      facet_nested(.~ Treatment + Replicate_num, scales = "free_x", space = "free_x", nest_line = TRUE) + 
    annotation_logticks(base = 10, sides = "l", scaled = TRUE,
                short = unit(0.05, "cm"), mid = unit(0.1, "cm"), long = unit(0.2, "cm"),
                colour = "black", size = 0.5, linetype = 1, alpha = 1)

# Note: To save with specific dimensions, use ggsave
# ggsave("plot.pdf", p1, width = 10, height = 5)
print(p1, width = 100, height = 20)


```

## Retain only ASVs that have at least 1% relative abundance in minimum 1 sample

```{r}
library(genefilter)
mothur_proportions = transform_sample_counts(ps.noncontam.filt, function(x) {x/sum(x)})
Filtered = filter_taxa(mothur_proportions, filterfun(kOverA(1, 0.01)), TRUE)
```

## Plot bacterial stacked bars

```{r}
mdf = psmelt(Filtered)

mdf$Treatment <- factor(mdf$Treatment, 
                        levels = c("Ctr", "AntLow", "AntHigh", "AntLow+hom", "AntHigh+hom"), 
                        ordered = TRUE)
mdf$Replicate_num <- gsub("[^0-9]", "", mdf$Replicate)

library(RColorBrewer)


cbPalette<-colorRampPalette(brewer.pal(12, "Paired"))

set.seed(4) # Change the number within parentheses to change the order of colors in the plot. Here we use set.seed to make the colors reproducible, because in scale_fill_manual we use "sample" to shuffle the color order.
p2 = ggplot(mdf, aes_string(x = "SampleID", y = "Abundance", fill = "Genus")) +
        theme_minimal()+
        geom_bar(stat = "identity", position = "stack", color = "black") +
        ylab("Relative abundance") +
        facet_nested(.~ Treatment+Replicate_num, scales = "free") +
        theme(
          # Axes en gras et stylisés
          axis.title.y = element_text(size = 10.5, face = "bold"),
          axis.text.y = element_text(size = 10, color = "black"),
          axis.title.x = element_text(size = 10.5, face = "bold"),
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10, color = "black"),
          legend.position="bottom",
          legend.title = element_text(size = 10, face = "bold"),
          strip.text = element_text(size = 8.5, face = "bold")
        ) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position="bottom") +
        scale_fill_manual(values=sample(cbPalette(14)))

print(p2, width = 1000, height = 200)





# Get the ggplot grob
gt <- ggplotGrob(p2)

# Locate the tops of the plot panels
panels <- grep("panel", gt$layout$name)
top <- unique(gt$layout$t[panels])

# Remove the rows immediately above the plot panel
gt = gt[-(top-1), ]

# Draw it
grid.newpage()
grid.draw(gt)





```

## Plot qPCR and ampliseq bars together (and export to pdf)

```{r}
dev.new()
pdf(file="Barplots_Experimentalsamples.pdf",width=12,height=8, pointsize=4)

plot_grid(p1, gt, ncol = 1, nrow = , rel_widths=c(2,2), rel_heights = c(4,6), align = 'v')

dev.off()

plot_grid(p1, gt, ncol = 1, nrow = , rel_widths=c(2,2), rel_heights = c(4,6), align = 'v')
```

## Basic descriptive statistics

```{r}
library(phyloseq)
library(dplyr)
library(tidyr)

# 1. Transformer en abondance relative (%)
ps_rel <- transform_sample_counts(ps.noncontam.filt, function(x) 100 * x / sum(x))

# 2. Agglomérer au niveau Genre
ps_genus <- tax_glom(ps_rel, "Genus")

# 3. Transformer en data.frame pour calculs
mdf_rel <- psmelt(ps_genus)

# 4. Calculer Moyenne, Écart-type et Prévalence par Genre
genus_stats <- mdf_rel %>%
  group_by(Genus) %>%
  summarise(
    avg_abundance = mean(Abundance),
    sd_abundance = sd(Abundance),
    # Prévalence : % d'échantillons où le genre est présent (> 0.1%)
    prevalence = sum(Abundance > 0.1) / n_distinct(SampleID) * 100
  ) %>%
  arrange(desc(avg_abundance))

# Afficher les 10 premiers
print(head(genus_stats, 10))

```

## Calculate Number of ASV 

```{r}
ps_exp <- subset_samples(ps.noncontam.filt, 
                         !Treatment %in% c("BeforeExperiment", "GlycerolPBScontrol(FirstStep)", "GutHomogenate"))

otu_mat <- as(otu_table(ps_exp), "matrix")

if(taxa_are_rows(ps_exp)) {
  richness_vector <- colSums(otu_mat > 0)
} else {
  richness_vector <- rowSums(otu_mat > 0)
}

asv_stats <- data.frame(
    Mean_ASV   = mean(richness_vector),
    SD_ASV     = sd(richness_vector),
    Median_ASV = median(richness_vector),
    Min_ASV    = min(richness_vector),
    Max_ASV    = max(richness_vector)
)

print(asv_stats)
```
## Alpha Diversity 

```{r}
library(dplyr)

alpha_div <- estimate_richness(ps_exp, measures = c("Observed", "Shannon"))
df_alpha <- data.frame(alpha_div, sample_data(ps_exp))

alpha_stats <- df_alpha %>%
  group_by(Treatment) %>%
  summarise(
    N = n(),
    Mean_Observed = mean(Observed),
    SD_Observed   = sd(Observed),
    Mean_Shannon  = mean(Shannon),
    SD_Shannon    = sd(Shannon)
  ) %>%
  arrange(desc(Mean_Shannon))

alpha_stats

# Quick ANOVA to see if Treatment significantly affects Shannon
kruskal_test <- kruskal.test(Shannon ~ Treatment, data = df_alpha)
kruskal_test
```


## Let’s plot ordinations. Let's first combine qPCR and ampliseq data

```{r}
proportions = transform_sample_counts(ps.noncontam.filt, function(x) {x/sum(x)})

#calculate ratios
ratios = merge( tax_table(proportions),t(otu_table(proportions)), by="row.names")
copynum = data.frame(sample_data(proportions)[,"CopyNum"])

row.names(ratios) <- ratios[,1]

ratios <- ratios[,9:ncol(ratios)]

copies = data.frame(row.names = (row.names(ratios)))
for (i in 1:ncol(ratios)){
  numbers = ratios[,i]*copynum$CopyNum[i]  # or copynum$CopyNum_norm[i]
  copies = cbind(copies,numbers)
}
colnames(copies) = colnames(ratios)

rownames(samdf) <- samdf$SampleID

# Phyloseq object with the combined data 
ps_copies <- phyloseq(otu_table(copies, taxa_are_rows=T), 
               sample_data(samdf), 
               tax_table(proportions),
               phy_tree(proportions))

```

## Save ASV table

```{r}
write.table(otu_table(ps_copies), "LasioProject_ASVtable_qPCRnormalised.txt", sep="\t")
```

## Ordinations using qPCR-normalized counts for the whole dataset

```{r}
set.seed(91193)

# Ordinate using Principal Coordinate analysis
bee_pcoa <- ordinate(
  physeq = ps_copies, 
  method = "PCoA", 
  distance = "bray", 
  trymax = 100
)

# Scree plot
plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  type="scree"
)

# Plot PCoA with ellipses
p1 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "Treatment",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = Treatment), alpha = 1, size = 3) +
  stat_ellipse(aes(color = Treatment), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio.pdf",
  useDingbats = FALSE
)

# Plot PCoA with ellipses
p2 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "Antibiotic",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = Antibiotic), alpha = 1, size = 3) +
  stat_ellipse(aes(color = Antibiotic), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio_Antibiotic.pdf",
  useDingbats = FALSE
)

# Plot PCoA with ellipses
p3 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "AntibioticHigh",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = AntibioticHigh), alpha = 1, size = 3) +
  stat_ellipse(aes(color = AntibioticHigh), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio_AntibioticHigh.pdf",
  useDingbats = FALSE
)

# Plot PCoA with ellipses
p4 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "Homogenate",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = Homogenate), alpha = 1, size = 3) +
  stat_ellipse(aes(color = Homogenate), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio_Homogenate.pdf",
  useDingbats = FALSE
)
```

## Run a Permanova test with Adonis.

```{r}
set.seed(8650)

sink("Adonis-Betadisper_FourGroups.txt")
print("###############################################################")
print("Differences between four groups - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_copies, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_copies))

# Adonis test
adonis2(bray ~ Treatment, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$Treatment)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_FourGroups.txt")

sink("Adonis-Betadisper_AntibioticEffect.txt")
print("###############################################################")
print("Differences between antibiotic exposed and non exposed (any dose) - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_copies, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_copies))

# Adonis test
adonis2(bray ~ Antibiotic, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$Antibiotic)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_AntibioticEffect.txt")

sink("Adonis-Betadisper_AntibioticHigh.txt")
print("###############################################################")
print("Differences between antibiotic high and other treatments - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_copies, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_copies))

# Adonis test
adonis2(bray ~ AntibioticHigh, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$AntibioticHigh)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_AntibioticHigh.txt")

sink("Adonis-Betadisper_Homogenate.txt")
print("###############################################################")
print("Differences between homogenate recolonization Yes vs no - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_copies, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_copies))

# Adonis test
adonis2(bray ~ Homogenate, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$Homogenate)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_Homogenate.txt")

sink("Adonis-Betadisper_AntibioticEffect_interaction.txt")
print("###############################################################")
print("Differences between antibiotic exposed and non exposed (any dose) - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_copies, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_copies))

# Adonis test
adonis2(formula=bray ~ Antibiotic+Homogenate, by = "margin", data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$Antibiotic)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_AntibioticEffect_interaction.txt")

sink("Adonis-Betadisper_AntibioticHigh_interaction.txt")
print("###############################################################")
print("Differences between antibiotic high and rest - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_copies, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_copies))

# Adonis test
adonis2(formula=bray ~ AntibioticHigh+Homogenate, by = "margin", data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$AntibioticHigh)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_AntibioticHigh_interaction.txt")
```

#  Testing after controlling pseudoreplication

```{r}
library(phyloseq)
library(vegan)

# 1. Setup output file
sink("Adonis-Betadisper_AntibioticHigh_AbsoluteCounts.txt")
cat("###############################################################\n")
cat("Differences between treatment groups (AntibioticHigh * Homogenate) - Bray-Curtis dissimilarities (pseudoreplicates accounted for)\n")
cat("--- Centroid-based PERMANOVA ---\n")

# Assuming ps_copies is already loaded
# 1️⃣ Compute Bray-Curtis distance matrix
bray <- phyloseq::distance(ps_copies, method = "bray", weighted = TRUE)

# 2️⃣ Extract sample metadata
sampledf <- data.frame(sample_data(ps_copies))

# 3️⃣ Collapse pseudoreplicates: compute centroids per arena (Replicate)
# This results in one 'observation' (centroid) per Replicate
beta <- betadisper(bray, sampledf$Replicate)
centroids <- beta$centroids 

# 4️⃣ Build metadata for centroids
# Since treatment is consistent within Replicate, unique() works perfectly
meta_centroids <- unique(sampledf[, c("Replicate", "AntibioticHigh", "Homogenate")])

# Align metadata to the centroids' order
meta_centroids <- meta_centroids[match(rownames(centroids), meta_centroids$Replicate), ]

# Sanity check: ensure the number of rows matches
# if(nrow(meta_centroids) != nrow(centroids)) stop("Metadata alignment error!") 

# 5️⃣ PERMANOVA on collapsed data (centroids)
adonis_res <- adonis2(
 centroids ~ AntibioticHigh + Homogenate, # Interaction is not significant
 data = meta_centroids,
 permutations = 999,
 by = "margin" # Tests each term's unique contribution
)
print(adonis_res)

# ---
cat("\n###############################################################\n")
cat("--- Homogeneity of Dispersion Check (Betadisper) ---\n")

# 6️⃣ Optional: check homogeneity of dispersion on original samples
# NOTE: This is run on the original (non-collapsed) distances (bray) 
# and the combined factor (or a single factor like AntibioticHigh)
# It tests the assumption that the variance (spread) around the group mean is equal.
beta_disp <- betadisper(bray, sampledf$Treatment) # Use Treatment or the combined factor for a complete test
perm_res <- permutest(beta_disp)
print(perm_res)

sink()
file.show("Adonis-Betadisper_AntibioticHigh_AbsoluteCounts.txt")

```

# Some stats - checking which bacteria differ between treatments - Based only on relative abundance

```{r}
library(DESeq2)
# subset your phyloseq object
ps.exp <- subset_samples(ps.noncontam.filt, Sample_Type == "Experimental")

# (optional but recommended)
ps.exp <- prune_samples(sample_sums(ps.exp) > 0, ps.exp)

# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ Treatment)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_all <- results(dds)
res_all <- res_all[order(res_all$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_all <- res_all[res_all$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_all <- cbind(tax[rownames(sig_all), ], sig_all)

# Preview
tax_all

# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ Antibiotic)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_ant <- results(dds, contrast = c("Antibiotic", "Yes", "No"))
res_ant <- res_ant[order(res_ant$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_ant <- res_ant[res_ant$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_ant <- cbind(tax[rownames(sig_ant), ], sig_ant)

# Preview
tax_ant


# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ AntibioticHigh)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_anthigh <- results(dds, contrast = c("AntibioticHigh", "Yes", "No"))
res_anthigh <- res_anthigh[order(res_anthigh$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_anthigh <- res_anthigh[res_anthigh$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_anthigh <- cbind(tax[rownames(sig_anthigh), ], sig_anthigh)

# Preview
tax_anthigh
print(tax_anthigh)

# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ Homogenate)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_hom <- results(dds, contrast = c("Homogenate", "Yes", "No"))
res_hom <- res_hom[order(res_hom$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_hom <- res_hom[res_hom$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_hom <- cbind(tax[rownames(sig_hom), ], sig_hom)

# Preview
tax_hom


############################################################
# Volcano plot 
############################################################

library(ggplot2)
library(ggrepel)

volcano_df <- as.data.frame(res_ant)
volcano_df$Taxon <- rownames(volcano_df)
volcano_df$Significant <- ifelse(volcano_df$padj < 0.05, "Yes", "No")

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(aes(label = Taxon), size = 3, max.overlaps = 10) +
  scale_color_manual(values = c("grey70", "red")) +
  theme_minimal() +
  labs(
    title = "Any antibiotic dose Yes vs No — Differential Abundance",
    x = "Log2 Fold Change (Yes / No)",
    y = "-log10 adjusted p-value"
  )

volcano_df <- as.data.frame(res_anthigh)
volcano_df$Taxon <- rownames(volcano_df)
volcano_df$Significant <- ifelse(volcano_df$padj < 0.05, "Yes", "No")

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(aes(label = Taxon), size = 3, max.overlaps = 10) +
  scale_color_manual(values = c("grey70", "red")) +
  theme_minimal() +
  labs(
    title = "Antibiotic high dose Yes vs No — Differential Abundance",
    x = "Log2 Fold Change (Yes / No)",
    y = "-log10 adjusted p-value"
  )

volcano_df <- as.data.frame(res_hom)
volcano_df$Taxon <- rownames(volcano_df)
volcano_df$Significant <- ifelse(volcano_df$padj < 0.05, "Yes", "No")

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(aes(label = Taxon), size = 3, max.overlaps = 10) +
  scale_color_manual(values = c("grey70", "red")) +
  theme_minimal() +
  labs(
    title = "Homogenate reinoculation Yes vs No — Differential Abundance",
    x = "Log2 Fold Change (Yes / No)",
    y = "-log10 adjusted p-value"
  )





```

# Let's test if any ASV differs between treatment groups - Absolute abundances

```{r}
library(multcomp)

# --- 1. DATA PREP ---
copynums <- psmelt(ps_copies)
colnames(copynums)[colnames(copynums) == "OTU"] <- "Group"
copynums$Abundance[copynums$Abundance == 0] <- 1
copynums$log_cop <- log10(as.numeric(copynums$Abundance))
copynums$Treatment <- as.factor(copynums$Treatment)

# --- 2. PERMUTATION ANOVA FUNCTION ---
anovaFF <- function(rpse, v1, nrep) {
  lmer <- lm(rpse ~ v1) 
  obs.F_real <- anova(lmer)$F[1] 
  obs.F_sim <- numeric(nrep)
  obs.F_sim[1] <- obs.F_real 
  
  for (i in 2:nrep) {
    rpse_shuffled <- sample(rpse)
    obs.F_sim[i] <- anova(lm(rpse_shuffled ~ v1))$F[1] 
  }
  
  pval <- sum(obs.F_sim >= obs.F_real) / nrep
  return(pval)
}

# --- 3. RESULTS STORAGE ---
results_summary <- data.frame(
  ASV = character(),
  anova_pval = numeric(),
  anova_sig = logical(),
  any_pairwise_sig = logical(),
  stringsAsFactors = FALSE
)

# store ALL pairwise tests (important!)
sig_pairs_all <- data.frame(
  ASV = character(),
  group1 = character(),
  group2 = character(),
  p_bonf = numeric(),   # from Tukey (Bonferroni)
  stringsAsFactors = FALSE
)

# helper
p_to_sig <- function(p) {
  if      (p < 0.001) "***"
  else if (p < 0.01)  "**"
  else if (p < 0.05)  "*"
  else if (p < 0.1)   "."
  else                "ns"
}

# --- 4. MAIN LOOP ---
t <- levels(as.factor(copynums$Group))

sink("Tukey-ASVstats.txt") 
cat("--- ASV STATISTICAL SUMMARY ---\n\n")

for (ii in t){ 
  
  dsub = copynums[copynums$Group == ii, ]  
  
  if(var(dsub$log_cop) == 0) next 
  
  p_val <- anovaFF(dsub$log_cop, dsub$Treatment, nrep = 10000)
  
  cat("=========================================\n")
  cat("ASV:", ii, "\n")
  cat("ANOVA Permutation P-value:", p_val, "\n")
  
  anova_sig <- p_val < 0.05
  any_pairwise_sig <- FALSE
  
  if (anova_sig) {
    cat("--> SIGNIFICANT! Running Tukey Post-Hoc...\n")
    
    mod <- lm(log_cop ~ Treatment, data = dsub)
    
    tryCatch({
      res <- summary(
        glht(mod, linfct = mcp(Treatment = "Tukey")),
        test = adjusted(type = "BH")
      )
      
      print(res)
      
      pairwise_pvals <- res$test$pvalues
      contrast_names <- names(res$test$coefficients)
      
      # store ALL pairwise comparisons
      for (j in seq_along(contrast_names)) {
        
        p_bonf <- pairwise_pvals[j]
        
        parts  <- strsplit(contrast_names[j], " - | == 0")[[1]]
        group1 <- trimws(parts[1])
        group2 <- trimws(parts[2])
        
        sig_pairs_all <- rbind(sig_pairs_all, data.frame(
          ASV    = ii,
          group1 = group1,
          group2 = group2,
          p_bonf = p_bonf,
          stringsAsFactors = FALSE
        ))
      }
      
      any_pairwise_sig <- any(pairwise_pvals < 0.05)
      
    }, error = function(e) {
      cat("Tukey failed: ", conditionMessage(e), "\n")
    })
    
  } else {
    cat("--> Not significant. Skipping Tukey.\n")
  }
  
  cat("\n")
  
  results_summary <- rbind(results_summary, data.frame(
    ASV = ii,
    anova_pval = p_val,
    anova_sig = anova_sig,
    any_pairwise_sig = any_pairwise_sig,
    stringsAsFactors = FALSE
  ))
}

sink()

# --- 5. FDR CORRECTION (ANOVA LEVEL) ---
results_summary$anova_pval_fdr <- p.adjust(results_summary$anova_pval, method = "BH")
results_summary$anova_sig      <- results_summary$anova_pval_fdr < 0.05

# --- 6. FDR CORRECTION (PAIRWISE LEVEL, GLOBAL) ---
sig_pairs_all$p_fdr <- p.adjust(sig_pairs_all$p_bonf, method = "BH")
sig_pairs_all$sig_fdr <- sig_pairs_all$p_fdr < 0.05
sig_pairs_all$sig_symbol <- sapply(sig_pairs_all$p_fdr, p_to_sig)

# OPTIONAL: keep only FDR-significant pairs
sig_pairs_sig <- sig_pairs_all[sig_pairs_all$sig_fdr, ]

# --- 7. SUMMARY ---
total_tested      <- nrow(results_summary)
n_anova_sig       <- sum(results_summary$anova_sig)
n_pairwise_sig    <- length(unique(sig_pairs_sig$ASV))

asvs_anova_sig    <- results_summary$ASV[results_summary$anova_sig]
asvs_pairwise_sig <- unique(sig_pairs_sig$ASV)

cat("=== SUMMARY ===\n")
cat("Total ASVs tested:                        ", total_tested, "\n")
cat("ASVs with ANOVA (FDR < 0.05):             ", n_anova_sig, "\n")
cat("  ->", paste(asvs_anova_sig, collapse = ", "), "\n\n")
cat("ASVs with >=1 FDR-significant pairwise:   ", n_pairwise_sig, "\n")
cat("  ->", paste(asvs_pairwise_sig, collapse = ", "), "\n")

# --- 8. EXPORT ---
write.csv(results_summary, "ASV_stats_summary.csv", row.names = FALSE)
write.csv(sig_pairs_all,   "ASV_all_pairs.csv",     row.names = FALSE)
write.csv(sig_pairs_sig,   "ASV_sig_pairs.csv",     row.names = FALSE)

```


## Who are the ASV that significantly changed ? anova < 0.05 .

```{r}

head(target_asvs)
head(taxa_names(ps_copies))

## Who are the ASV that significantly changed? We plot the ones with ANOVA < 0.05.
sig_summary <- read.csv("ASV_stats_summary.csv")

# Explicitly use FDR column
target_asvs <- sig_summary$ASV[sig_summary$anova_pval_fdr < 0.05]

tax_info <- as.data.frame(tax_table(ps_copies)[target_asvs, ])
print(tax_info)

write.csv(tax_info, "ASV_FDR_Significant_Taxonomy.csv")

```

## Plot indvidual anova < 0.05 + at least one pairwise comparison that is p<0.05 

```{r}
## Plot individual differences — ASVs with ANOVA p < 0.05 AND ≥1 significant pairwise comparison
library(ggplot2)
library(dplyr)
library(tidyr)

# --- 1. CHARGEMENT DES DONNÉES ---
# On utilise les fichiers générés par l'analyse statistique précédente
sig_summary   <- read.csv("ASV_stats_summary.csv")
sig_pairs_all <- read.csv("ASV_sig_pairs.csv") # Ce fichier contient les colonnes 'sig_symbol'

# FILTRE : Uniquement les ASV qui sont significatifs en ANOVA ET qui ont au moins un Tukey significatif
target_asvs <- sig_summary$ASV[sig_summary$anova_sig == TRUE & sig_summary$any_pairwise_sig == TRUE]

# --- 2. PRÉPARATION DES LABELS (TAXONOMIE + P-VALUE ANOVA) ---
tax_df <- as.data.frame(as(tax_table(ps_copies), "matrix"))
tax_df$Group <- rownames(tax_df)
tax_df <- tax_df %>%
  mutate(Genus = replace_na(Genus, "Unknown Genus")) %>%
  left_join(sig_summary[, c("ASV", "anova_pval")], by = c("Group" = "ASV")) %>%
  mutate(
    anova_label = case_when(
      anova_pval < 0.001 ~ "p < 0.001",
      !is.na(anova_pval) ~ paste0("p = ", round(anova_pval, 3)),
      TRUE               ~ ""
    ),
    ASV_Label = paste0(Group, " (", Genus, ")\n", anova_label)
  )

# --- 3. PRÉPARATION DES DONNÉES DU GRAPHIQUE ---
df_plot <- copynums %>%
  filter(Group %in% target_asvs) %>%
  mutate(Replicate_num = as.factor(gsub("[^0-9]", "", Replicate))) %>%
  left_join(tax_df[, c("Group", "ASV_Label")], by = "Group") %>%
  mutate(Treatment = factor(Treatment, levels = c("Ctr", "AntLow", "AntHigh", "AntLow+hom", "AntHigh+hom")))

# --- 4. CONSTRUCTION DES CROCHETS DE SIGNIFICATIVITÉ (BRACKETS) ---
treatment_levels <- c("Ctr", "AntLow", "AntHigh", "AntLow+hom", "AntHigh+hom")
x_pos <- setNames(seq_along(treatment_levels), treatment_levels)

sig_brackets <- do.call(rbind, lapply(target_asvs, function(asv) {
  pairs <- sig_pairs_all[sig_pairs_all$ASV == asv, ]
  if (nrow(pairs) == 0) return(NULL)

  label   <- unique(df_plot$ASV_Label[df_plot$Group == asv])
  y_max   <- max(df_plot$log_cop[df_plot$Group == asv], na.rm = TRUE)
  y_range <- diff(range(df_plot$log_cop[df_plot$Group == asv], na.rm = TRUE))
  # Ajustement de l'espace vertical entre les barres
  step    <- max(y_range * 0.15, 0.2) 

  pairs$x1   <- x_pos[pairs$group1]
  pairs$x2   <- x_pos[pairs$group2]
  pairs$span <- abs(pairs$x2 - pairs$x1)
  pairs      <- pairs[order(pairs$span), ]

  do.call(rbind, lapply(seq_len(nrow(pairs)), function(i) {
    data.frame(
      Group     = asv,
      ASV_Label = label,
      x_left    = min(pairs$x1[i], pairs$x2[i]),
      x_right   = max(pairs$x1[i], pairs$x2[i]),
      y_bracket = y_max + step * i,
      sig       = pairs$sig_symbol[i], # <-- CORRECTION : on utilise sig_symbol
      stringsAsFactors = FALSE
    )
  }))
}))


p_sig_pairs <- ggplot(df_plot, aes(x = Treatment, y = log_cop)) +
  # Boxplot en premier plan pour qu'il soit bien visible
  geom_boxplot(outlier.shape = NA, fill = NA, color = "black", size = 0.5) +
  # Points plus transparents (alpha = 0.4) avec une seed fixe pour la reproductibilité
  geom_point(aes(color = Replicate_num), alpha = 0.4, size = 2, position = position_jitter(width = 0.2, seed = 42)) + 
  
  # Crochets de significativité
  geom_segment(data = sig_brackets,
               aes(x = x_left, xend = x_right, y = y_bracket, yend = y_bracket),
               inherit.aes = FALSE, color = "black", linewidth = 0.4) +
  geom_segment(data = sig_brackets,
               aes(x = x_left, xend = x_left, y = y_bracket - 0.05, yend = y_bracket),
               inherit.aes = FALSE, color = "black", linewidth = 0.4) +
  geom_segment(data = sig_brackets,
               aes(x = x_right, xend = x_right, y = y_bracket - 0.05, yend = y_bracket),
               inherit.aes = FALSE, color = "black", linewidth = 0.4) +
  
  # Étoiles de significativité
  geom_text(data = sig_brackets,
            aes(x = (x_left + x_right) / 2, y = y_bracket + 0.05, label = sig),
            inherit.aes = FALSE, size = 5, fontface = "bold") +
  
  facet_wrap(~ ASV_Label, scales = "free_y", ncol = 3) +
  theme_classic() + 
  theme(
    strip.background = element_blank(), 
    strip.text = element_text(face = "bold.italic", size = 11),
    axis.line = element_line(color = "black"),
    axis.title = element_text(size = 12, face = "bold"),
    panel.grid.major.y = element_line(color = "grey95"),
    legend.position = "none" # Légende supprimée
  ) +
  labs(
    x = "Experimental Treatment",
    y = expression(bold(Log[10]) ~ bold(Absolute ~ Abundance))
  ) +
  scale_color_brewer(palette = "Dark2")

print(p_sig_pairs)

ggsave("Significant_ASVs_FDR_Boxplots.png", plot = p_sig_pairs, width = 12, height = 8, dpi = 300)

```



## Extra-tests // Correlation between the number of Wolbachia and the absolute number of genes 
```{r}
library(phyloseq)
library(ggplot2)
library(ggpubr) #

# 1. Isoler la quantité de Wolbachia par échantillon
# On somme tous les ASVs identifiés comme Wolbachia
ps_wol <- subset_taxa(ps_copies, Genus == "Wolbachia")
wol_counts <- sample_sums(ps_wol)

# 2. Calculer le total de bactéries (Total Load) par échantillon
total_load <- sample_sums(ps_copies)

# 3. Créer un data frame pour la corrélation
df_corr <- data.frame(
  SampleID = names(total_load),
  Wolbachia_Load = wol_counts,
  Total_Bacterial_Load = total_load
)

# 4. Appliquer le log10 (avec +1 pour éviter log(0))
df_corr$log_Wol <- log10(df_corr$Wolbachia_Load + 1)
df_corr$log_Total <- log10(df_corr$Total_Bacterial_Load + 1)

# 5. Test de corrélation de Spearman
spearman_test <- cor.test(df_corr$log_Wol, df_corr$log_Total, method = "spearman")
print(spearman_test)

# 6. Visualisation
p_corr <- ggplot(df_corr, aes(x = log_Wol, y = log_Total)) +
  geom_point(aes(color = log_Total), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  stat_cor(method = "spearman", label.x = 2, label.y = max(df_corr$log_Total)) +
  labs(
    title = "Corrélation entre Wolbachia et Charge Bactérienne Totale",
    x = "Log10(Abondance Wolbachia)",
    y = "Log10(Total qPCR)"
  ) +
  theme_minimal()

p_corr
# ggsave(p_corr,
#   height = 5, width = 7, dpi = 300,
#   filename = "corr2.pdf",
#   useDingbats = FALSE
# )
```

## Correlation test between the relative propotion and the absolute number of genes
```{r}
# Calcul de la proportion relative (0 à 1)
df_corr$Relative_Wol <- df_corr$Wolbachia_Load / df_corr$Total_Bacterial_Load
df_corr$Percent_Wol <- df_corr$Relative_Wol * 100

# Test de Spearman entre la charge totale et la proportion
spearman_prop <- cor.test(df_corr$log_Total, df_corr$Relative_Wol, method = "spearman")
print(spearman_prop)

p1 <- ggplot(df_corr, aes(x = log_Total, y = Percent_Wol)) +
  geom_point(color = "darkorange", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", linetype = "solid") + # "loess" pour voir la tendance non-linéaire
  stat_cor(method = "spearman") +
  labs(
    title = "Influence de la Charge Totale sur la Dominance de Wolbachia",
    x = "Log10(Charge Bactérienne Totale)",
    y = "Proportion de Wolbachia (%)"
  ) +
  theme_minimal()

p1

# ggsave(
#   height = 5, width = 7, dpi = 300,
#   filename = "corr.pdf",
#   useDingbats = FALSE
# )
```


#Let's try to filter out Wolbachia and run analyses again

```{r}
# Filter out Wolbachia
ps_no_wolbachia <- subset_taxa(ps.noncontam.filt, Genus != "Wolbachia")
```

## Retain only ASVs that have at least 1% relative abundance in minimum 1 sample

```{r}
library(genefilter)
mothur_proportions = transform_sample_counts(ps_no_wolbachia, function(x) {x/sum(x)})
Filtered = filter_taxa(mothur_proportions, filterfun(kOverA(3, 0.01)), TRUE)
```

## Plot bacterial stacked bars

```{r}
mdf = psmelt(Filtered)

mdf <- mdf[order(factor(mdf$Treatment, levels = c("Ctr", "AntLow", "AntHigh", "AntLow+hom", "AntHigh+hom"), ordered=TRUE), mdf$Sample), ]  # Let's first sort by Treatment, then hive, then sample ID

library(RColorBrewer)
cbPalette<-colorRampPalette(brewer.pal(12, "Paired"))

set.seed(4) # Change the number within parentheses to change the order of colors in the plot. Here we use set.seed to make the colors reproducible, because in scale_fill_manual we use "sample" to shuffle the color order.
p2 = ggplot(mdf, aes_string(x = "SampleID", y = "Abundance", fill = "Genus")) +
        geom_bar(stat = "identity", position = "stack", color = "black") +
        ylab("Relative abundance") +
        facet_grid(.~ Treatment, scales = "free") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position="bottom") +
        scale_fill_manual(values=sample(cbPalette(13)))

print(p2, width = 1000, height = 200)

```

## Let’s plot ordinations

```{r}
proportions = transform_sample_counts(ps_no_wolbachia, function(x) {x/sum(x)})

ps_copies = proportions 

```

## Save ASV table

```{r}
# write.table(otu_table(ps_copies), "BeeTracking_ASVtable_qPCRnormalised_Filtered.txt", sep="\t")
```

## Ordinations using qPCR-normalized counts for the whole dataset

```{r}
set.seed(91193)

# Ordinate using Principal Coordinate analysis
bee_pcoa <- ordinate(
  physeq = ps_copies, 
  method = "PCoA", 
  distance = "bray", 
  trymax = 100
)

# Scree plot
plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  type="scree"
)

# Plot PCoA with ellipses
p1 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "Treatment",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = Treatment), alpha = 1, size = 3) +
  stat_ellipse(aes(color = Treatment), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio_NoWolbachia.pdf",
  useDingbats = FALSE
)

# Plot PCoA with ellipses
p2 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "Antibiotic",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = Antibiotic), alpha = 1, size = 3) +
  stat_ellipse(aes(color = Antibiotic), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio_Antibiotic_NoWolbachia.pdf",
  useDingbats = FALSE
)

# Plot PCoA with ellipses
p3 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "AntibioticHigh",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = AntibioticHigh), alpha = 1, size = 3) +
  stat_ellipse(aes(color = AntibioticHigh), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio_AntibioticHigh_NoWolbachia.pdf",
  useDingbats = FALSE
)

# Plot PCoA with ellipses
p4 <- plot_ordination(
  physeq = ps_copies,
  ordination = bee_pcoa,
  label = "Sample_ID",
  axes = c(1,2),
  color = "Homogenate",
  title = "PCoA of Bray-Curtis dissimilarities"
) +
  scale_fill_viridis_b() +
  theme_bw() +
  geom_point(aes(color = Homogenate), alpha = 1, size = 3) +
  stat_ellipse(aes(color = Homogenate), linewidth = 1)

ggsave(
  height = 5, width = 7, dpi = 300,
  filename = "Ordinations_PCoA_Bray-Lasio_Homogenate_NoWolbachia.pdf",
  useDingbats = FALSE
)
```

## Run a Permanova test with Adonis.

```{r}
set.seed(8650)

sink("Adonis-Betadisper_FourGroups_NoWolbachia.txt")
print("###############################################################")
print("Differences between four groups - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_no_wolbachia, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_no_wolbachia))

# Adonis test
adonis2(bray ~ Treatment, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$Treatment)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_FourGroups_NoWolbachia.txt")

sink("Adonis-Betadisper_AntibioticEffect_NoWolbachia.txt")
print("###############################################################")
print("Differences between antibiotic exposed and non exposed (any dose) - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_no_wolbachia, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_no_wolbachia))

# Adonis test
adonis2(bray ~ Antibiotic, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$Antibiotic)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_AntibioticEffect_NoWolbachia.txt")

sink("Adonis-Betadisper_AntibioticHigh_NoWolbachia.txt")
print("###############################################################")
print("Differences between antibiotic high and other treatments - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_no_wolbachia, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_no_wolbachia))

# Adonis test
adonis2(bray ~ AntibioticHigh, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$AntibioticHigh)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_AntibioticHigh_NoWolbachia.txt")

sink("Adonis-Betadisper_Homogenate_NoWolbachia.txt")
print("###############################################################")
print("Differences between homogenate recolonization Yes vs no - Bray-Curtis dissimilarities")
# Calculate bray curtis distance matrix
bray <- phyloseq::distance(ps_no_wolbachia, method = "bray", weighted=TRUE)

# make a data frame from the scaled sample_data
sampledf <- data.frame(sample_data(ps_no_wolbachia))

# Adonis test
adonis2(bray ~ Homogenate, data = sampledf)

# test of Homegeneity of dispersion
beta <- betadisper(bray, sampledf$Homogenate)

# run a permutation test to get a statistic and a significance score
permutest(beta)

sink()
file.show("Adonis-Betadisper_Homogenate_NoWolbachia.txt")
```

# Some stats - checking which bacteria differ between treatments

```{r}
library(DESeq2)
# subset your phyloseq object
ps.exp <- subset_samples(ps_no_wolbachia, Sample_Type == "Experimental")

# (optional but recommended)
ps.exp <- prune_samples(sample_sums(ps.exp) > 0, ps.exp)

# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ Treatment)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_all <- results(dds)
res_all <- res_all[order(res_all$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_all <- res_all[res_all$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_all <- cbind(tax[rownames(sig_all), ], sig_all)

# Preview
tax_all

# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ Antibiotic)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_ant <- results(dds, contrast = c("Antibiotic", "Yes", "No"))
res_ant <- res_ant[order(res_ant$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_ant <- res_ant[res_ant$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_ant <- cbind(tax[rownames(sig_ant), ], sig_ant)

# Preview
tax_ant


# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ AntibioticHigh)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_anthigh <- results(dds, contrast = c("AntibioticHigh", "Yes", "No"))
res_anthigh <- res_anthigh[order(res_anthigh$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_anthigh <- res_anthigh[res_anthigh$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_anthigh <- cbind(tax[rownames(sig_anthigh), ], sig_anthigh)

# Preview
tax_anthigh

# Convert phyloseq to DESeq2 object
dds <- phyloseq_to_deseq2(ps.exp, ~ Homogenate)

dds <- estimateSizeFactors(dds, type = "poscounts")
dds <- DESeq(dds)

# Get results for S vs C
res_hom <- results(dds, contrast = c("Homogenate", "Yes", "No"))
res_hom <- res_hom[order(res_hom$padj, na.last = NA), ]

# Significant taxa (FDR < 0.05)
sig_hom <- res_hom[res_hom$padj < 0.05, ]

# Combine with taxonomy for interpretation
tax <- as.data.frame(tax_table(ps.exp))
tax_hom <- cbind(tax[rownames(sig_hom), ], sig_hom)

# Preview
tax_hom


############################################################
# Volcano plot 
############################################################

library(ggplot2)
library(ggrepel)

volcano_df <- as.data.frame(res_ant)
volcano_df$Taxon <- rownames(volcano_df)
volcano_df$Significant <- ifelse(volcano_df$padj < 0.05, "Yes", "No")

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(aes(label = Taxon), size = 3, max.overlaps = 10) +
  scale_color_manual(values = c("grey70", "red")) +
  theme_minimal() +
  labs(
    title = "Any antibiotic dose Yes vs No — Differential Abundance",
    x = "Log2 Fold Change (Yes / No)",
    y = "-log10 adjusted p-value"
  )

volcano_df <- as.data.frame(res_anthigh)
volcano_df$Taxon <- rownames(volcano_df)
volcano_df$Significant <- ifelse(volcano_df$padj < 0.05, "Yes", "No")

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(aes(label = Taxon), size = 3, max.overlaps = 10) +
  scale_color_manual(values = c("grey70", "red")) +
  theme_minimal() +
  labs(
    title = "Antibiotic high dose Yes vs No — Differential Abundance",
    x = "Log2 Fold Change (Yes / No)",
    y = "-log10 adjusted p-value"
  )

volcano_df <- as.data.frame(res_hom)
volcano_df$Taxon <- rownames(volcano_df)
volcano_df$Significant <- ifelse(volcano_df$padj < 0.05, "Yes", "No")

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(aes(label = Taxon), size = 3, max.overlaps = 10) +
  scale_color_manual(values = c("grey70", "red")) +
  theme_minimal() +
  labs(
    title = "Homogenate reinoculation Yes vs No — Differential Abundance",
    x = "Log2 Fold Change (Yes / No)",
    y = "-log10 adjusted p-value"
  )

```

# Alpha diversity

```{r}
plot_richness(ps.noncontam.filt, x="Treatment", color="Treatment", measures=c("Chao1", "Shannon", "Simpson"))

alpha_df <- estimate_richness(ps.noncontam.filt, measures = c("Chao1", "Shannon"))
alpha_df$SampleID <- rownames(alpha_df)

meta <- data.frame(sample_data(ps.noncontam.filt))
meta$SampleID <- rownames(meta)

alpha_merged <- merge(alpha_df, meta, by = "SampleID")

model_chao1 <- lm(Chao1 ~ AntibioticHigh + Homogenate, data = alpha_merged)
anova(model_chao1)

model_shannon <- lm(Shannon ~ AntibioticHigh + Homogenate, data = alpha_merged)
anova(model_shannon)


```

# Without Wolbachia

```{r}
plot_richness(ps_no_wolbachia, x="Treatment", color="Treatment", measures=c("Chao1", "Shannon", "Simpson"))

alpha_df <- estimate_richness(ps_no_wolbachia, measures = c("Chao1", "Shannon", "Simpson"))
alpha_df$SampleID <- rownames(alpha_df)

meta <- data.frame(sample_data(ps_no_wolbachia))
meta$SampleID <- rownames(meta)

alpha_merged <- merge(alpha_df, meta, by = "SampleID")

model_chao1 <- lm(Chao1 ~ AntibioticHigh + Homogenate, data = alpha_merged)
anova(model_chao1)

model_shannon <- lm(Shannon ~ AntibioticHigh + Homogenate, data = alpha_merged)
anova(model_shannon)

model_simpson <- lm(Simpson ~ AntibioticHigh + Homogenate, data = alpha_merged)
anova(model_simpson)
```

## To save the session

```{r}
save.image("/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/MelinaProject_Ampliseq.rdata")
# Close R, Re-open R
#load("/Volumes/s-genev-liberti/users/Liberti/Lmalachurum/Amplicon-sequencing/MelinaProject_Ampliseq.rdata")
```

## Session info

```{r}
sessionInfo()
```

```         
## R version 4.1.0 (2021-05-18)
## Platform: x86_64-apple-darwin17.0 (64-bit)
## Running under: macOS Mojave 10.14.6
## 
## Matrix products: default
## BLAS:   /Library/Frameworks/R.framework/Versions/4.1/Resources/lib/libRblas.dylib
## LAPACK: /Library/Frameworks/R.framework/Versions/4.1/Resources/lib/libRlapack.dylib
## 
## locale:
## [1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
## 
## attached base packages:
## [1] stats     graphics  grDevices utils     datasets  methods   base     
## 
## loaded via a namespace (and not attached):
##  [1] compiler_4.1.0  magrittr_2.0.1  fastmap_1.1.0   tools_4.1.0    
##  [5] htmltools_0.5.2 yaml_2.2.1      stringi_1.7.6   rmarkdown_2.11 
##  [9] knitr_1.36      stringr_1.4.0   xfun_0.28       digest_0.6.29  
## [13] rlang_0.4.12    evaluate_0.14
```
