---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,          
  message = FALSE,      
  warning = FALSE 
)
```

# Behavior and activity statistics
## Load library

```{r}
library(here)         # Relative file path tracking

library(tidyverse)    
library(readr)        
library(dplyr)        
library(tidyr)        

library(lme4)         
library(lmerTest)     
library(glmmTMB)      

library(DHARMa)      
library(car)         
library(emmeans)     

library(vegan)        
library(dimensio)    

library(ggplot2)      
library(ggbeeswarm)
library(patchwork)
library(scales)

set.seed(67)
options(contrasts = c("contr.sum", "contr.poly"))
```

# Data pre-processing 
### Behavioral data
```{r}
categories_all <- list(
  control  = c(5, 8, 13, 40, 41, 42, 43),
  low_ant  = c(2, 14, 15, 16, 24, 25, 26, 27),
  high_ant = c(3, 7, 10, 18, 28, 29, 30, 31),
  low_gut  = c(1, 6, 12, 20, 21, 22, 23, 34),
  high_gut = c(4, 11, 17, 19, 32, 33),
  gut_ctrl = c(35, 36, 37, 38, 39) # 2025 dataset only
)

category_conditions <- tibble(
  treatment    = c('control', 'low_ant', 'high_ant', 'low_gut', 'high_gut', 'gut_ctrl'),
  antibiotic  = c('zero',    'low',     'high',     'low',     'high',     'zero'),
  reinoculation = c('no',      'no',      'no',       'yes',     'yes',      'yes')
)

data_list <- list()
counter <- 1
for (cat_name in names(categories_all)) {
  cat_files <- categories_all[[cat_name]]
  for (x in cat_files) {
    # Determine the experimental year based on the dish index threshold
    year <- if_else(x <= 15, 2024, 2025)
    file_name <- paste0(year, "_petridish", x, ".csv")
    full_path <- here::here("activity-and-behavior", "data", "behavior_csv_raw", file_name) 
    if (!file.exists(full_path)) {
      warning(paste("File not found:", file_name))
      next
    }
    
    df <- read_csv(full_path, show_col_types = FALSE)
    df_filtered <- df %>% filter(time < 1800) #we get exactly 30min
    
    bump_count      <- sum(df_filtered$class == 1, na.rm = TRUE)
    touch_count     <- sum(df_filtered$class == 2, na.rm = TRUE)
    avoidance_count <- sum(df_filtered$class == 5, na.rm = TRUE)
    head_count      <- sum(df_filtered$class == 6, na.rm = TRUE)
    
    data_list[[counter]] <- tibble(
      `petridish number` = x,
      treatment = cat_name,
      year = year,
      `nb of bump` = bump_count,
      `nb of avoidance` = avoidance_count,
      `nb of head-to-head` = head_count,
      `nb of touch` = touch_count
    )
    counter <- counter + 1
  }
}

if (length(data_list) > 0) {
  df_behavior <- bind_rows(data_list) %>% 
    left_join(category_conditions, by = "treatment") %>% 
    dplyr::select(
      `petridish number`, treatment, antibiotic, reinoculation, year,
      `nb of bump`, `nb of avoidance`, `nb of head-to-head`, `nb of touch`
    )
}

colnames(df_behavior)<- c("arena", "treatment", "antibiotic", "reinoculation",
                   "year", "nb.of.bump", "nb.of.avoidance", "nb.of.head.to.head",
                   "nb.of.touch")

df_behavior <- df_behavior %>%
  mutate(treatment = dplyr::recode(
      treatment,
      control  = "zero_no",
      low_ant  = "low_no",
      high_ant = "high_no",
      low_gut  = "low_yes",
      high_gut = "high_yes",
      gut_ctrl = "zero_yes"
    )
  )

df_behavior$antibiotic <- factor(df_behavior$antibiotic, levels = c("zero", "low", "high"))
df_behavior$reinoculation <- factor(df_behavior$reinoculation)
df_behavior$reinoculation <- relevel(df_behavior$reinoculation, ref = "no")
df_behavior$year <- factor(df_behavior$year)

df_behavior
```

### Inter-individual distance data
```{r}
data_dir <- here::here("activity-and-behavior", "data", "animal_ta")
file_path <- file.path(data_dir, "inderindividual_dist.csv")
           
df_interind <- read_delim(file_path,
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

colnames(df_interind)<- c("treatment", "antibiotic", "reinoculation",
                   "year", "arena", "mean.dist")

df_interind <- df_interind %>%
  mutate(treatment = factor(treatment),      
    treatment = relevel(treatment, ref = "zero_no"),
    antibiotic = factor(antibiotic, levels = c("zero", "low", "high")),
    reinoculation = factor(reinoculation),
    reinoculation = relevel(reinoculation, ref = "no"),
    year = factor(year)   
  )
df_interind
```

### Activity data
```{r}
data_dir <- here::here("activity-and-behavior", "data", "animal_ta")
file_path <- file.path(data_dir, "moving.csv")

df_move <- read_delim(file = file_path,
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

colnames(df_move)<- c("treatment", "antibiotic", "reinoculation",
                   "year", "arena", "prop.time.moving", "average.speed")

df_move <- df_move %>%
  mutate(treatment = factor(treatment), 
    treatment = relevel(treatment, ref = "zero_no"),
    antibiotic = factor(antibiotic, levels = c("zero", "low", "high")),
    reinoculation = factor(reinoculation),
    reinoculation = relevel(reinoculation, ref = "no"),
    year = factor(year), 
    arena = factor(arena)
  )
df_move
```


### PCA on the 7 metrics 

```{r}
df_move_aggregated <- df_move %>%
  group_by(year, arena, treatment, antibiotic, reinoculation) %>%
  summarise(moy_prop_time_moving = mean(prop.time.moving, na.rm = TRUE),
    moy_average_speed    = mean(average.speed, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(arena = as.numeric(as.character(arena)))

df_behavior <- df_behavior %>%
  mutate(arena = as.numeric(arena))

df_interind <- df_interind %>%
  mutate(arena = as.numeric(as.character(arena)))

df_pca_7 <- df_move_aggregated %>%
  left_join(df_behavior, by = c("year", "arena", "antibiotic", "reinoculation", "treatment")) %>%
  left_join(df_interind, by = c("year", "arena", "antibiotic", "reinoculation", "treatment")) %>%
  mutate(
    year        = factor(year),
    arena       = factor(arena),
    treatment   = factor(treatment),
    treatment   = relevel(treatment, ref = "zero_no"),
    antibiotic  = factor(antibiotic, levels = c("zero", "low", "high")),
    reinoculation = factor(reinoculation, levels = c("no", "yes"))
  )

df_pca_7
```

# ==============================
Now that the data is correctly pre-processed, we can start the EDA and the analysis 

## Behavioral analysis
### Basic stats
```{r}
moyenne_bump <- tapply(df_behavior$nb.of.bump, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_bump <- tapply(df_behavior$nb.of.bump, df_behavior$treatment, sd, na.rm = TRUE)

resultats_bump <- data.frame(
  Moyenne = moyenne_bump,
  Ecart_Type = ecarts_bump
)

moyenne_avoidance <- tapply(df_behavior$nb.of.avoidance, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_avoidance <- tapply(df_behavior$nb.of.avoidance, df_behavior$treatment, sd, na.rm = TRUE)

resultats_avoidance <- data.frame(
  Moyenne = moyenne_avoidance,
  Ecart_Type = ecarts_avoidance
)

moyenne_head_to_head <- tapply(df_behavior$nb.of.head.to.head, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_head_to_head <- tapply(df_behavior$nb.of.head.to.head, df_behavior$treatment, sd, na.rm = TRUE)

resultats_head_to_head <- data.frame(
  Moyenne = moyenne_head_to_head,
  Ecart_Type = ecarts_head_to_head
)

moyenne_touch <- tapply(df_behavior$nb.of.touch, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_touch <- tapply(df_behavior$nb.of.touch, df_behavior$treatment, sd, na.rm = TRUE)

resultats_touch <- data.frame(
  Moyenne = moyenne_touch,
  Ecart_Type = ecarts_touch
)

formater_stats <- function(moyenne, et) {
  paste0(round(moyenne, 1), " \u00B1 ", round(et, 1)) # \u00B1 is the symbol +/-
}

tableau_publication <- data.frame(
  Category = rownames(resultats_bump), 
  stringsAsFactors = FALSE
)

tableau_publication$`Random bump` <- formater_stats(
  moyenne_bump, 
  ecarts_bump
)

tableau_publication$`Short antennal touch` <- formater_stats(
  moyenne_touch, 
  ecarts_touch
)

tableau_publication$`Head-to-head interaction` <- formater_stats(
  moyenne_head_to_head, 
  ecarts_head_to_head
)

tableau_publication$Avoidance <- formater_stats(
  moyenne_avoidance, 
  ecarts_avoidance
)

colnames(tableau_publication)[1] <- "Treatment"

tableau_publication

#Save table 
result_dir <- here::here("activity-and-behavior", "result", "behavior")

if (!dir.exists(result_dir)) {
  dir.create(result_dir, recursive = TRUE)
}
output_file_path <- file.path(result_dir, "table_behavior_stats.csv")
write.csv(tableau_publication, file = output_file_path, row.names = FALSE)
```

#------------------------------------------------------
## Linear mixed models (3 x 2 model)
### Single antennal touch
```{r}
aggregate(nb.of.touch ~ antibiotic, df_behavior, sum) 

# Full model
model_nb2_single_touch <- glmmTMB(
  nb.of.touch ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_single_touch, type=3)

#Additive model
model_nb2_add_single_touch <- glmmTMB(
  nb.of.touch ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_add_single_touch, type = 2)

#to test if the interaction if significant
anova(model_nb2_single_touch,model_nb2_add_single_touch)

#pairwise comparison
emm_antibio <- emmeans(model_nb2_add_single_touch, ~ antibiotic)
summary(emm_antibio, type = "response")
pairs(emm_antibio, adjust = "tukey", type = "response") # HSD for the 3 doses

emm_reino <- emmeans(model_nb2_add_single_touch, ~ reinoculation)
summary(emm_reino, type = "response")
pairs(emm_reino, type = "response")
```

### Head to head interaction
```{r}
aggregate(nb.of.head.to.head ~ antibiotic, df_behavior, sum)

# Full model
model_nb2_head <- glmmTMB(
  nb.of.head.to.head ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_head, type=3)

#Additive model
model_nb2_head_add <- glmmTMB(
  nb.of.head.to.head ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_head_add, type=2)

#to test if the interaction if significant
anova(model_nb2_head,model_nb2_head_add)

#pairwise comparison
emm_antibio <- emmeans(model_nb2_head_add, ~ antibiotic)
summary(emm_antibio, type = "response")
pairs(emm_antibio, adjust = "tukey", type = "response") # HSD for the 3 doses

emm_reino <- emmeans(model_nb2_head_add, ~ reinoculation)
summary(emm_reino, type = "response")
pairs(emm_reino, type = "response")
```

### Avoidance

```{r}
aggregate(nb.of.avoidance ~ antibiotic, df_behavior, sum) 

#Full model
model_nb2_avoid <- glmmTMB(
  nb.of.avoidance ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_avoid, type=3)

#Additive model
model_nb2_add_avoid <- glmmTMB(
  nb.of.avoidance ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_add_avoid, type=2)

#to test if the interaction if significant
anova(model_nb2_avoid, model_nb2_add_avoid)

#pairwise comparison
emm_antibio <- emmeans(model_nb2_add_avoid, ~ antibiotic)
summary(emm_antibio, type = "response")
pairs(emm_antibio, adjust = "tukey", type = "response") # HSD for the 3 doses

emm_reino <- emmeans(model_nb2_add_avoid, ~ reinoculation)
summary(emm_reino, type = "response")
pairs(emm_reino, type = "response")
```

### Bumps
```{r}
aggregate(nb.of.bump ~ antibiotic, df_behavior, sum) 

#Full model
model_nb2_bump <- glmmTMB(
  nb.of.bump ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_bump, type=3)

#Additive model
model_nb2_add_bump <- glmmTMB(
  nb.of.bump ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)
Anova(model_nb2_add_bump, type = 2)

#to test if the interaction if significant
anova(model_nb2_bump, model_nb2_add_bump)

#pairwise comparison
emm_antibio <- emmeans(model_nb2_add_bump, ~ antibiotic)
summary(emm_antibio, type = "response")
pairs(emm_antibio, adjust = "tukey", type = "response") # HSD for the 3 doses

emm_reino <- emmeans(model_nb2_add_bump, ~ reinoculation)
summary(emm_reino, type = "response")
pairs(emm_reino, type = "response")
```

## Test residuals 
```{r}
### Function to run DHARMa diagnostics
check_my_model <- function(model, title = "Model Diagnostics") {
  cat("\n--- Diagnostics for:", title, "---\n")
  res <- simulateResiduals(fittedModel = model, plot = FALSE)
  plot(res)
  print(testDispersion(res))      # Over/Under-dispersion
  print(testZeroInflation(res))   # Zero-inflation
  print(testOutliers(res))        # Outlier detection
}

check_my_model(model_nb2_single_touch, "3x2 Single Antennal Touch")
check_my_model(model_nb2_head, "3x2 Head to Head")
check_my_model(model_nb2_avoid, "3x2 Avoidance")
check_my_model(model_nb2_bump, "3x2 Bump")
```

## Visualisation behaviors
```{r}
### Extraction of predictions (EMMEANS) for the 4 behavioral metrics (additive model as they were the one retained)
get_emm_data <- function(model, label_name) {
  emm <- emmeans(model, ~ antibiotic + reinoculation, type = "response")
  df <- as.data.frame(emm)
  df$Behavior <- label_name
  return(df)
}

df_touch <- get_emm_data(model_nb2_single_touch, "Single Antennal Touch")
df_head  <- get_emm_data(model_nb2_head, "Head to Head")
df_avoid <- get_emm_data(model_nb2_avoid, "Avoidance")
df_bump  <- get_emm_data(model_nb2_bump, "Bump")

emm_all <- rbind(df_touch, df_head, df_avoid, df_bump)
```

### Preparation of raw data (long format)
```{r}
df_behavior_long <- df_behavior %>%
  pivot_longer(
    cols = c("nb.of.touch", "nb.of.head.to.head", "nb.of.avoidance", "nb.of.bump"),
    names_to = "variable",
    values_to = "Count"
  ) %>%
  mutate(Behavior = case_when(
    variable == "nb.of.touch" ~ "Single Antennal Touch",
    variable == "nb.of.head.to.head" ~ "Head to Head",
    variable == "nb.of.avoidance" ~ "Avoidance",
    variable == "nb.of.bump" ~ "Bump"
  ))

#In order to have the right order for the plot
emm_all$Behavior <- factor(emm_all$Behavior, 
                           levels = c("Single Antennal Touch", "Bump", "Head to Head", "Avoidance"))
df_behavior_long$Behavior <- factor(df_behavior_long$Behavior, 
                             levels = c("Single Antennal Touch", "Bump","Head to Head", "Avoidance"))
```

### Boxplot
```{r}
p_combined_boxplot <- ggplot(df_behavior_long, aes(x = antibiotic, y = Count, fill = reinoculation)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.1,
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    mapping = aes(ymin = after_stat(y), ymax = after_stat(y)),
    width = 0.3,
    linewidth = 0.8,
    position = position_dodge(0.8)
  ) +
  geom_point(
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  facet_wrap(~ Behavior, scales = "free_y") +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Number of interactions",
    fill = "Re-inoculation"
  )

p_combined_boxplot

ggsave(
  filename = here::here("activity-and-behavior","result", "behavior","behavior_combined_3x2_boxplot.png"), 
  plot = p_combined_boxplot, 
  width = 10, 
  height = 7, 
  dpi = 300
)
```

### Model prediction plot
```{r}
p_model_pred <- ggplot() +
  geom_errorbar(
    data = as.data.frame(emm_all),
    aes(x = antibiotic, ymin = asymp.LCL, ymax = asymp.UCL, group = reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = as.data.frame(emm_all),
    aes(x = antibiotic, ymin = response, ymax = response, group = reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_behavior_long,
    aes(x = antibiotic, y = Count, fill = reinoculation),
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  facet_wrap(~ Behavior, scales = "free_y") +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Number of interactions",
    fill = "Re-inoculation"
  )

p_model_pred

ggsave(
  filename = here::here("activity-and-behavior","result", "behavior","behavior_combined_3x2_model_prediction.png"), 
  plot = p_model_pred, 
  width = 10, 
  height = 7, 
  dpi = 300
)
```

## PCA on 3 x 2 model 
### PCA for Treatment (3X2)
```{r}
set.seed(67) 

data2 <- df_behavior[, !(names(df_behavior) %in% c("antibiotic", "reinoculation", "year", "arena"))]
head(data2)
X <- dimensio::pca(data2, center = TRUE, scale = FALSE, sup_quali = "treatment")
get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

p1 <- viz_individuals(
  x = X,
  extra_quali = data2$treatment,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"),
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)

# Data Preparation for Vegan
stats_matrix <- data2[, sapply(data2, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# Check for Homogeneity of Dispersion (Betadisper). This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, df_behavior$treatment)
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat <- permutest(dispersion_mod, permutations = 10000)
permu_treat

# PERMANOVA (Adonis2) "Are the centers of the groups significantly different?"
ado_treat <- adonis2(dist_matrix ~ antibiotic*reinoculation, strata=df_behavior$year, data = df_behavior, method = "euclidean", permutations = 10000)
ado_treat

ado_ant <- adonis2(dist_matrix ~ antibiotic, strata = df_behavior$year, data = df_behavior, permutations = 10000)
ado_gut <- adonis2(dist_matrix ~ reinoculation, strata = df_behavior$year, data = df_behavior, permutations = 10000)
```

## Save result for LMM and PCA 3 x 2 model 
```{r}
sink(here::here("activity-and-behavior", "result", "behavior", "behavior_3x2_summary.txt"), split = TRUE)

cat("=========================================================================\n")
cat("            3 x 2 FACTORIAL MODEL — BEHAVIORAL GLMM RESULTS\n")
cat("=========================================================================\n\n")
cat("\n=====================================\n")
cat("--- 1. Single Antennal Touch ---\n")
cat("=====================================\n")
cat("\n[1A. Interaction & Model Comparison (LRT)]\n")
print(Anova(model_nb2_single_touch, type = "III"))
cat("\n--> Model Comparison (Full vs Additive):\n")
print(anova(model_nb2_single_touch, model_nb2_add_single_touch))
cat("\n[1B. Main Effects Test (Additive Model Type II)]\n")
print(Anova(model_nb2_add_single_touch, type = "II"))
cat("\n[1C. Pairwise Comparisons: Antibiotic Doses (Tukey HSD)]\n")
print(pairs(emmeans(model_nb2_add_single_touch, ~ antibiotic), adjust = "tukey", type = "response"))
cat("\n[1D. Pairwise Comparisons: Reinoculation (Yes vs No)]\n")
print(pairs(emmeans(model_nb2_add_single_touch, ~ reinoculation), type = "response"))
cat("\n=====================================\n")
cat("--- 2. Head-to-Head Interaction ---\n")
cat("=====================================\n")
cat("\n[2A. Interaction & Model Comparison (LRT)]\n")
print(Anova(model_nb2_head, type = "III"))
cat("\n--> Model Comparison (Full vs Additive):\n")
print(anova(model_nb2_head, model_nb2_head_add))
cat("\n[2B. Main Effects Test (Additive Model Type II)]\n")
print(Anova(model_nb2_head_add, type = "II"))
cat("\n[2C. Pairwise Comparisons: Antibiotic Doses (Tukey HSD)]\n")
print(pairs(emmeans(model_nb2_head_add, ~ antibiotic), adjust = "tukey", type = "response"))
cat("\n[2D. Pairwise Comparisons: Reinoculation (Yes vs No)]\n")
print(pairs(emmeans(model_nb2_head_add, ~ reinoculation), type = "response"))
cat("\n=====================================\n")
cat("--- 3. Avoidance ---\n")
cat("=====================================\n")
cat("\n[3A. Interaction & Model Comparison (LRT)]\n")
print(Anova(model_nb2_avoid, type = "III"))
cat("\n--> Model Comparison (Full vs Additive):\n")
print(anova(model_nb2_avoid, model_nb2_add_avoid))
cat("\n[3B. Main Effects Test (Additive Model Type II)]\n")
print(Anova(model_nb2_add_avoid, type = "II"))
cat("\n[3C. Pairwise Comparisons: Antibiotic Doses (Tukey HSD)]\n")
print(pairs(emmeans(model_nb2_add_avoid, ~ antibiotic), adjust = "tukey", type = "response"))
cat("\n[3D. Pairwise Comparisons: Reinoculation (Yes vs No)]\n")
print(pairs(emmeans(model_nb2_add_avoid, ~ reinoculation), type = "response"))
cat("\n=====================================\n")
cat("--- 4. Bumps ---\n")
cat("=====================================\n")
cat("\n[4A. Interaction & Model Comparison (LRT)]\n")
print(Anova(model_nb2_bump, type = "III"))
cat("\n--> Model Comparison (Full vs Additive):\n")
print(anova(model_nb2_bump, model_nb2_add_bump))
cat("\n[4B. Main Effects Test (Additive Model Type II)]\n")
print(Anova(model_nb2_add_bump, type = "II"))
cat("\n[4C. Pairwise Comparisons: Antibiotic Doses (Tukey HSD)]\n")
print(pairs(emmeans(model_nb2_add_bump, ~ antibiotic), adjust = "tukey", type = "response"))
cat("\n[4D. Pairwise Comparisons: Reinoculation (Yes vs No)]\n")
print(pairs(emmeans(model_nb2_add_bump, ~ reinoculation), type = "response"))
cat("\n=====================================\n")
cat("--- 5. PERMANOVA & PCA 3x2 ---\n")
cat("=====================================\n")
cat("\n[5A. Homogeneity of Multivariate Dispersions]\n")
print(permu_treat)
cat("\n[5B. Global PERMANOVA (Treatment)]\n")
print(ado_treat)
cat("\n[5C. PERMANOVA (Antibiotic Main Effect)]\n")
print(ado_ant)
cat("\n[5D. PERMANOVA (Reinoculation Main Effect)]\n")
print(ado_gut)
sink()
```

#-----------------------------------------------------
## Linear mixed models (2 x 2 model)
```{r}
# We join togheter low and zero into the "zero category"
df_behavior_reduce <- df_behavior

df_behavior_reduce$antibiotic[df_behavior_reduce$antibiotic %in% c("low", "zero")] <- "zero"
df_behavior_reduce$antibiotic <- factor(df_behavior_reduce$antibiotic)
levels(df_behavior_reduce$antibiotic)
df_behavior_reduce$treatment <- paste(df_behavior_reduce$antibiotic, df_behavior_reduce$reinoculation, sep = "_")

#Make sure the reference is zero antibiotic and no recolonisation, year as a factor
df_behavior_reduce$antibiotic <- relevel(df_behavior_reduce$antibiotic, ref = "zero")
df_behavior_reduce$reinoculation <- relevel(df_behavior_reduce$reinoculation, ref = "no")
df_behavior_reduce$year <- factor(df_behavior_reduce$year)
```

### Single antennal touch
```{r}
aggregate(nb.of.touch ~ antibiotic, df_behavior_reduce, sum) 

# Full model
model_nb2_single_touch_red <- glmmTMB(
  nb.of.touch ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_single_touch_red)

#Additive model
model_nb2_add_single_touch_red <- glmmTMB(
  nb.of.touch ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_add_single_touch_red)

#to test if the interaction if significant
anova(model_nb2_add_single_touch_red, model_nb2_single_touch_red)
```

### Head to head interaction
```{r}
aggregate(nb.of.head.to.head ~ antibiotic, df_behavior_reduce, sum)

#Full model
model_nb2_head_red <- glmmTMB(
  nb.of.head.to.head ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_head_red)

#Additive model
model_nb2_add_head_red <- glmmTMB(
  nb.of.head.to.head ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_add_head_red)

#to test if the interaction if significant
anova(model_nb2_head_red, model_nb2_add_head_red)
```

### Avoidance
```{r}
aggregate(nb.of.avoidance ~ antibiotic, df_behavior_reduce, sum) 

# Full model
model_nb2_avoid_red <- glmmTMB(
  nb.of.avoidance ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_avoid_red)

#Additive model 
model_nb2_add_avoid_red <- glmmTMB(
  nb.of.avoidance ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_add_avoid_red)

#to test if the interaction if significant
anova(model_nb2_avoid_red, model_nb2_add_avoid_red)
```

### Bumps
```{r}
aggregate(nb.of.bump ~ antibiotic, df_behavior_reduce, sum) 

# Full model
model_nb2_bump_red <- glmmTMB(
  nb.of.bump ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_bump_red)

#Additive model
model_nb2_add_bump_red <- glmmTMB(
  nb.of.bump ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_add_bump_red)

#to test if the interaction if significant
anova(model_nb2_bump_red, model_nb2_add_bump_red)
```

## PCA on 2 x 2 model 

### PCA for Treatment (2X2)

```{r}
set.seed(67) 

data2 <- df_behavior_reduce[, !(names(df_behavior_reduce) %in% c("antibiotic", "reinoculation", "year", "arena"))]
head(data2)

X <- dimensio::pca(data2, center = TRUE, scale = FALSE, sup_quali = "treatment")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data2$treatment,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), 
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)

# Data Preparation for Vegan
stats_matrix <- data2[, sapply(data2, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# Check for Homogeneity of Dispersion (Betadisper) "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, df_behavior_reduce$treatment)
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat2x2 <- permutest(dispersion_mod, permutations = 10000)
permu_treat2x2
# PERMANOVA (Adonis2) "Are the centers of the groups significantly different?"

ado_treat2x2 <- adonis2(dist_matrix ~ antibiotic*reinoculation, strata=df_behavior_reduce$year, data = df_behavior_reduce, method = "euclidean", permutations = 10000)
ado_treat2x2

ado_ant2x2 <- adonis2(dist_matrix ~ antibiotic, strata=df_behavior_reduce$year, data = df_behavior_reduce, method = "euclidean", permutations = 10000)
ado_gut2x2 <- adonis2(dist_matrix ~ reinoculation, strata=df_behavior_reduce$year, data = df_behavior_reduce, method = "euclidean", permutations = 10000)
```

## Save result for LMM and PCA 2 x 2 model 
```{r}
sink(here::here("activity-and-behavior", "result", "behavior", "behavior_2x2_summary.txt"), split = TRUE)
cat("=========================================================================\n")
cat("            2 x 2 FACTORIAL MODEL — BEHAVIORAL GLMM RESULTS\n")
cat("=========================================================================\n\n")
cat("\n=====================================\n")
cat("--- 1. Single Antennal Touch ---\n")
cat("=====================================\n")
cat("\n[1A. Full Model (Interaction Type III)]\n")
print(Anova(model_nb2_single_touch_red, type = "III"))
cat("\n[1B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_nb2_add_single_touch_red, type = "II"))
cat("\n[1C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_nb2_add_single_touch_red, model_nb2_single_touch_red))
cat("\n=====================================\n")
cat("--- 2. Head-to-Head Interaction ---\n")
cat("=====================================\n")
cat("\n[2A. Full Model (Interaction Type III)]\n")
print(Anova(model_nb2_head_red, type = "III"))
cat("\n[2B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_nb2_add_head_red, type = "II"))
cat("\n[2C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_nb2_add_head_red, model_nb2_head_red))
cat("\n=====================================\n")
cat("--- 3. Avoidance ---\n")
cat("=====================================\n")
cat("\n[3A. Full Model (Interaction Type III)]\n")
print(Anova(model_nb2_avoid_red, type = "III"))
cat("\n[3B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_nb2_add_avoid_red, type = "II"))
cat("\n[3C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_nb2_add_avoid_red, model_nb2_avoid_red))
cat("\n=====================================\n")
cat("--- 4. Bumps ---\n")
cat("=====================================\n")
cat("\n[4A. Full Model (Interaction Type III)]\n")
print(Anova(model_nb2_bump_red, type = "III"))
cat("\n[4B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_nb2_add_bump_red, type = "II"))
cat("\n[4C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_nb2_add_bump_red, model_nb2_bump_red))
cat("\n=====================================\n")
cat("--- 5. PERMANOVA & PCA 2x2 ---\n")
cat("=====================================\n")
cat("\n[5A. Homogeneity of Multivariate Dispersions]\n")
print(permu_treat2x2)
cat("\n[5B. Global PERMANOVA (Treatment)]\n")
print(ado_treat2x2)
cat("\n[5C. PERMANOVA (Antibiotic Main Effect)]\n")
print(ado_ant2x2)
cat("\n[5D. PERMANOVA (Reinoculation Main Effect)]\n")
print(ado_gut2x2)
cat("\n=====================================\n")
cat("END OF ANALYSIS\n")
cat("=====================================\n")
sink()
```

#==============================
# Analysis for interindividual data 
## Basic stats
```{r}
moyenne_Mean_dist <- tapply(df_interind$mean.dist, df_interind$treatment, mean, na.rm = TRUE)
ecarts_Mean_dist <- tapply(df_interind$mean.dist, df_interind$treatment, sd, na.rm = TRUE)

resultats_mean_dist <- data.frame(
  Moyenne_cm = moyenne_Mean_dist,
  Ecart_Type_cm = ecarts_Mean_dist
)

resultats_mean_dist

#save table
dir.create(here::here("activity-and-behavior", "result", "inter_individual_dist"), recursive = TRUE, showWarnings = FALSE)

write.csv(resultats_mean_dist, here::here("activity-and-behavior", "result", "inter_individual_dist", "table_dist_interind.csv"), row.names = TRUE)
```

# LMM model on 3 x 2 
```{r}
hist(df_interind$mean.dist, breaks = 15)
qqnorm(df_interind$mean.dist)

#fit the model
model_gauss <- lmer(mean.dist ~ antibiotic * reinoculation + (1 | year),data = df_interind)

#residuals tests
simulateResiduals(model_gauss) |> plot()
testDispersion(model_gauss)
testResiduals(model_gauss)

a1 <- Anova(model_gauss, type = "III")
a1

#Additive Model
model_gauss_add <- lmer(mean.dist ~ antibiotic + reinoculation + (1 | year), data = df_interind)

a2 <- Anova(model_gauss_add, type = "II")
a2

#to test if the interaction if significant
anova(model_gauss_add, model_gauss)  #Here the interaction model is a better fit

emm <- emmeans(model_gauss, ~ antibiotic * reinoculation)
pairs(emm, by = "antibiotic")
pairs(emm, by = "reinoculation", adjust = "tukey")
```

### Visualisation
```{r}
emm <- emmeans(model_gauss, ~ antibiotic * reinoculation)
emm_df <- as.data.frame(emm)
pairs(emm)

p1 <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = lower.CL, ymax = upper.CL, group = reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = emmean, ymax = emmean, group = reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_interind,
    aes(x = antibiotic, y = mean.dist, fill = reinoculation),
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Mean interindividual distance [cm]",
    fill = "Re-inoculation"
  )

p1

ggsave(
  filename = here::here("activity-and-behavior", "result", "inter_individual_dist", "mean_interind_3x2_plot.png"), 
  plot = p1, 
  width = 12, 
  height = 8, 
  dpi = 300
)
```

# Reduced model 
```{r}
# Joining low and zero
df_interind_reduced <- df_interind
df_interind_reduced$antibiotic[df_interind_reduced$antibiotic %in% c("low", "zero")] <- "zero"

df_interind_reduced<- df_interind_reduced %>%
  mutate(treatment = paste(antibiotic, reinoculation, sep = "_"))

df_interind_reduced <- df_interind_reduced %>%
  mutate(treatment = factor(treatment),              
    treatment = relevel(treatment, ref = "zero_no"),
    antibiotic = factor(antibiotic),
    antibiotic = relevel(antibiotic, ref = "zero"),
    reinoculation = factor(reinoculation),
    reinoculation = relevel(reinoculation, ref = "no"),
    year = factor(year)   
  )

levels(df_interind_reduced$treatment)

hist(df_interind_reduced$mean.dist, breaks = 15)
qqnorm(df_interind_reduced$mean.dist)

#fit the model
model_gauss_red <- lmer(mean.dist ~ antibiotic * reinoculation + (1 | year),data = df_interind_reduced)

#residuals tests
simulateResiduals(model_gauss_red) |> plot()
testDispersion(model_gauss_red)
testResiduals(model_gauss_red)

a3 <- Anova(model_gauss_red, type = "III")
a3

#additive Model
model_gauss_red_add <- lmer(mean.dist ~ antibiotic + reinoculation + (1 | year), data = df_interind_reduced)

a4 <- Anova(model_gauss_red_add, type = "II")
a4

#to test if the interaction if significant
anova(model_gauss_red, model_gauss_red_add) 
```

### Visualisation
```{r}
emm <- emmeans(model_gauss_red, ~ antibiotic * reinoculation)
emm_df <- as.data.frame(emm)
pairs(emm)

p2 <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = lower.CL, ymax = upper.CL, group = reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = emmean, ymax = emmean, group = reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_interind_reduced,
    aes(x = antibiotic, y = mean.dist, fill = reinoculation),
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Mean interindividual distance [cm]",
    fill = "Re-inoculation"
  )

p2

ggsave(
  filename = here::here("activity-and-behavior", "result", "inter_individual_dist", "mean_interind_2x2_plot.png"), 
  plot = p2, 
  width = 10, 
  height = 7, 
  dpi = 300
)
```

# Save the result in a .txt file
```{r}
sink(here::here("activity-and-behavior", "result", "inter_individual_dist", "interind_dist_result.txt"), split = TRUE)
cat("=========================================================================\n")
cat("            MEAN INTERINDIVIDUAL DISTANCE — LMM RESULTS\n")
cat("=========================================================================\n\n")
cat("=====================================\n")
cat("--- 1. FULL MODEL (3 x 2) ---\n")
cat("=====================================\n")
cat("\n[1A. Full Model (Interaction Type III)]\n")
print(Anova(model_gauss, type = "III"))
cat("\n[1B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_gauss_add, type = "II"))
cat("\n[1C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_gauss_add, model_gauss))
cat("\n[1D. the interaction model is a better fit\n")
emm <- emmeans(model_gauss, ~ antibiotic * reinoculation)
pairs(emm, by = "antibiotic")
pairs(emm, by = "reinoculation", adjust = "tukey")
cat("\n=====================================\n")
cat("--- 2. REDUCED MODEL (2 x 2) ---\n")
cat("=====================================\n")
cat("\n[2A. Full Model (Interaction Type III)]\n")
print(Anova(model_gauss_red, type = "III"))
cat("\n[2B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_gauss_red_add, type = "II"))
cat("\n[2C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_gauss_red_add, model_gauss_red))
cat("\n=====================================\n")
cat("END OF ANALYSIS\n")
cat("=====================================\n")
sink()
```

#==============================
#Analysis for activity (avg speed + time spent moving) 
## Basic stats
```{r}
moyenne_speed <- tapply(df_move$average.speed, df_move$treatment, mean, na.rm = TRUE)
ecarts_speed <- tapply(df_move$average.speed, df_move$treatment, sd, na.rm = TRUE)

resultats_speed <- data.frame(
  Moyenne = moyenne_speed,
  Ecart_Type = ecarts_speed
)

moyenne_prop <- tapply(df_move$prop.time.moving, df_move$treatment, mean, na.rm = TRUE)
ecarts_prop <- tapply(df_move$prop.time.moving, df_move$treatment, sd, na.rm = TRUE)

resultats_prop <- data.frame(
  Moyenne = moyenne_prop,
  Ecart_Type = ecarts_prop
)

#save table
dir.create(here::here("activity-and-behavior", "result", "activity"), recursive = TRUE, showWarnings = FALSE)
write.csv(resultats_speed, here::here("activity-and-behavior", "result", "activity", "avg_speed_table.csv"), row.names = TRUE)
write.csv(resultats_prop, here::here("activity-and-behavior", "result", "activity", "prop_of_time_moving_table.csv"), row.names = TRUE)
```

#------------------------------------------------------
# Average speed analysis
## LLM full model

```{r}
hist(df_move$average.speed, breaks = 50)
qqnorm(df_move$average.speed)

#Full model
model_avg.speed <- lmer(average.speed ~ antibiotic * reinoculation + (1 | year:arena), data = df_move)
simulateResiduals(model_avg.speed) |> plot()
testDispersion(model_avg.speed)
testResiduals(model_avg.speed)

a1 <- Anova(model_avg.speed, type = "III")
a1 

#Additive Model
model_avg.speed_add <- lmer(average.speed ~ antibiotic + reinoculation + (1 | year:arena), data = df_move)
a2 <- Anova(model_avg.speed_add, type = "II")
a2

#to test if the interaction if significant
anova(model_avg.speed_add, model_avg.speed)  #Here the interaction model is a better fit

emm <- emmeans(model_avg.speed, ~ antibiotic * reinoculation)
pairs(emm, by = "antibiotic")
pairs(emm, by = "reinoculation", adjust = "tukey")
```

## Visualisation
### Simple box plot 
```{r}
p_boxplot <- ggplot(df_move, aes(x = antibiotic, y = average.speed, fill = reinoculation)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.1,
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    mapping = aes(ymin = after_stat(y), ymax = after_stat(y)),
    width = 0.5,
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Average speed [cm/sec]",
    fill = "Re-inoculation"
  )

p_boxplot

ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "avg_speed_boxplot.pdf"), 
  plot = p_boxplot, 
  width = 9, 
  height = 5
)
```

### The model prediction
```{r}
emm <- emmeans(model_avg.speed, ~ antibiotic * reinoculation)
emm_df <- as.data.frame(emm)

p_model <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = lower.CL, ymax = upper.CL, group = reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = emmean, ymax = emmean, group = reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_move, 
    aes(x = antibiotic, y = average.speed, fill = reinoculation), 
    shape = 21,            
    colour = "black",       
    stroke = 0.3,          
    alpha = 0.9, 
    size = 3.5, 
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Average speed [cm/sec]",
    fill = "Re-inoculation"   
  )

p_model

ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "avg_speed_model_pred.pdf"), 
  plot = p_model, 
  width = 9, 
  height = 5
)
```

## Reduced model
```{r}
#Same stats but joining low and zero
df_move_red <- df_move
df_move_red$antibiotic[df_move_red$antibiotic %in% c("low", "zero")] <- "zero"
df_move_red<- df_move_red %>%
  mutate(treatment = paste(antibiotic, reinoculation, sep = "_"))

df_move_red <- df_move_red %>%
  mutate(
    treatment = factor(treatment),                  
    treatment = relevel(treatment, ref = "zero_no"),
    antibiotic = factor(antibiotic),
    antibiotic = relevel(antibiotic, ref = "zero"),
    reinoculation = factor(reinoculation),
    reinoculation = relevel(reinoculation, ref = "no"),
    year = factor(year),
    arena = factor(arena)
  )
df_move_red$antibiotic <- factor(df_move_red$antibiotic, levels = c("zero", "high"))

hist(df_move_red$average.speed, breaks = 50)
qqnorm(df_move_red$average.speed)

#Full model
model_avg.speed_red <- lmer(average.speed ~ antibiotic * reinoculation + (1 | year:arena), data = df_move_red)
simulateResiduals(model_avg.speed_red) |> plot()
testDispersion(model_avg.speed_red)
testResiduals(model_avg.speed_red)

a3 <- Anova(model_avg.speed_red, type = "III")
a3 

#Additive Model
model_avg.speed_red_add <- lmer(average.speed ~ antibiotic + reinoculation + (1 | year:arena), data = df_move_red)
a4 <- Anova(model_avg.speed_red_add, type = "II")
a4

#to test if the interaction if significant
anova(model_avg.speed_red_add, model_avg.speed_red)  #Here the interaction model is a better fit

emm <- emmeans(model_avg.speed_red, ~ antibiotic * reinoculation)
pairs(emm, by = "antibiotic")
pairs(emm, by = "reinoculation", adjust = "tukey")

```

## Save the result in a .txt file
```{r}
sink(here::here("activity-and-behavior", "result", "activity", "avg_speed_result.txt"), split = TRUE)

cat("=========================================================================\n")
cat("                AVERAGE SPEED ANALYSIS — LMM RESULTS\n")
cat("=========================================================================\n\n")
cat("=====================================\n")
cat("--- 1. FULL MODEL (3 x 2) ---\n")
cat("=====================================\n")
cat("\n[1A. Full Model (Interaction Type III)]\n")
print(Anova(model_avg.speed, type = "III"))
cat("\n[1B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_avg.speed_add, type = "II"))
cat("\n[1C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_avg.speed_add, model_avg.speed))
cat("\n[1D. Post-Hoc Simple Effects Breakdown]\n")
emm_full <- emmeans(model_avg.speed, ~ antibiotic * reinoculation)
cat("\n* Reinoculation effect within each Antibiotic dose:\n")
print(pairs(emm_full, by = "antibiotic"))
cat("\n* Antibiotic effect within each Reinoculation status (Tukey HSD):\n")
print(pairs(emm_full, by = "reinoculation", adjust = "tukey"))
cat("\n=====================================\n")
cat("--- 2. REDUCED MODEL (2 x 2) ---\n")
cat("=====================================\n")
cat("\n[2A. Full Model (Interaction Type III)]\n")
print(Anova(model_avg.speed_red, type = "III"))
cat("\n[2B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_avg.speed_red_add, type = "II"))
cat("\n[2C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_avg.speed_red_add, model_avg.speed_red))
cat("\n[2D. Post-Hoc Simple Effects Breakdown]\n")
emm_red <- emmeans(model_avg.speed_red, ~ antibiotic * reinoculation)
cat("\n* Reinoculation effect within each Antibiotic dose:\n")
print(pairs(emm_red, by = "antibiotic"))
cat("\n* Antibiotic effect within each Reinoculation status (Tukey HSD):\n")
print(pairs(emm_red, by = "reinoculation", adjust = "tukey"))
cat("\n=====================================\n")
cat("END OF ANALYSIS\n")
cat("=====================================\n")
sink()
```

# ------------------------------------------------------
# Proportion of time spent moving 
# LLM Full model
```{r}
hist(df_move$prop.time.moving, breaks = 35)
qqnorm(df_move$prop.time.moving)
range(df_move$prop.time.moving)

# transform y to (0,1) range
n <- length(df_move$prop.time.moving)
y_adj <- (df_move$prop.time.moving * (n - 1) + 0.5) / n
df_move$prop_time_adj <- y_adj
range(df_move$prop_time_adj)

model_beta <- glmmTMB(prop_time_adj ~ antibiotic * reinoculation + (1|year:arena),
                      family = beta_family(link="logit"), #Beta-regression is best suited for this
                      data=df_move) 

#Verifications check
res <- simulateResiduals(model_beta)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta) 

a1 <- Anova(model_beta, type = "III")
a1 


#Additive Model
model_beta_add <- glmmTMB(prop_time_adj ~ antibiotic + reinoculation + (1|year:arena),
                      family = beta_family(link="logit"), #Beta-regression is best suited for this
                      data=df_move) 

a2 <- Anova(model_beta_add, type = "II")
a2

#to test if the interaction if significant
anova(model_beta_add, model_beta)  #Here the interaction model is a better fit

emm <- emmeans(model_beta, ~ antibiotic * reinoculation)
pairs(emm, by = "antibiotic")
pairs(emm, by = "reinoculation", adjust = "tukey")
```

## Visualisation
### Simple boxplot
```{r}
p_boxplot_beta <- ggplot(df_move, aes(x = antibiotic, y = prop_time_adj, fill = reinoculation)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.1,
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    mapping = aes(ymin = after_stat(y), ymax = after_stat(y)),
    width = 0.5,
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Proportion of time moving",
    fill = "Re-inoculation"
  )

p_boxplot_beta

ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "prop_box_plot.pdf"), 
  plot = p_boxplot_beta, 
  width = 9, 
  height = 5
)
```

### The model prediction
```{r}
emm <- emmeans(model_beta, ~ antibiotic * reinoculation, type = "response") 
emm_df <- as.data.frame(emm)

p_pred_model_prop <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = asymp.LCL, ymax = asymp.UCL, group = reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = response, ymax = response, group = reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_move, 
    aes(x = antibiotic, y = prop_time_adj, fill = reinoculation), 
    shape = 21,            
    colour = "black",       
    stroke = 0.3,          
    alpha = 0.9, 
    size = 3.5, 
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Proportion of time moving",
    fill = "Re-inoculation"   
  )

p_pred_model_prop

ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "prop_pred_model.pdf"), 
  plot = p_pred_model_prop, 
  width = 9, 
  height = 5
)
```

# Reduced model
## LLM reduced
```{r}
hist(df_move_red$prop.time.moving, breaks = 35)
qqnorm(df_move_red$prop.time.moving)
range(df_move_red$prop.time.moving)

# transform y to (0,1) range
n <- length(df_move_red$prop.time.moving)
y_adj <- (df_move_red$prop.time.moving * (n - 1) + 0.5) / n
df_move_red$prop_time_adj <- y_adj
range(df_move_red$prop_time_adj)

model_beta_red <- glmmTMB(prop_time_adj ~ antibiotic * reinoculation + (1|year/arena),
                      family = beta_family(link="logit"),
                      data=df_move_red) 

#Verifications check
res <- simulateResiduals(model_beta_red)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta_red) 

a3 <- Anova(model_beta_red, type=3)
a3

#Additive Model
model_beta_red_add <- glmmTMB(prop_time_adj ~ antibiotic + reinoculation + (1|year:arena),
                      family = beta_family(link="logit"), #Beta-regression is best suited for this
                      data=df_move_red) 

a4 <- Anova(model_beta_red_add, type = "II")
a4

#to test if the interaction if significant
anova(model_beta_red_add, model_beta_red)  #Here the interaction model is a better fit

emm <- emmeans(model_beta_red, ~ antibiotic * reinoculation)
pairs(emm, by = "antibiotic")
pairs(emm, by = "reinoculation", adjust = "tukey")

```

#Save the result in a .txt file

```{r}
sink(here::here("activity-and-behavior", "result", "activity", "prop_time_moving_result.txt"), split = TRUE)
cat("=========================================================================\n")
cat("          PROPORTION OF TIME SPENT MOVING — GLMM BETA RESULTS\n")
cat("=========================================================================\n\n")
cat("=====================================\n")
cat("--- 1. FULL MODEL (3 x 2) ---\n")
cat("=====================================\n")
cat("\n[1A. Full Model (Interaction Type III)]\n")
print(Anova(model_beta, type = "III"))
cat("\n[1B. Additive Model (Main Effects Type II)]\n")
print(Anova(model_beta_add, type = "II"))
cat("\n[1C. Likelihood Ratio Test (LRT)]\n")
print(anova(model_beta_add, model_beta))
cat("\n[1D. Post-Hoc Simple Effects Breakdown (Back-transformed to Real Proportions)]\n")
emm_full <- emmeans(model_beta, ~ antibiotic * reinoculation, type = "response")
cat("\n* Reinoculation effect within each Antibiotic dose:\n")
print(pairs(emm_full, by = "antibiotic"))
cat("\n* Antibiotic effect within each Reinoculation status (Tukey HSD):\n")
print(pairs(emm_full, by = "reinoculation", adjust = "tukey"))
cat("\n=====================================\n")
cat("--- 2. REDUCED MODEL (2 x 2) ---\n")
cat("=====================================\n")
cat("\n[2A. Reduced Model (Interaction Type III)]\n")
print(Anova(model_beta_red, type = "III"))
cat("\n[2B. Reduced Additive Model (Main Effects Type II)]\n")
print(Anova(model_beta_red_add, type = "II"))
cat("\n[2C. Likelihood Ratio Test (LRT) for Reduced Model]\n")
print(anova(model_beta_red_add, model_beta_red))
cat("\n[2D. Post-Hoc Simple Effects Breakdown (Back-transformed to Real Proportions)]\n")
emm_red <- emmeans(model_beta_red, ~ antibiotic * reinoculation, type = "response")
cat("\n* Reinoculation effect within each Antibiotic dose:\n")
print(pairs(emm_red, by = "antibiotic"))
cat("\n* Antibiotic effect within each Reinoculation status (Tukey HSD):\n")
print(pairs(emm_red, by = "reinoculation", adjust = "tukey"))
cat("\n=====================================\n")
cat("END OF ANALYSIS\n")
cat("=====================================\n")
sink()
```

#===============================
# PCA ON 7 METRICS
### PCA for Treatment (3 x 2 levels)
```{r}
data2 <- df_pca_7[, !(names(df_pca_7) %in% c("treatment", "year", "arena", "antibiotic", "reinoculation"))]
head(data2)

X <- dimensio::pca(data2, center = TRUE, scale = TRUE)
eig <- get_eigenvalues(X)
eig
screeplot(X, cumulative = TRUE)

viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = df_pca_7$treatment,
  color = c("#4477AA", "#EE6677", "#528833", "#9477AA", "#AE6677", "#E28833"), 
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)


set.seed(67) 
# Data Preparation for Vegan
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper) "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix,df_pca_7$treatment)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat <- permutest(dispersion_mod, permutations = 10000)
permu_treat

# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?" Test 1: Simple effect of Treatment
ado_treat <- adonis2(dist_matrix ~ antibiotic*reinoculation, strata = df_pca_7$year, data = df_pca_7, method = "euclidean", permutations = 10000)
ado_treat

ado_ant <- adonis2(dist_matrix ~ antibiotic, strata=df_pca_7$year, data = df_pca_7, method = "euclidean", permutations = 10000)
ado_ant

ado_reino <- adonis2(dist_matrix ~ reinoculation, strata=df_pca_7$year, data = df_pca_7, method = "euclidean", permutations = 10000)
ado_reino
```

### PCA on the PC1
```{r}
#### extract the PC1
data3 <- df_pca_7[, !(names(df_pca_7) %in% c("year", "arena", "antibiotic", "reinoculation"))]
head(data3)

X <- dimensio::pca(data3, center = TRUE, scale = TRUE, sup_quali = "treatment")
get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

pca_scores <- get_coordinates(X)

combined_interaction_data <- pca_scores$F1

df_pca_7$combined <- combined_interaction_data
```

#### LMM on PC1
```{r}
hist(df_pca_7$combined, breaks = 10)
qqnorm(df_pca_7$combined)

model_lmm <- lmer(combined ~ antibiotic * reinoculation + (1 | year), data = df_pca_7)

res <- simulateResiduals(model_lmm)
plot(res)
testDispersion(res)
testUniformity(res) # This is the standard test for residual distribution

a1 <- Anova(model_lmm, type=3)
a1

emm <- emmeans(model_lmm, ~ antibiotic * reinoculation)
pairs(emm, by = "antibiotic")
pairs(emm, by = "reinoculation", adjust = "tukey")
```

### Visualisation
```{r}
emm <- emmeans(model_lmm, ~ antibiotic * reinoculation)
all_pairwise <- pairs(emm)
emm_df <- as.data.frame(emm)

p1 <- ggplot() +
  # 1. Résultats du modèle LMM : Moustaches IC95 (En noir, fines)
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = lower.CL, ymax = upper.CL, group = reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  # 2. Résultats du modèle LMM : Ligne horizontale pour la Moyenne prédite
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = emmean, ymax = emmean, group = reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  # 3. Données brutes : Points individuels avec contour noir (style habituel)
  geom_point(
    data = df_pca_7, 
    aes(x = antibiotic, y = combined, fill = reinoculation), 
    shape = 21,            
    colour = "black",       
    stroke = 0.3,          
    alpha = 0.9, 
    size = 3.5, 
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  # 4. Palette de couleurs et étiquettes épurées
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  # 5. Thème classique épuré (Cadre noir complet, pas de lignes d'axes)
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "PC1 value",
    fill = "Re-inoculation"   
  )

p1

dir.create(
  path = here::here("activity-and-behavior", "result", "PCA_on_7metrics"), 
  showWarnings = FALSE, 
  recursive = TRUE
)

ggsave(
  filename = here::here("activity-and-behavior", "result", "PCA_on_7metrics", "PC1_Plot.pdf"), 
  plot = p1, 
  width = 9, 
  height = 5
)

```

# Save the result in a .txt file
```{r}
sink(
  file = here::here("activity-and-behavior", "result", "PCA_on_7metrics", "pca_result.txt"), 
  split = TRUE
)

cat("=========================================================================\n")
cat("          MULTIVARIATE PCA (7 METRICS) & LMM ON PC1 RESULTS\n")
cat("=========================================================================\n\n")
cat("=====================================\n")
cat("--- 1. MULTIVARIATE DIMENSIONALITY ---\n")
cat("=====================================\n")
cat("\n[1A. Eigenvalues & Variance Explained]\n")
print(eig)
cat("\n=====================================\n")
cat("--- 2. MULTIVARIATE DISPERSION & PERMANOVA ---\n")
cat("=====================================\n")
cat("\n[2A. Homogeneity of Multivariate Dispersions (Betadisper)]\n")
print(permu_treat)
cat("\n[2B. Global PERMANOVA (Interaction antibiotic * reinoculation)]\n")
print(ado_treat)
cat("\n[2C. PERMANOVA (Antibiotic Main Effect)]\n")
print(ado_ant)
cat("\n[2D. PERMANOVA (Reinoculation Main Effect)]\n")
print(ado_reino)
cat("\n=====================================\n")
cat("--- 3. LINEAR MIXED MODEL (LMM) ON PC1 ---\n")
cat("=====================================\n")
cat("\n[3A. Type III Wald Chisquare Tests (Anova)]\n")
print(Anova(model_lmm, type = 3))
cat("\n[3B. Post-Hoc Breakdown: Reinoculation effect within each Antibiotic dose]\n")
print(pairs(emm, by = "antibiotic"))
cat("\n[3C. Post-Hoc Breakdown: Antibiotic effect within each Reinoculation status]\n")
print(pairs(emm, by = "reinoculation", adjust = "tukey"))
cat("\n[3D. All Pairwise Comparisons]\n")
print(all_pairwise)
sink()
```
