# Behavior and activity statistics

## Load library

```{r}
library(here)         # Relative file path tracking

library(tidyverse)    
library(readr)        
library(dplyr)        
library(tidyr)        

library(lme4)         
library(lmerTest)     
library(glmmTMB)      

library(DHARMa)      
library(car)         
library(emmeans)     

library(vegan)        
library(dimensio)    

library(ggplot2)      
library(ggbeeswarm)
library(patchwork)
library(scales)

set.seed(67)
options(contrasts = c("contr.sum", "contr.poly"))
```

# Data pre-processing 
### Behavioral data
```{r}
categories_all <- list(
  control  = c(5, 8, 13, 40, 41, 42, 43),
  low_ant  = c(2, 14, 15, 16, 24, 25, 26, 27),
  high_ant = c(3, 7, 10, 18, 28, 29, 30, 31),
  low_gut  = c(1, 6, 12, 20, 21, 22, 23, 34),
  high_gut = c(4, 11, 17, 19, 32, 33),
  gut_ctrl = c(35, 36, 37, 38, 39) # 2025 dataset only
)

category_conditions <- tibble(
  treatment    = c('control', 'low_ant', 'high_ant', 'low_gut', 'high_gut', 'gut_ctrl'),
  antibiotic  = c('zero',    'low',     'high',     'low',     'high',     'zero'),
  reinoculation = c('no',      'no',      'no',       'yes',     'yes',      'yes')
)

data_list <- list()
counter <- 1

for (cat_name in names(categories_all)) {
  cat_files <- categories_all[[cat_name]]
  
  for (x in cat_files) {
    # Determine the experimental year based on the dish index threshold
    year <- if_else(x <= 15, 2024, 2025)
    file_name <- paste0(year, "_petridish", x, ".csv")
    
    full_path <- here::here("activity-and-behavior", "data", "behavior_csv_raw", file_name) 
    
    if (!file.exists(full_path)) {
      warning(paste("File not found:", file_name))
      next
    }
    
    df <- read_csv(full_path, show_col_types = FALSE)
        df_filtered <- df %>% filter(time < 1800) #we get exactly 30min
    
    bump_count      <- sum(df_filtered$class == 1, na.rm = TRUE)
    touch_count     <- sum(df_filtered$class == 2, na.rm = TRUE)
    avoidance_count <- sum(df_filtered$class == 5, na.rm = TRUE)
    head_count      <- sum(df_filtered$class == 6, na.rm = TRUE)
    
    data_list[[counter]] <- tibble(
      `petridish number` = x,
      treatment = cat_name,
      year = year,
      `nb of bump` = bump_count,
      `nb of avoidance` = avoidance_count,
      `nb of head-to-head` = head_count,
      `nb of touch` = touch_count
    )
    
    counter <- counter + 1
  }
}

if (length(data_list) > 0) {
  df_behavior <- bind_rows(data_list) %>% 
    left_join(category_conditions, by = "treatment") %>% 
    dplyr::select(
      `petridish number`, treatment, antibiotic, reinoculation, year,
      `nb of bump`, `nb of avoidance`, `nb of head-to-head`, `nb of touch`
    )
}

colnames(df_behavior)<- c("arena", "treatment", "antibiotic", "reinoculation",
                   "year", "nb.of.bump", "nb.of.avoidance", "nb.of.head.to.head",
                   "nb.of.touch")


df_behavior <- df_behavior %>%
  mutate(
    treatment = dplyr::recode(
      treatment,
      control  = "zero_no",
      low_ant  = "low_no",
      high_ant = "high_no",
      low_gut  = "low_yes",
      high_gut = "high_yes",
      gut_ctrl = "zero_yes"
    )
  )

df_behavior$antibiotic <- factor(df_behavior$antibiotic, levels = c("zero", "low", "high"))
df_behavior$reinoculation <- factor(df_behavior$reinoculation)
df_behavior$reinoculation <- relevel(df_behavior$reinoculation, ref = "no")
df_behavior$year <- factor(df_behavior$year)

df_behavior
```

### Inter-individual distance data

```{r}
data_dir <- here::here("activity-and-behavior", "data", "animal_ta")
file_path <- file.path(data_dir, "inderindividual_dist.csv")
           
df_interind <- read_delim(file_path,
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

colnames(df_interind)<- c("treatment", "antibiotic", "reinoculation",
                   "year", "arena", "mean.dist")

df_interind <- df_interind %>%
  mutate(
    treatment = factor(treatment),      
    treatment = relevel(treatment, ref = "zero_no"),
    antibiotic = factor(antibiotic, levels = c("zero", "low", "high")),
    reinoculation = factor(reinoculation),
    reinoculation = relevel(reinoculation, ref = "no"),
    year = factor(year)   
  )
df_interind
```
### Activity data

```{r}
data_dir <- here::here("activity-and-behavior", "data", "animal_ta")
file_path <- file.path(data_dir, "moving.csv")

df_move <- read_delim(file = file_path,
                       delim = ";",
                       locale = locale(decimal_mark = ","),
                       trim_ws = TRUE,
                       col_names = TRUE,
                       progress = FALSE,
                       guess_max = 2000)

colnames(df_move)<- c("treatment", "antibiotic", "reinoculation",
                   "year", "arena", "prop.time.moving", "average.speed")

df_move <- df_move %>%
  mutate(
    treatment = factor(treatment), 
    treatment = relevel(treatment, ref = "zero_no"),
    antibiotic = factor(antibiotic, levels = c("zero", "low", "high")),
    reinoculation = factor(reinoculation),
    reinoculation = relevel(reinoculation, ref = "no"),
    year = factor(year), 
    arena = factor(arena)
  )
df_move
```


### PCA on the 7 metrics 

```{r}

df_move_aggregated <- df_move %>%
  group_by(year, arena, treatment, antibiotic, reinoculation) %>%
  summarise(moy_prop_time_moving = mean(prop.time.moving, na.rm = TRUE),
    moy_average_speed    = mean(average.speed, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    arena = as.numeric(as.character(arena))
  )

df_behavior <- df_behavior %>%
  mutate(arena = as.numeric(arena))

df_interind <- df_interind %>%
  mutate(arena = as.numeric(as.character(arena)))


df_pca_7 <- df_move_aggregated %>%
  left_join(df_behavior, by = c("year", "arena", "antibiotic", "reinoculation", "treatment")) %>%
  left_join(df_interind, by = c("year", "arena", "antibiotic", "reinoculation", "treatment")) %>%
  mutate(
    year        = factor(year),
    arena       = factor(arena),
    treatment   = factor(treatment),
    treatment   = relevel(treatment, ref = "zero_no"),
    antibiotic  = factor(antibiotic, levels = c("zero", "low", "high")),
    reinoculation = factor(reinoculation, levels = c("no", "yes"))
  )

df_pca_7

```


# ==============================

Now that the data is correctly pre-processed, we can start the EDA and the analysis 


## Behavioral analysis

### Basic stats
```{r}
moyenne_bump <- tapply(df_behavior$nb.of.bump, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_bump <- tapply(df_behavior$nb.of.bump, df_behavior$treatment, sd, na.rm = TRUE)

resultats_bump <- data.frame(
  Moyenne = moyenne_bump,
  Ecart_Type = ecarts_bump
)

moyenne_avoidance <- tapply(df_behavior$nb.of.avoidance, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_avoidance <- tapply(df_behavior$nb.of.avoidance, df_behavior$treatment, sd, na.rm = TRUE)

resultats_avoidance <- data.frame(
  Moyenne = moyenne_avoidance,
  Ecart_Type = ecarts_avoidance
)

moyenne_head_to_head <- tapply(df_behavior$nb.of.head.to.head, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_head_to_head <- tapply(df_behavior$nb.of.head.to.head, df_behavior$treatment, sd, na.rm = TRUE)

resultats_head_to_head <- data.frame(
  Moyenne = moyenne_head_to_head,
  Ecart_Type = ecarts_head_to_head
)

moyenne_touch <- tapply(df_behavior$nb.of.touch, df_behavior$treatment, mean, na.rm = TRUE)
ecarts_touch <- tapply(df_behavior$nb.of.touch, df_behavior$treatment, sd, na.rm = TRUE)

resultats_touch <- data.frame(
  Moyenne = moyenne_touch,
  Ecart_Type = ecarts_touch
)

formater_stats <- function(moyenne, et) {
  paste0(round(moyenne, 1), " \u00B1 ", round(et, 1)) # \u00B1 is the symbol +/-
}

tableau_publication <- data.frame(
  Category = rownames(resultats_bump), 
  stringsAsFactors = FALSE
)

tableau_publication$`Random bump` <- formater_stats(
  moyenne_bump, 
  ecarts_bump
)

tableau_publication$`Short antennal touch` <- formater_stats(
  moyenne_touch, 
  ecarts_touch
)

tableau_publication$`Head-to-head interaction` <- formater_stats(
  moyenne_head_to_head, 
  ecarts_head_to_head
)

tableau_publication$Avoidance <- formater_stats(
  moyenne_avoidance, 
  ecarts_avoidance
)

colnames(tableau_publication)[1] <- "Treatment"

tableau_publication

#Save table 
result_dir <- here::here("activity-and-behavior", "result", "behavior")

if (!dir.exists(result_dir)) {
  dir.create(result_dir, recursive = TRUE)
}
output_file_path <- file.path(result_dir, "table_behavior_stats.csv")
write.csv(tableau_publication, file = output_file_path, row.names = FALSE)
```
#------------------------------------------------------
## Linear mixed models (3 x 2 model)

### Single antennal touch

```{r}
aggregate(nb.of.touch ~ antibiotic, df_behavior, sum) 


# Negative Binomial GLMM 
model_nb2_single_touch <- glmmTMB(
  nb.of.touch ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_single_touch)
Anova(model_nb2_single_touch, type=3)


#Additive model
model_nb2_add_single_touch <- glmmTMB(
  nb.of.touch ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_add_single_touch)
Anova(model_nb2_add_single_touch, type = 2)

#to test if the interaction if significant
anova(model_nb2_single_touch,model_nb2_add_single_touch)
```

### Head to head interaction

```{r}
aggregate(nb.of.head.to.head ~ antibiotic, df_behavior, sum)

#Negative Binomial GLMM (NB2)
model_nb2_head <- glmmTMB(
  nb.of.head.to.head ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_head)
Anova(model_nb2_head, type=3)


#Additive model
model_nb2_head_add <- glmmTMB(
  nb.of.head.to.head ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_head_add)
Anova(model_nb2_head_add, type=2)

#to test if the interaction if significant
anova(model_nb2_head,model_nb2_head_add)
```

### Avoidance

```{r}
aggregate(nb.of.avoidance ~ antibiotic, df_behavior, sum) 

# Negative Binomial GLMM (NB2)
model_nb2_avoid <- glmmTMB(
  nb.of.avoidance ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_avoid)
Anova(model_nb2_avoid, type=3)


#Additive model
model_nb2_add_avoid <- glmmTMB(
  nb.of.avoidance ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_add_avoid)
Anova(model_nb2_add_avoid, type=2)

#to test if the interaction if significant
anova(model_nb2_avoid, model_nb2_add_avoid)
```

### Bumps

```{r}
aggregate(nb.of.bump ~ antibiotic, df_behavior, sum) 


# Negative Binomial GLMM
model_nb2_bump <- glmmTMB(
  nb.of.bump ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_bump)
Anova(model_nb2_bump, type=3)


#Additive model
model_nb2_add_bump <- glmmTMB(
  nb.of.bump ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior
)

summary(model_nb2_add_bump)
Anova(model_nb2_add_bump, type = 2)

#to test if the interaction if significant
anova(model_nb2_bump, model_nb2_add_bump)
```

## Test residuals 

```{r}
### Function to run DHARMa diagnostics

check_my_model <- function(model, title = "Model Diagnostics") {
  cat("\n--- Diagnostics for:", title, "---\n")
  # 1. Simulate residuals
  res <- simulateResiduals(fittedModel = model, plot = FALSE)
  # 2. Plot QQ-plot and Residuals vs. Predicted
  plot(res)
  # 3. Specific tests
  print(testDispersion(res))      # Over/Under-dispersion
  print(testZeroInflation(res))   # Zero-inflation
  print(testOutliers(res))        # Outlier detection
}


# Diagnostics for Single Antennal Touch
check_my_model(model_nb2_single_touch, "3x2 Single Antennal Touch")

# Diagnostics for Head to Head
check_my_model(model_nb2_head, "3x2 Head to Head")

# Diagnostics for Avoidance
check_my_model(model_nb2_avoid, "3x2 Avoidance")

# Diagnostics for Bump
check_my_model(model_nb2_bump, "3x2 Bump")
```


## Visualisation behaviors


```{r}
### Extraction of predictions (EMMEANS) for the 4 behavioral metrics
get_emm_data <- function(model, label_name) {
  emm <- emmeans(model, ~ antibiotic + reinoculation, type = "response")
  df <- as.data.frame(emm)
  df$Behavior <- label_name
  return(df)
}

df_touch <- get_emm_data(model_nb2_single_touch, "Single Antennal Touch")
df_head  <- get_emm_data(model_nb2_head, "Head to Head")
df_avoid <- get_emm_data(model_nb2_avoid, "Avoidance")
df_bump  <- get_emm_data(model_nb2_bump, "Bump")

emm_all <- rbind(df_touch, df_head, df_avoid, df_bump)
```

### Preparation of raw data (long format)

```{r}
df_behavior_long <- df_behavior %>%
  pivot_longer(
    cols = c("nb.of.touch", "nb.of.head.to.head", "nb.of.avoidance", "nb.of.bump"),
    names_to = "variable",
    values_to = "Count"
  ) %>%
  mutate(Behavior = case_when(
    variable == "nb.of.touch" ~ "Single Antennal Touch",
    variable == "nb.of.head.to.head" ~ "Head to Head",
    variable == "nb.of.avoidance" ~ "Avoidance",
    variable == "nb.of.bump" ~ "Bump"
  ))

# Have the right order for the plot
emm_all$Behavior <- factor(emm_all$Behavior, 
                           levels = c("Single Antennal Touch", "Bump", "Head to Head", "Avoidance"))
df_behavior_long$Behavior <- factor(df_behavior_long$Behavior, 
                             levels = c("Single Antennal Touch", "Bump","Head to Head", "Avoidance"))


```

### Boxplot
```{r}
p_combined_boxplot <- ggplot(df_behavior_long, aes(x = antibiotic, y = Count, fill = reinoculation)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.1,
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    mapping = aes(ymin = after_stat(y), ymax = after_stat(y)),
    width = 0.3,
    linewidth = 0.8,
    position = position_dodge(0.8)
  ) +
  geom_point(
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  facet_wrap(~ Behavior, scales = "free_y") +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Number of interactions",
    fill = "Re-inoculation"
  )

print(p_combined_boxplot)


ggsave(
  filename = here::here("activity-and-behavior","result", "behavior","behavior_combined_3x2_boxplot.png"), 
  plot = p_combined_boxplot, 
  width = 10, 
  height = 7, 
  dpi = 300
)
```
### Model prediction plot
```{r}
p_model_pred <- ggplot() +
  geom_errorbar(
    data = as.data.frame(emm_all),
    aes(x = antibiotic, ymin = asymp.LCL, ymax = asymp.UCL, group = reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = as.data.frame(emm_all),
    aes(x = antibiotic, ymin = response, ymax = response, group = reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_behavior_long,
    aes(x = antibiotic, y = Count, fill = reinoculation),
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  facet_wrap(~ Behavior, scales = "free_y") +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Number of interactions",
    fill = "Re-inoculation"
  )

print(p_model_pred)


ggsave(
  filename = here::here("activity-and-behavior","result", "behavior","behavior_combined_3x2_model_prediction.png"), 
  plot = p_model_pred, 
  width = 10, 
  height = 7, 
  dpi = 300
)

```


## PCA on 3 x 2 model 

### PCA for Treatment (3X2)

```{r}
data2 <- df_behavior[, !(names(df_behavior) %in% c("antibiotic", "reinoculation", "year", "arena"))]
head(data2)
X <- dimensio::pca(data2, center = TRUE, scale = FALSE, sup_quali = "treatment")
get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

p1 <- viz_individuals(
  x = X,
  extra_quali = data2$treatment,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)

set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data2[, sapply(data2, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, df_behavior$treatment)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat <- permutest(dispersion_mod, permutations = 10000)
permu_treat
# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(67) # Ensures results are reproducible
ado_treat <- adonis2(dist_matrix ~ antibiotic*reinoculation, strata=df_behavior$year, data = df_behavior, method = "euclidean", permutations = 10000)
ado_treat


ado_ant <- adonis2(dist_matrix ~ antibiotic, strata = df_behavior$year, data = df_behavior, permutations = 10000)
ado_gut <- adonis2(dist_matrix ~ reinoculation, strata = df_behavior$year, data = df_behavior, permutations = 10000)
```


## Save result for LMM and PCA 3 x 2 model 

```{r}

sink(here::here("activity-and-behavior", "result", "behavior", "behavior_3x2_summary.txt"), split = TRUE)

cat("\n\n=====================================\n")
cat("3 x 2 MODEL — GLMM RESULTS\n")
cat("=====================================\n\n")

cat("\n--- Single Antennal Touch ---\n")
summary(model_nb2_single_touch)
Anova(model_nb2_single_touch, type = 3)
summary(model_nb2_add_single_touch)
Anova(model_nb2_add_single_touch, type = 2)

anova(model_nb2_single_touch, model_nb2_add_single_touch)

cat("\n--- Head-to-Head Interaction ---\n")
summary(model_nb2_head)
Anova(model_nb2_head, type = 3)
summary(model_nb2_head_add)
Anova(model_nb2_head_add, type = 2)

anova(model_nb2_head, model_nb2_head_add)

cat("\n--- Avoidance ---\n")
summary(model_nb2_avoid)
Anova(model_nb2_avoid, type = 3)
summary(model_nb2_add_avoid)
Anova(model_nb2_add_avoid, type = 2)

anova(model_nb2_avoid, model_nb2_add_avoid)

cat("\n--- Bump ---\n")
summary(model_nb2_bump)
Anova(model_nb2_bump, type = 3)
summary(model_nb2_add_bump)
Anova(model_nb2_add_bump, type = 2)

anova(model_nb2_bump, model_nb2_add_bump)

cat("\n--- PCA 3x2 ---\n")
permu_treat 
ado_treat

ado_ant
ado_gut

sink()
```
#-----------------------------------------------------
## Linear mixed models (2 x 2 model)
```{r}
# We join togheter low and zero into the "zero category"
df_behavior_reduce <- df_behavior

df_behavior_reduce$antibiotic[df_behavior_reduce$antibiotic %in% c("low", "zero")] <- "zero"
df_behavior_reduce$antibiotic <- factor(df_behavior_reduce$antibiotic)
levels(df_behavior_reduce$antibiotic)

df_behavior_reduce$treatment <- paste(df_behavior_reduce$antibiotic, df_behavior_reduce$reinoculation, sep = "_")

#Make sure the reference is zero antibiotic and no recolonisation, year as a factor
df_behavior_reduce$antibiotic <- relevel(df_behavior_reduce$antibiotic, ref = "zero")
df_behavior_reduce$reinoculation <- relevel(df_behavior_reduce$reinoculation, ref = "no")
df_behavior_reduce$year <- factor(df_behavior_reduce$year)



```




### Single antennal touch

```{r}
aggregate(nb.of.touch ~ antibiotic, df_behavior_reduce, sum) 


# Negative Binomial GLMM 
model_nb2_single_touch_red <- glmmTMB(
  nb.of.touch ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)

summary(model_nb2_single_touch_red)


#Additive model
model_nb2_add_single_touch_red <- glmmTMB(
  nb.of.touch ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)

summary(model_nb2_add_single_touch_red)

anova(model_nb2_add_single_touch_red, model_nb2_single_touch_red)
```

### Head to head interaction

```{r}
aggregate(nb.of.head.to.head ~ antibiotic, df_behavior, sum)

#Negative Binomial GLMM (NB2)
model_nb2_head_red <- glmmTMB(
  nb.of.head.to.head ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)

summary(model_nb2_head_red)


model_nb2_add_head_red <- glmmTMB(
  nb.of.head.to.head ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)
summary(model_nb2_add_head_red)


anova(model_nb2_head_red, model_nb2_add_head_red)
```

### Avoidance

```{r}
aggregate(nb.of.avoidance ~ antibiotic, df_behavior_reduce, sum) 

# Negative Binomial GLMM (NB2)
model_nb2_avoid_red <- glmmTMB(
  nb.of.avoidance ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)

summary(model_nb2_avoid_red)


#Additive model 
model_nb2_add_avoid_red <- glmmTMB(
  nb.of.avoidance ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)

summary(model_nb2_add_avoid_red)

anova(model_nb2_avoid_red, model_nb2_add_avoid_red)
```

### Bumps

```{r}
aggregate(nb.of.bump ~ antibiotic, df_behavior_reduce, sum) 

# Negative Binomial GLMM
model_nb2_bump_red <- glmmTMB(
  nb.of.bump ~ antibiotic * reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)

summary(model_nb2_bump_red)

#Additive model
model_nb2_add_bump_red <- glmmTMB(
  nb.of.bump ~ antibiotic + reinoculation + (1 | year),
  family = nbinom2,
  data = df_behavior_reduce
)

summary(model_nb2_add_bump_red)

anova(model_nb2_bump_red, model_nb2_add_bump_red)
```

## PCA on 2 x 2 model 

### PCA for Treatment (2X2)

```{r}
data2 <- df_behavior_reduce[, !(names(df_behavior_reduce) %in% c("antibiotic", "reinoculation", "year", "arena"))]
head(data2)

X <- dimensio::pca(data2, center = TRUE, scale = FALSE, sup_quali = "treatment")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data2$treatment,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)



set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data2[, sapply(data2, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, df_behavior_reduce$treatment)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat2x2 <- permutest(dispersion_mod, permutations = 999)
permu_treat2x2
# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_treat2x2 <- adonis2(dist_matrix ~ antibiotic*reinoculation, strata=df_behavior_reduce$year, data = df_behavior_reduce, method = "euclidean", permutations = 10000)
ado_treat2x2

ado_ant2x2 <- adonis2(dist_matrix ~ antibiotic, strata=df_behavior_reduce$year, data = df_behavior_reduce, method = "euclidean", permutations = 10000)
ado_ant2x2

ado_gut2x2 <- adonis2(dist_matrix ~ reinoculation, strata=df_behavior_reduce$year, data = df_behavior_reduce, method = "euclidean", permutations = 10000)
ado_gut2x2
```

## Save result for LMM and PCA 2 x 2 model 
```{r}
sink(here::here("activity-and-behavior", "result", "behavior", "behavior_2x2_summary.txt"), split = TRUE)

cat("\n\n=====================================\n")
cat("2 x 2 MODEL — GLMM RESULTS\n")
cat("=====================================\n\n")

cat("\n--- Single Antennal Touch ---\n")
summary(model_nb2_single_touch_red)
Anova(model_nb2_single_touch_red, type = 3)
summary(model_nb2_add_single_touch_red)
Anova(model_nb2_add_single_touch_red, type = 2)
anova(model_nb2_add_single_touch_red, model_nb2_single_touch_red)

cat("\n--- Head-to-Head Interaction ---\n")
summary(model_nb2_head_red)
Anova(model_nb2_head_red, type = 3)
summary(model_nb2_add_head_red)
Anova(model_nb2_add_head_red, type = 2)
anova(model_nb2_add_head_red, model_nb2_head_red)

cat("\n--- Avoidance ---\n")
summary(model_nb2_avoid_red)
Anova(model_nb2_avoid_red, type = 3)
summary(model_nb2_add_avoid_red)
Anova(model_nb2_add_avoid_red, type = 2)
anova(model_nb2_add_avoid_red, model_nb2_avoid_red)

cat("\n--- Bump ---\n")
summary(model_nb2_bump_red)
Anova(model_nb2_bump_red, type = 3)
summary(model_nb2_add_bump_red)
Anova(model_nb2_add_bump_red, type = 2)
anova(model_nb2_add_bump_red, model_nb2_bump_red)

cat("\n--- PCA 2x2 ---\n")
permu_treat2x2
ado_treat2x2
ado_ant2x2
ado_gut2x2

cat("\n=====================================\n")
cat("END OF ANALYSIS\n")
cat("=====================================\n")

sink()
```

#==============================
# Analysis for interindividual data 

## Basic stats
```{r}
moyenne_Mean_dist <- tapply(df_interind$Mean_dist, df_interind$Treatment, mean, na.rm = TRUE)
ecarts_Mean_dist <- tapply(df_interind$Mean_dist, df_interind$Treatment, sd, na.rm = TRUE)

resultats_mean_dist <- data.frame(
  Moyenne_cm = moyenne_Mean_dist,
  Ecart_Type_cm = ecarts_Mean_dist
)

resultats_mean_dist

#save table
dir.create(here::here("activity-and-behavior", "result", "inter_individual_dist"), recursive = TRUE, showWarnings = FALSE)

write.csv(resultats_mean_dist, here::here("activity-and-behavior", "result", "inter_individual_dist", "table_dist_interind.csv"), row.names = TRUE)

```



# LMM model on 3 x 2 
```{r}
hist(df_interind$Mean_dist, breaks = 15)
qqnorm(df_interind$Mean_dist)

#fit the model
model_gauss <- lmer(Mean_dist ~ Antibiotic * Reinoculation + (1 | Year),data = df_interind)

#residuals tests
simulateResiduals(model_gauss) |> plot()
testDispersion(model_gauss)
testResiduals(model_gauss)

a1 <- Anova(model_gauss, type = "III") 
a1

s1 <- summary(model_gauss)
s1

emm <- emmeans(model_gauss, ~ Antibiotic * Reinoculation)
emm_df <- as.data.frame(emm)
pairs(emm)
```

### Visualisation
```{r}
emm <- emmeans(model_gauss, ~ Antibiotic * Reinoculation)
emm_df <- as.data.frame(emm)
pairs(emm)

p1 <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = lower.CL, ymax = upper.CL, group = Reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = emmean, ymax = emmean, group = Reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_interind,
    aes(x = Antibiotic, y = Mean_dist, fill = Reinoculation),
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Mean interindividual distance [cm]",
    fill = "Re-inoculation"
  )

print(p1)

#save the plot
ggsave(
  filename = here::here("activity-and-behavior", "result", "inter_individual_dist", "mean_interind_3x2_plot.png"), 
  plot = p1, 
  width = 12, 
  height = 8, 
  dpi = 300
)
```


### additive model (Just to have it - The interaction is significant)
```{r}
model_gauss_add <- lmer(Mean_dist ~ Antibiotic + Reinoculation + (1 | Year),data = df_interind)
simulateResiduals(model_gauss_add) |> plot()
testDispersion(model_gauss_add) 
testResiduals(model_gauss_add) 

summary(model_gauss_add)
Anova(model_gauss, type = "III") 


anova(model_gauss, model_gauss_add)
```


# LMM model on 2x2 (reduced) model 
```{r}
# joining LOW and ZERO
df_interind$Antibiotic[df_interind$Antibiotic %in% c("low", "zero")] <- "zero"

df_interind<- df_interind %>%
  mutate(Treatment = paste(Antibiotic, Reinoculation, sep = "_"))

df_interind <- df_interind %>%
  mutate(
    Treatment = factor(Treatment),              
    Treatment = relevel(Treatment, ref = "zero_no"),
    Antibiotic = factor(Antibiotic),
    Antibiotic = relevel(Antibiotic, ref = "zero"),
    Reinoculation = factor(Reinoculation),
    Reinoculation = relevel(Reinoculation, ref = "no"),
    Year = factor(Year)   
  )


levels(df_interind$Treatment)


hist(df_interind$Mean_dist, breaks = 15)
qqnorm(df_interind$Mean_dist)

model_gauss_low <- lmer(Mean_dist ~ Antibiotic * Reinoculation + (1 | Year),data = df_interind)
simulateResiduals(model_gauss_low) |> plot()
testDispersion(model_gauss_low)
testResiduals(model_gauss_low)

s2 <- summary(model_gauss_low)
s2
a2 <- Anova(model_gauss_low, type = "III") 
a2

emm <- emmeans(model_gauss_low, ~ Antibiotic * Reinoculation)
emm_df <- as.data.frame(emm)
pairs(emm)
```

### Visualisation
```{r}
emm <- emmeans(model_gauss_low, ~ Antibiotic * Reinoculation)
emm_df <- as.data.frame(emm)
pairs(emm)

p2 <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = lower.CL, ymax = upper.CL, group = Reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = emmean, ymax = emmean, group = Reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_interind,
    aes(x = Antibiotic, y = Mean_dist, fill = Reinoculation),
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Mean interindividual distance [cm]",
    fill = "Re-inoculation"
  )

print(p2)

#save the plot
ggsave(
  filename = here::here("activity-and-behavior", "result", "inter_individual_dist", "mean_interind_2x2_plot.png"), 
  plot = p2, 
  width = 10, 
  height = 7, 
  dpi = 300
)
```

### additive model
```{r}
model_gauss_low_add <- lmer(Mean_dist ~ Antibiotic + Reinoculation + (1 | Year),data = df_interind)
simulateResiduals(model_gauss_low_add) |> plot()
testDispersion(model_gauss_low_add); 
testResiduals(model_gauss_low_add) # tests supplémentaires

summary(model_gauss_low_add)
anova(model_gauss_low_add)
```


# Save the result in a .txt file

```{r}
sink(here::here("activity-and-behavior", "result", "inter_individual_dist", "interind_dist_result.txt"), split = TRUE)

cat("Interind dist for full 3x2 model\n\n")
a1 
s1

cat("Interind dist for reduced 2x2 model\n\n")
a2
s2

sink()
```



#==============================
#Analysis for activity (avg speed + time spent moving) 

## Basic stats
```{r}
#DESCRIPTIVES STATISTICS
moyenne_speed <- tapply(df_move$Average_Speed, df_move$Treatment, mean, na.rm = TRUE)
ecarts_speed <- tapply(df_move$Average_Speed, df_move$Treatment, sd, na.rm = TRUE)

resultats_speed <- data.frame(
  Moyenne = moyenne_speed,
  Ecart_Type = ecarts_speed
)

moyenne_prop <- tapply(df_move$Prop_time_moving, df_move$Treatment, mean, na.rm = TRUE)
ecarts_prop <- tapply(df_move$Prop_time_moving, df_move$Treatment, sd, na.rm = TRUE)

resultats_prop <- data.frame(
  Moyenne = moyenne_prop,
  Ecart_Type = ecarts_prop
)

#save table
dir.create(here::here("activity-and-behavior", "result", "activity"), recursive = TRUE, showWarnings = FALSE)

write.csv(resultats_speed, here::here("activity-and-behavior", "result", "activity", "avg_speed_table.csv"), row.names = TRUE)
write.csv(resultats_prop, here::here("activity-and-behavior", "result", "activity", "prop_of_time_moving_table.csv"), row.names = TRUE)
```



#------------------------------------------------------
# Average speed analysis

## LLM full model

```{r}
hist(df_move$Average_Speed, breaks = 50)
qqnorm(df_move$Average_Speed)

model_lmm <- lmer(Average_Speed ~ Antibiotic * Reinoculation + (1 | Year:Arena), data = df_move)
simulateResiduals(model_lmm) |> plot()
testDispersion(model_lmm)
testResiduals(model_lmm)

s1 <- summary(model_lmm)
s1
a1 <- Anova(model_lmm, type=3)
a1 

```

## Visualisation
### Simple box plot 
```{r}
p_boxplot <- ggplot(df_move, aes(x = Antibiotic, y = Average_Speed, fill = Reinoculation)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.1,
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    mapping = aes(ymin = after_stat(y), ymax = after_stat(y)),
    width = 0.5,
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Average speed [cm/sec]",
    fill = "Re-inoculation"
  )

print(p_boxplot)


ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "avgspeed_boxplot.pdf"), 
  plot = p_boxplot, 
  width = 9, 
  height = 5
)
```

### The model prediction
```{r}
emm <- emmeans(model_lmm, ~ Antibiotic * Reinoculation)
emm_df <- as.data.frame(emm)

p_model <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = lower.CL, ymax = upper.CL, group = Reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = emmean, ymax = emmean, group = Reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_move, 
    aes(x = Antibiotic, y = Average_Speed, fill = Reinoculation), 
    shape = 21,            
    colour = "black",       
    stroke = 0.3,          
    alpha = 0.9, 
    size = 3.5, 
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Average speed [cm/sec]",
    fill = "Re-inoculation"   
  )

print(p_model)

ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "avg_speed_model_pred.pdf"), 
  plot = p_model, 
  width = 9, 
  height = 5
)
```

### Additive model

```{r}
model_lmm_add <- lmer(Average_Speed ~ Antibiotic + Reinoculation + (1 | Year:Arena), data = df_move)
simulateResiduals(model_lmm_add) |> plot()
testDispersion(model_lmm_add)
testResiduals(model_lmm_add)

s2 <- summary(model_lmm_add)
s2
```

## Reduced model

```{r}
#Same stats but joining LOW and ZERO
df_move_red <- df_move
df_move_red$Antibiotic[df_move_red$Antibiotic %in% c("low", "zero")] <- "zero"
df_move_red<- df_move_red %>%
  mutate(Treatment = paste(Antibiotic, Reinoculation, sep = "_"))

df_move_red <- df_move_red %>%
  mutate(
    Treatment = factor(Treatment),                   # d’abord en facteur
    Treatment = relevel(Treatment, ref = "zero_no"),
    Antibiotic = factor(Antibiotic),
    Antibiotic = relevel(Antibiotic, ref = "zero"),
    Reinoculation = factor(Reinoculation),
    Reinoculation = relevel(Reinoculation, ref = "no"),
    Year = factor(Year),
    Arena = factor(Arena)
  )
df_move_red$Antibiotic <- factor(df_move_red$Antibiotic, levels = c("zero", "high"))
```

## LLM reduced

```{r}
hist(df_move_red$Average_Speed, breaks = 50)
qqnorm(df_move_red$Average_Speed)

model_lmm_red <- lmer(Average_Speed ~ Antibiotic * Reinoculation + (1 | Year:Arena), data = df_move_red)
simulateResiduals(model_lmm_red) |> plot()
testDispersion(model_lmm_red)
testResiduals(model_lmm_red)

s3 <- summary(model_lmm_red)
s3
a3 <- Anova(model_lmm_red, type=3)
a3 
```

### additive model

```{r}

model_lmm_add <- lmer(Average_Speed ~ Antibiotic + Reinoculation + (1 | Year:Arena), data = df_move_red)
simulateResiduals(model_lmm_add) |> plot()
testDispersion(model_lmm_add)
testResiduals(model_lmm_add)

s4 <- summary(model_lmm_add)
s4
```

## Save the result in a .txt file

```{r}
sink(here::here("activity-and-behavior", "result", "activity", "avg_speed_result.txt"), split = TRUE)

cat("\n\n   Full model 3x2 for avg speed \n\n")
 
s1
a1
cat("\n\n additive \n\n")
s2


cat("\n\n Reduced model 2x2 for avg speed \n\n")
cat("\n")
s3
a3
cat("\n\n additive \n\n")
s4

sink()
```

# ------------------------------------------------------

# Proportion of time spent moving 



# LLM Full model

```{r}
hist(df_move$Prop_time_moving, breaks = 35)
qqnorm(df_move$Prop_time_moving)
range(df_move$Prop_time_moving)

#Beta-regression should be the best

# transform y to (0,1) range
n <- length(df_move$Prop_time_moving)
y_adj <- (df_move$Prop_time_moving * (n - 1) + 0.5) / n
df_move$prop_time_adj <- y_adj
range(df_move$prop_time_adj)

model_beta <- glmmTMB(prop_time_adj ~ Antibiotic * Reinoculation + (1|Year/Arena),
                      family = beta_family(link="logit"),
                      data=df_move) 

#Verifications check
res <- simulateResiduals(model_beta)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta) 

# Models summary
s1 <- summary(model_beta)
s1
a1 <- Anova(model_beta, type=3)
a1 
```

### Visualisation
```{r}
p_boxplot_beta <- ggplot(df_move, aes(x = Antibiotic, y = prop_time_adj, fill = Reinoculation)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.1,
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    mapping = aes(ymin = after_stat(y), ymax = after_stat(y)),
    width = 0.5,
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    shape = 21,
    colour = "black",
    stroke = 0.3,
    size = 3.5,         
    alpha = 0.9,         
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Proportion of time moving",
    fill = "Re-inoculation"
  )

print(p_boxplot_beta)

ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "prop_box_plot.pdf"), 
  plot = p_boxplot_beta, 
  width = 9, 
  height = 5
)

```


```{r}
emm <- emmeans(model_beta, ~ Antibiotic * Reinoculation, type = "response") 
emm_df <- as.data.frame(emm)

p_pred_model_prop <- ggplot() +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = asymp.LCL, ymax = asymp.UCL, group = Reinoculation),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = Antibiotic, ymin = response, ymax = response, group = Reinoculation),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  geom_point(
    data = df_move, 
    aes(x = Antibiotic, y = prop_time_adj, fill = Reinoculation), 
    shape = 21,            
    colour = "black",       
    stroke = 0.3,          
    alpha = 0.9, 
    size = 3.5, 
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "Proportion of time moving",
    fill = "Re-inoculation"   
  )

print(p_pred_model_prop)

ggsave(
  filename = here::here("activity-and-behavior", "result", "activity", "prop_pred_model.pdf"), 
  plot = p_pred_model_prop, 
  width = 9, 
  height = 5
)

```


### Additive model (just to have it)

```{r}
model_beta_add <- glmmTMB(prop_time_adj ~ Antibiotic + Reinoculation + (1|Year/Arena),
                          family = beta_family(link="logit"),
                          data=df_move) 

#Verifications check
res <- simulateResiduals(model_beta_add)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta_add) 

#Résumé modele
s2 <- summary(model_beta_add)
s2
a2 <- Anova(model_beta_add, type=3)
a2
```

# Reduced model

## LLM reduced

```{r}

hist(df_move_red$Prop_time_moving, breaks = 35)
qqnorm(df_move_red$Prop_time_moving)
range(df_move_red$Prop_time_moving)

#Beta-regression should be the best
# transform y to (0,1) range
n <- length(df_move_red$Prop_time_moving)
y_adj <- (df_move_red$Prop_time_moving * (n - 1) + 0.5) / n
df_move_red$prop_time_adj <- y_adj
range(df_move_red$prop_time_adj)

model_beta <- glmmTMB(prop_time_adj ~ Antibiotic * Reinoculation + (1|Year/Arena),
                      family = beta_family(link="logit"),
                      data=df_move_red) 

#Verifications check
res <- simulateResiduals(model_beta)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta) 


#Résumé modele
s3 <- summary(model_beta)
s3
a3 <- Anova(model_beta, type=3)
a3
```
### additive model

```{r}
model_beta_add <- glmmTMB(prop_time_adj ~ Antibiotic + Reinoculation + (1|Year/Arena),
                          family = beta_family(link="logit"),
                          data=df_move_red)

#Verifications check
res <- simulateResiduals(model_beta_add)
plot(res); 
testDispersion(res)
testUniformity(res)
testResiduals(res) 
testZeroInflation(model_beta) 

#Résumé modele
s4 <- summary(model_beta_add)
s4
a4 <- Anova(model_beta_add, type=3)
a4
```

#Save the result in a .txt file

```{r}
sink(here::here("activity-and-behavior", "result", "activity", "prop_time_moving_result.txt"), split = TRUE)

cat("\n\n  Full model 3x2 for prop of time moving \n\n")
cat("\n")

s1
a1

cat("\n\n additive \n\n")
cat("\n")

s2
a2


cat("\n\n Reduced model 2x2 for pro of time moving \n\n")
cat("\n")

s3
a3
cat("\n\n additive \n\n")
cat("\n")

s4
a4

sink()
```

##Combined plot 
```{r}
model_lmm <- lmer(Average_Speed ~ Antibiotic * Reinoculation + (1 | Year:Arena), data = df_move)
model_beta <- glmmTMB(prop_time_adj ~ Antibiotic * Reinoculation + (1 | Year/Arena), family = beta_family(link = "logit"), data = df_move)

emm_lmm_df <- as.data.frame(emmeans(model_lmm, ~ Antibiotic * Reinoculation))
emm_beta_df <- as.data.frame(emmeans(model_beta, ~ Antibiotic * Reinoculation, type = "response"))

p_model <- ggplot() +
  geom_errorbar(data = emm_lmm_df, aes(x = Antibiotic, ymin = lower.CL, ymax = upper.CL, group = Reinoculation), width = 0.1, linewidth = 0.5, position = position_dodge(0.8)) +
  geom_errorbar(data = emm_lmm_df, aes(x = Antibiotic, ymin = emmean, ymax = emmean, group = Reinoculation), width = 0.5, linewidth = 1.2, position = position_dodge(0.8)) +
  geom_point(data = df_move, aes(x = Antibiotic, y = Average_Speed, fill = Reinoculation), shape = 21, colour = "black", stroke = 0.3, size = 3.5, alpha = 0.9, position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)) +
  scale_fill_manual(values = c("no" = "#4F77B4", "yes" = "#FFBF00"), labels = c("No", "Yes")) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1), axis.text.x = element_text(margin = margin(t = 5))) +
  labs(x = "Antibiotic dose", y = "Average speed [cm/sec]", fill = "Re-inoculation")

p_pred_model_prop <- ggplot() +
  geom_errorbar(data = emm_beta_df, aes(x = Antibiotic, ymin = asymp.LCL, ymax = asymp.UCL, group = Reinoculation), width = 0.1, linewidth = 0.5, position = position_dodge(0.8)) +
  geom_errorbar(data = emm_beta_df, aes(x = Antibiotic, ymin = response, ymax = response, group = Reinoculation), width = 0.5, linewidth = 1.2, position = position_dodge(0.8)) +
  geom_point(data = df_move, aes(x = Antibiotic, y = prop_time_adj, fill = Reinoculation), shape = 21, colour = "black", stroke = 0.3, size = 3.5, alpha = 0.9, position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)) +
  scale_fill_manual(values = c("no" = "#4F77B4", "yes" = "#FFBF00"), labels = c("No", "Yes")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  theme_classic() + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1), axis.text.x = element_text(margin = margin(t = 5))) +
  labs(x = "Antibiotic dose", y = "Proportion of time moving", fill = "Re-inoculation")

combined_plot <- (p_model + p_pred_model_prop) + 
  plot_layout(guides = "collect") + 
  plot_annotation(tag_levels = 'A') & 
  theme(legend.position = "bottom", plot.tag = element_text(face = "bold", size = 16))

print(combined_plot)

ggsave(filename = here::here("activity-and-behavior", "result", "activity", "combined_activity_metrics.pdf"), plot = combined_plot, width = 12, height = 5.5)

```
#===============================
# PCA ON 7 METRICS


### PCA for Treatment (3 x 2 levels)

```{r}
data2 <- df_pca_7[, !(names(df_pca_7) %in% c("antibiotic", "year", "reinoculate","category", "nb.of.antennal.interaction","arena"))]
head(data2)

X <- dimensio::pca(data2, center = TRUE, scale = TRUE, sup_quali = "treatment")
eig <- get_eigenvalues(X)
eig
screeplot(X, cumulative = TRUE)

viz_variables(X)


viz_individuals(
  x = X,
  extra_quali = data2$treatment,
  color = c("#4477AA", "#EE6677", "#528833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)


# === 3) MULTIVARIATE STATISTICS (PERMANOVA & BETADISPER) ===
set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data2[, sapply(data2, is.numeric)]

# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.

dispersion_mod <- betadisper(dist_matrix,data2$treatment)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_treat <- permutest(dispersion_mod, permutations = 999)
permu_treat

# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Treatment
ado_treat <- adonis2(dist_matrix ~ treatment, strata = df_pca_7$year, data = data2, method = "euclidean")
ado_treat

```

### PCA for Antibiotic

```{r}
data3 <- df_pca_7[, !(names(df_pca_7) %in% c("treatment", "year", "reinoculate","category", "nb.of.antennal.interaction","arena"))]
data3$antibiotic <- as.character(data3$antibiotic)
head(data3)

X <- dimensio::pca(data3, center = TRUE, scale = TRUE, sup_quali = "antibiotic")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data3$antibiotic,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)


# === 3) MULTIVARIATE STATISTICS (PERMANOVA & BETADISPER) ===
set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data3[, sapply(data3, is.numeric)]

# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")

# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data3$antibiotic)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_ant <- permutest(dispersion_mod, permutations = 999)
permu_ant
# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_ant <- adonis2(dist_matrix ~ antibiotic, strata=df_pca_7$year, data = data3, method = "euclidean")
ado_ant

```

### PCA for réinoculation

```{r}
data4 <- df_pca_7[, !(names(df_pca_7) %in% c("treatment", "antibiotic", "year","category", "nb.of.antennal.interaction","arena"))]

head(data4)

X <- dimensio::pca(data4, center = TRUE, scale = TRUE, sup_quali = "reinoculate")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

viz_individuals(
  x = X,
  extra_quali = data4$reinoculate,
  color = c("#4477AA", "#EE6677", "#228833", "#9477AA", "#AE6677", "#E28833"), # Custom color scheme
  #symbol = c(1, 2, 3, 4, 5, 6, 7), # Custom symbols
  legend = list(x = "topright") ,
  ellipse = list(type = "tolerance", level = 0.95)
)


set.seed(123) 
# A. Data Preparation for Vegan
stats_matrix <- data4[, sapply(data4, is.numeric)]
# Calculate Euclidean Distance Matrix // We use Euclidean because PCA preserves Euclidean distances.
dist_matrix <- vegdist(stats_matrix, method = "euclidean")


# B. Check for Homogeneity of Dispersion (Betadisper)
# This tests: "Do some groups have more variable behavior than others?" If this is significant, the groups have different shapes/spreads.
dispersion_mod <- betadisper(dist_matrix, data4$reinoculate)

# Visualize the dispersion
plot(dispersion_mod, main = "Multivariate Dispersion (Betadisper)")
boxplot(dispersion_mod, main = "Distance to Centroid per Group")

# BETADIASPER RESULT - Test significance of dispersion
permu_reino <- permutest(dispersion_mod, permutations = 999)
permu_reino

# C. PERMANOVA (Adonis2)
# This tests: "Are the centers of the groups significantly different?"
set.seed(123) # Ensures results are reproducible
# Test 1: Simple effect of Antibiotic
ado_reino <- adonis2(dist_matrix ~ reinoculate, strata=df_pca_7$year, data = data4, method = "euclidean")
ado_reino

```

### PCA on the PC1

#### extract the PC1

```{r}
data5 <- df_pca_7[, !(names(df_pca_7) %in% c("antibiotic", "year", "reinoculate","category", "nb.of.antennal.interaction","arena"))]
head(data5)

X <- dimensio::pca(data5, center = TRUE, scale = TRUE, sup_quali = "treatment")

get_eigenvalues(X)
screeplot(X, cumulative = TRUE)
biplot(X, type = "form", labels = "variables")
biplot(X, type = "covariance", labels = "variables")
viz_variables(X)

pca_scores <- get_coordinates(X)

combined_interaction_data <- pca_scores$F1

df_pca_7$combined <- combined_interaction_data
```

#### LMM on PC1

```{r}
hist(df_pca_7$combined, breaks = 10)
qqnorm(df_pca_7$combined)


model_lmm <- lmer(combined ~ antibiotic * reinoculate + (1 | year), data = df_pca_7)


res <- simulateResiduals(model_lmm)
plot(res)
testDispersion(res)
testUniformity(res) # This is the standard test for residual distribution


s1 <- summary(model_lmm)
s1
a1 <- Anova(model_lmm, type=3)
a1

```

### Visualisation

```{r}

emm <- emmeans(model_lmm, ~ antibiotic * reinoculate)
all_pairwise <- pairs(emm)
emm_df <- as.data.frame(emm)

p1 <- ggplot() +
  # 1. Résultats du modèle LMM : Moustaches IC95 (En noir, fines)
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = lower.CL, ymax = upper.CL, group = reinoculate),
    width = 0.1, 
    linewidth = 0.5,
    position = position_dodge(0.8)
  ) +
  # 2. Résultats du modèle LMM : Ligne horizontale pour la Moyenne prédite
  geom_errorbar(
    data = emm_df,
    aes(x = antibiotic, ymin = emmean, ymax = emmean, group = reinoculate),
    width = 0.5, 
    linewidth = 1.2,
    position = position_dodge(0.8)
  ) +
  # 3. Données brutes : Points individuels avec contour noir (style habituel)
  geom_point(
    data = df_pca_7, 
    aes(x = antibiotic, y = combined, fill = reinoculate), 
    shape = 21,            
    colour = "black",       
    stroke = 0.3,          
    alpha = 0.9, 
    size = 3.5, 
    position = position_jitterdodge(jitter.width = 0.37, dodge.width = 0.8, seed = 32)
  ) +
  # 4. Palette de couleurs et étiquettes épurées
  scale_fill_manual(
    values = c("no" = "#4F77B4", "yes" = "#FFBF00"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05))) +
  # 5. Thème classique épuré (Cadre noir complet, pas de lignes d'axes)
  theme_classic() + 
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "bottom",
    axis.text.x = element_text(margin = margin(t = 5))
  ) +
  labs(
    x = "Antibiotic dose",
    y = "PC1 value",
    fill = "Re-inoculation"   
  )

print(p1)

dir.create(
  path = here::here("activity-and-behavior", "result", "PCA_on_7metrics"), 
  showWarnings = FALSE, 
  recursive = TRUE
)

ggsave(
  filename = here::here("activity-and-behavior", "result", "PCA_on_7metrics", "PC1_Plot.pdf"), 
  plot = p1, 
  width = 9, 
  height = 5
)

```

# Save the result in a .txt file

```{r}
sink(
  file = here::here("activity-and-behavior", "result", "PCA_on_7metrics", "pca_result.txt"), 
  split = TRUE
)

cat("===================================================\n")
cat("               PCA ON THE 7 METRICS                \n")
cat("===================================================\n\n")

cat("--- Eigenvalues & Variance Explained ---\n")
print(eig)
cat("\n\n")

cat("--- PCA for Treatment ---\n")
print(permu_treat)
print(ado_treat)
cat("\n\n")

cat("--- PCA for Antibiotic ---\n")
print(permu_ant)
print(ado_ant)
cat("\n\n")

cat("--- PCA for Reinoculation ---\n")
print(permu_reino)
print(ado_reino)
cat("\n\n")

cat("--- Linear Mixed Model on PC1 ---\n")
print(a1) 
print(s1) 

sink()

```
